### 2015-05-11 ms.tex
version submitted to EJS

\documentclass{report}

\usepackage{amsmath,amsfonts}
\usepackage{authblk}
\usepackage{graphicx}
\usepackage{caption,subcaption}
%\usepackage{color}
%\usepackage{etoolbox}
\usepackage{float}
\usepackage{fullpage}
%\usepackage{mathtools}
\usepackage{natbib}
\usepackage{paralist}
%\usepackage{setspace}

\renewcommand\thesection{\arabic{section}}
\usepackage{xspace}
\newcommand{\textcompute}{\textsf}
\newcommand{\N}{\text{N}} % Normal 
\newcommand{\R}{\textcompute{R}\xspace}
\newcommand{\IG}{\text{InvGam}}
\newcommand{\RL}{f}
\newcommand{\RLorig}{\text{RL}}
\newcommand{\logRL}{\log\RL}
\newcommand{\logRLorig}{\log\RLorig}
\newcommand{\sigssq}{\sigma_s^2}
\newcommand{\sigesq}{\sigma_e^2}
\newcommand{\sshat}{\hat\sigma^2_e,\hat\sigma^2_s}
\newcommand{\logRLss}{\logRL(\sigesq,\sigssq)}
\newcommand{\logRLssorig}{\logRLorig(\sigesq,\sigssq)}
\newcommand{\ass}{a_j\sigssq + \sigesq}
\newcommand{\abss}{a_j\sigssq + b_j\sigesq}
\newcommand{\fss}{f(\sigesq,\sigssq)\,}
%\newcommand{\mrle}{MRLE}
\newcommand{\mrle}{$\argsup\log f$}
%\newcommand{\lone}{\textbf{l}_1}
%\newcommand{\pone}{\textbf{p}_1}
%\newcommand{\ltwo}{\textbf{l}_2}
%\newcommand{\ptwo}{\textbf{p}_2}
%\newcommand{\pthree}{\textbf{p}_3}
%\newcommand{\given}{\,|\,}
\newcommand{\g}{\,|\,}
\newcommand{\maxit}{\textcompute{maxit}}
\newcommand{\argmax}{\operatornamewithlimits{argmax}}
\newcommand{\argsup}{\operatornamewithlimits{argsup}}
\renewcommand{\bibname}{References}

\begin{document}

\title{Approximately Exact Calculations for Linear Mixed Models}
\author[1]{Michael Lavine}
\author[2]{James Hodges}
\affil[1]{Department of Mathematics and Statistics, University of Massachusetts, Amherst, MA 01003 USA}
\affil[2]{Division of Biostatistics, University of Minnesota, Minneapolis, MN 55455 USA}

\maketitle

\begin{abstract}
This paper is about computations for linear mixed models having two variances, $\sigma^2_e$ for residuals and $\sigma^2_s$ for random effects, though the ideas can be extended to some linear mixed models having more variances.  Researchers are often interested in either the restricted (residual) likelihood $\RLorig(\sigesq,\sigssq)$ or the joint posterior $\pi(\sigesq,\sigssq\g y)$ or their logarithms.  Both $\log\RLorig$ and $\log\pi$ can be multimodal and computations often rely on either a general purpose optimization algorithm or MCMC, both of which can fail to find regions where the target function is high.  This paper presents an alternative.  Letting $f$ stand for either $\RLorig$ or $\pi$, we show how to find a box $B$ in the $(\sigesq,\sigssq)$ plane such that
\begin{enumerate}
\item all local and global maxima of $\log f$ lie within $B$;
\item $\sup_{(\sigesq,\sigssq) \in B^c} \log f(\sigesq,\sigssq) \le \sup_{(\sigesq,\sigssq) \in B} \log f(\sigesq,\sigssq) - M$ for a prespecified $M>0$; and
\item $\log f$ can be estimated to within a prespecified tolerance $\epsilon$ everywhere in $B$ with no danger of missing regions where $\log f$ is large.
\end{enumerate}
Taken together these conditions imply that the $(\sigesq,\sigssq)$ plane can be divided into two parts: $B$, where we know $\log f$ as accurately as we wish,  and $B^c$, where $\log f$ is small enough to be safely ignored.  We provide algorithms to find $B$ and to evaluate $\log f$ as accurately as desired everywhere in $B$.
\end{abstract}

\section{Introduction}
\label{sec:intro}
Linear mixed models are an important class of statistical models.  Books are written about them (e.g.\ \citealt{bryk_raudenbush:1992, verbeke&molenberghs:2000, hodges:2013, west&welch&galecki:2014}),  courses are taught about them, and they have many applications.  Examples include random intercept models (including balanced and unbalanced one-way random effect models), additive models with one penalized spline, spatial models with one intrinsic conditional autoregression (ICAR) random effect, dynamic linear models with one system-level variance, and some multiple membership models (e.g.\ \citealt{browne_etal:2001, mccaffrey_etal:2004}).  Typical notation, which we adopt, is
\begin{equation}
\label{eq:lmm}
	y = X\beta + Zu + \epsilon
\end{equation}
where $y$ is a vector of $n$ observations, $X$ is a known $n \times p$ matrix, $\beta$ is a vector of $p$ unknown coefficients called fixed effects, $Z$ is a known $n \times q$ matrix, $u$ is a vector of $q$ unknown coefficients called random effects, and $\epsilon$ is a vector of $n$ errors. The term ``mixed" is used when we treat $u$ as a vector of random variables, thus mixing fixed and random effects in the same model.  For linear mixed models where $u$ and $\epsilon$ are modelled as Normal, researchers are often interested in the restricted likelihood function
\begin{equation}
\label{eq:rll}
  \RLorig(\theta) = K |V(\theta)|^{-1/2}|X^tV^{-1}(\theta)X|^{-1/2}
                       \exp\left\{ -\frac{1}{2} \left(y^tV^{-1}(\theta)y -
                                                                \tilde\beta^t(\theta)X^tV^{-1}(\theta)X\tilde\beta(\theta)
                                                         \right)
                              \right\}
\end{equation}
where $K$ is an unimportant constant, $\theta$ is a vector of unknown parameters in the covariance matrices of $u$ and $\epsilon$, $V(\theta)$ is the marginal covariance matrix of $y$ implied by the covariance matrices of $u$ and $\epsilon$, and $\tilde\beta(\theta)$ is the generalized least-squares estimate of $\beta$, given $V(\theta)$.
This manuscript deals with the special case in which we adopt the model
\begin{equation*}
	\epsilon \sim \N (0, \sigma_e^2 \Sigma_e) \qquad u \sim \N (0, \sigma_s^2 \Sigma_s)
\end{equation*}
where $\Sigma_e$ and $\Sigma_s$ are known matrices, often the identity, of the appropriate sizes and $\theta \equiv (\sigma_e^2, \sigma_s^2)$, two unknown variance parameters.  The key for this manuscript is that $\theta$ contains only those two unknown variances and no others.  \cite{hodges:2013} gives examples and explains the importance of this special case.

\cite{hodges:2013} also unifies and generalizes \cite{reich_hodges:2008} and \cite{welham_thompson:2009} to show that in our special case, and a few others, $\logRLssorig$ can be expressed as
\begin{equation}
\label{eq:reexpress}
  \logRLssorig = B - \frac{n_e}{2}\log(\sigesq) - \frac{y^t \Gamma_c \Gamma^t_c y}{2\sigesq} 
    - \frac{1}{2} \sum_{j=1}^{s_z} \left[ \log(\ass) + \frac{\hat v_j^2}{\ass}\right]
\end{equation}
where $B$ is an unimportant known constant; $n_e$ is $n$ minus the dimension of the space spanned by the columns of $[X|Z]$; $\Gamma_c$ is $n \times n_e$ and spans the space orthogonal to $[X|Z]$ (so $y^t \Gamma_c \Gamma^t_c y$ is the residual sum of squares); $s_z$ is the dimension of the space spanned by the columns of $Z$ not already in the span of the columns of $X$; and the $\{a_j\}$ and $\{\hat v_j\}$ are known constants with all $a_j>0$.  Thus the only unknowns are $(\sigma^2_s, \sigma^2_e)$ and $\logRLssorig$ is a function of just those two arguments.  Specifically, $\logRLssorig$ is a linear combination of logs and inverses of linear functions $\abss$ of the two unknowns.

As \cite{hodges:2013} further observes, if $\beta$ is given an improper flat prior and $\sigesq$ and $\sigssq$ are given conjugate priors --- say $\sigesq \sim \IG(\alpha_e,\beta_e)$ and $\sigssq \sim \IG(\alpha_s,\beta_s)$ --- then \begin{equation*}
  -(\alpha_e+1) \log\sigesq - \beta_e/\sigesq -(\alpha_s+1) \log\sigssq - \beta_s/\sigssq
\end{equation*}
is added to \eqref{eq:reexpress} to yield
the log posterior
\begin{equation}
\label{eq:logpost}
  \begin{split}
  \log\pi(\sigesq,\sigssq\g y) &=
  B - \frac{n_e + 2\alpha_e + 2}{2}\log(\sigesq) -
    \frac{y^t \Gamma_c \Gamma^t_c y + 2\beta_e}{2\sigesq}\\
    &\qquad\qquad - \frac{2\alpha_s + 2} {2}\log(\sigssq) - \frac{2\beta_s}{2\sigssq}
    - \frac{1}{2} \sum_{j=1}^{s_z} \left[ \log(\ass) + \frac{\hat v_j^2}{\ass}\right]\\
    &= B - \frac{1}{2}\sum_j\left[ c_j \log(\abss) + \frac{d_j}{\abss}\right]
  \end{split}
\end{equation}
for known nonnegative constants $\{a_j, b_j, c_j, d_j\}$ ---
still a linear combination of logs and inverses of linear functions $\abss$.  Our derivation will proceed from \eqref{eq:logpost}, though it applies also to \eqref{eq:reexpress} and other linear combinations of logs and inverses of linear functions $\abss$.  We use $f$ to denote such functions generically.

It is known (e.g.\ \citealt{henn&hodges:2014}) that $\logRLss$\  
can have multiple maxima and that existing general purpose algorithms for linear mixed models may fail to find all of them, as shown by examples in \cite{hodges:2013}, \cite{henn&hodges:2014}, and elsewhere.
\cite{mullen:2014} examines 18 optimization functions available in \R, tests them on 48 objective functions (admittedly more complicated than $\logRLss$) and finds that even the best of them fail in over 10\% of the cases.  \cite{henn&hodges:2014} examine conditions under which multiple maxima occur in posterior densities and conclude ``\dots second maxima in posterior distributions therefore may be more common than reports in the literature would suggest."  Thus, failure to find local and global maxima may be common.

Our point of view is that it is important to find regions where $f$ or $\log f$ is large relative to its maximum regardless of whether the regions contain local maxima.  (E.g., if $f$ or $\log f$ is relatively flat and large over a region, it matters little whether the region contains small bumps that are, technically, local maxima.)  This paper shows how to find $(\sshat) \equiv \argsup_{\sigesq, \sigssq}\logRLss$ globally (typically either the maximum restricted log likelihood estimate or the maximum \textit{a posteriori} estimate) and to evaluate $\logRLss$ everywhere, each to within a prespecified tolerance (hence ``approximately exact''), without fear of missing regions of high $f$ or $\logRL$.  The technique relies on the partial derivatives of $\logRLss$.  Analysis of the partial derivatives allows us to satisfy two desiderata.
\begin{description}
\item[D1] For any prespecified constant $M>0$ we can find a box $B$, a rectangle in the first quadrant of the $(\sigssq, \sigesq)$ plane whose sides are parallel to the axes, such that
\begin{equation}
\label{eq:boundingbox}
	\text{all local maxima are in $B$} \qquad \text{and} \qquad \sup_{(\sigesq, \sigssq) \in B^c} \logRLss \le \logRL(\sshat) - M.
\end{equation}
In practice we will take $M$ to be large enough to interpret (\ref{eq:boundingbox}) as meaning that we can restrict attention to $B$ because values of $(\sigesq, \sigssq) \in B^c$ have $\logRLss$ too low to be of further interest.

\item[D2] For any box $b$ we can quickly compute lower and upper bounds $(L^b,U^b)$ satisfying
\begin{equation*}
  L^b \le \inf_{(\sigesq, \sigssq) \in b}\logRLss \qquad\text{and}\qquad
  U^b \ge \sup_{(\sigesq, \sigssq) \in b}\logRLss
\end{equation*}
and such that $U^b-L^b \rightarrow 0$ as $b$ shrinks.  Therefore, partitioning the box $B$ from \textbf{D1} allows us to know $\logRLss$ everywhere in $B$ to within a pre-specified tolerance and also to locate $\argsup\logRL$ to within a pre-specified tolerance without fear of missing regions of high $\logRL$.
\end{description}
The next section shows how the partial derivatives are used to satisfy \textbf{D1} and \textbf{D2}.

\section{Satisfying the Desiderata}
\subsection{Partial Derivatives Determine Lines}
The partial derivatives of $\logRLss$ can be calculated from (\ref{eq:logpost}):
\begin{subequations}
\label{eq:partials}
\begin{equation}
\label{eq:drlds}
  \frac{\partial\logRLss}{\partial\sigssq}
  = -\frac{1}{2} \sum_j
        \left[ \frac{a_jc_j}{\abss} - \frac{a_j d_j}{(\abss)^2} \right]
        = -\frac{1}{2} \sum_j
           \frac{a_j \left( a_jc_j\sigssq + b_jc_j\sigesq  - d_j\right)}{(\abss)^2}\\
\end{equation}
and
\begin{equation}\label{eq:drlde}
  \frac{\partial\logRLss}{\partial\sigesq} =
    -\frac{1}{2} \sum_j \left[ \frac{b_jc_j}{\abss} - \frac{b_j d_j}{(\abss)^2} \right]
    = - \frac{1}{2} \sum_j \frac{b_j \left( a_jc_j\sigssq + b_jc_j\sigesq  - d_j\right)}{(\abss)^2}.
\end{equation}
\end{subequations}
We work with one term in (\ref{eq:partials})'s summations at a time; that is, one $j$ at a time.  The $j$'th terms in (\ref{eq:partials}) differ only by a multiplicative constant $a_j/b_j$; they have the same sign as each other and the same sign as $(a_jc_j\sigssq + b_jc_j\sigesq  - d_j)$, which determines a line $\sigssq = d_j/a_j c_j - b_j \sigesq/a_j$ --- call it the $j$'th line --- in the first quadrant of the $(\sigssq,\sigesq)$ plane.  The line has positive intercept and negative slope because all the constants are nonnegative.  Both partial derivatives of the $j$'th term are positive below the $j$'th line, 0 on the line, and negative above the line, as indicated in Figure~\ref{fig:oneline}.  The $j$'th term is constant on the $j$'th line and attains its maximum there.  Both \eqref{eq:reexpress} and \eqref{eq:logpost} contain a term, call it the $v$'th (for vertical) term,  with $a_j=0$.  The corresponding line is vertical at $\sigesq = \sigma_e^{2v} \equiv d_v / b_v c_v$.  The $v$'th term in (\ref{eq:partials}) is negative to the right of the $v$'th line, positive to the left, and 0 on the line; hence the $v$'th term in \eqref{eq:logpost} is maximized on the $v$'th line.  \eqref{eq:logpost} also contains a term, call it the $h$'th (for horizontal) term, with $b_j=0$.  The corresponding line is horizontal at $\sigssq = \sigma_e^{2h} \equiv d_h / a_h c_h$.  The $h$'th term in \eqref{eq:logpost} is maximized on the $h$'th line.

\begin{figure}[h]
	\centering
	\includegraphics[width=.5\linewidth]{figs/oneline.pdf}
	\caption{For a single summand, i.e., a fixed $j$, in (\ref{eq:logpost}),
	              the partial derivatives (\ref{eq:drlds}) and (\ref{eq:drlde})
	              are 0 on the line and positive and negative where indicated
	              by ``\textbf{+}" and ``\textbf{-}".}
	\label{fig:oneline}
\end{figure}

\subsection{Lines Determine a Bounding Box}
\label{sec:boundingbox}

Figure~(\ref{fig:bigboundingbox}) shows five lines --- realistic data sets may have more --- labelled $j=1, 2, 3, v, h$.  The largest intercept of those lines on the $\sigssq$ axis and the largest intercept on the $\sigesq$ axis determine a rectangle $B_1$ whose outline is marked in bold.  At any point above $B_1$, e.g., $p_1$, the partial derivatives \eqref{eq:partials}, for all $j$, are negative or, for $j=v$, zero.  Therefore, $\logRL(p_1^*) \ge \logRL(p_1)$.  Similarly, for points like $p_2$, $\logRL(p_2) \ge \logRL(p_2^*)$ and, for points like $p_3$, $\logRL(p_3^*) \ge \logRL(p_3)$.  That is, for points above and/or to the right of $B_1$, $\logRLss$ can be increased by moving down and/or to the left as far as $B_1$'s boundary.  Therefore \mrle\ must lie on or inside $B_1$ and there can be no maxima of $\logRLss$ in $B_1^c$.  $B_1$ could be passed to an optimizer such as \R's \textsf{optim} or \textsf{nlminb}, with potentially better results than using those functions without bounds.  However, even with known bounds, general purpose optimizers may still miss \mrle\ and regions of high $\logRL$.  This paper presents a computational method guaranteed not to miss \mrle\ and which evaluates $\logRLss$ everywhere to within a specified tolerance inside a box $B$ satisfying desideratum \textbf{D1}.  First, though, we pause to note that the region outlined in bold in Figure~(\ref{fig:smallboundingbox}) is a subset of $B_1$ that must also contain \mrle\ for the reasons given above.  But it is not rectangular, hence less convenient than $B_1$, so we don't pursue it further.

\begin{figure}[h]
  \begin{subfigure}{.5\textwidth}
	\centering
	\includegraphics[width=.8\linewidth]{figs/boundingbox.pdf}
	\caption{A rectangular region containing all maxima.}
	\label{fig:bigboundingbox}
  \end{subfigure}
  \begin{subfigure}{.5\textwidth}
	\centering
	\includegraphics[width=.8\linewidth]{figs/smallboundingregion.pdf}
	\caption{A smaller region containing all maxima.}
	\label{fig:smallboundingbox}
  \end{subfigure}
  \caption{all local and global maxima lie in the regions bounded by the solid
                dark line.
               }
  \label{fig:boundingbox}
\end{figure}

To find $B$, choose an arbitrary constant $M>0$ and an arbitrary point inside $B_1$, say $p_4$, as illustrated in Figure~(\ref{fig:boundingbox2}) and define $L \equiv \logRL(p_4)$.  We will find a box $B\supset B_1$, as shown in Figure~(\ref{fig:boundingbox2}), outside of which $\logRLss \le L-M$, thus satisfying \textbf{D1}.  We have drawn a figure without an $h$ line so the reader can see  the derivation without it, which would apply to $\logRLorig$.  If there were an $h$ line, it would influence the derivation for Figure~(\ref{fig:boundingbox2})'s $q_1$ in a manner analogous to how the $v$ line influences the argument for $q_2$ and $q_3$.

$B$ is determined by its intercepts $\sigma_e^{2*}$ and $\sigma_s^{2*}$ on the $\sigesq$ and $\sigssq$ axes respectively.  Let $q_1^* = (\sigma_e^{2*},0)$ be the intercept of $B$ and the $\sigesq$ axis.  Because each term of the partial derivatives \eqref{eq:partials} is nonpositive to the right of $B_1$, for any point $q_1$ to the right of $B$, every term in the summation of \eqref{eq:logpost} (with the possible exception of an $h$ term) is larger at $q_1^*$ than at $q_1$.  Therefore, $\logRL(\sshat) - \logRL(q_1) \ge L - \logRL(q_1^*)$.  The latter expression is greater than $M$ iff $\logRL(q_1^*) < L-M$.  But examination of (\ref{eq:logpost}) shows that for any fixed $\sigssq$, in particular for $\sigssq=0$, $\lim_{\sigesq \rightarrow \infty} \logRLss = -\infty$.  Thus by choosing $\sigma_e^{2*}$ large enough so $\logRL(q_1^*) < L-M$, we satisfy (\ref{eq:boundingbox}) for all points with $\sigesq \ge \sigma_e^{2*}$.

Let $q_2^* = (\sigma_e^{2v},\sigma_s^{2*})$.  For any point $q_2$ having $\sigssq \ge \sigma_s^{2*}$ and $\sigesq \in [\sigma_e^{2v}, \sigma_e^{2*}]$ --- i.e., above $B$ and to the right of the $v$'th line --- the partial derivatives are nonpositive so $\logRL(\sshat) - \logRL(q_2) \ge L - \logRL(q_2^*)$, which is larger than $M$ if $\logRL(q_2^*) < L-M$.  But for any fixed $\sigesq$, in particular for $\sigesq=\sigma_e^{2v}$, $\lim_{\sigssq \rightarrow \infty} \logRLss = -\infty$.  So by choosing $\sigma_s^{2*}$ large enough we satisfy (\ref{eq:boundingbox}) for all points with $\sigssq \ge \sigma_s^{2*}$ and $\sigesq \in [\sigma_e^{2v}, \sigma_e^{2*}]$.

Finally, Let $q_3^* = (0,\sigma_s^{2*})$ be the intercept of $B$ and the $\sigssq$ axis.  Points such as $q_3$ which satisfy $\sigssq \ge \sigma_s^{2*}$ and $\sigesq \le \sigma_e^{2v}$ --- above $B$ and to the left of the $v$'th line --- are in a region where the partial derivatives (\ref{eq:drlds}, \ref{eq:drlde}) are negative for $j \in \{1, \dots, s_z\}$ but nonnegative for $j=v$.  Let $\logRL_j(\cdot)$ be the $j$'th term of (\ref{eq:reexpress}) evaluated at $(\cdot)$.  For $j\ge1$ we compare $\logRL_j(q_3)$ to $\logRL_j(q_3^*)$ but we compare $\logRL_v(q_3)$ to $\logRL_v(q_2^*)$.
\begin{equation}
	\logRL(\sshat) - \logRL(q_2) \ge L - \sum_{j=1}^{s_z} \logRL_j(q_3^*) - \logRL_v(q_2^*)
\end{equation}
which is larger than $M$ if $\sum_{1}^{s_z} \logRL_j(q_3^*) < L - M - \logRL_v(q_2^*) = L - M + n_e(\log(\sigma_e^{2v}) + 1)/2$.

Combining the requirements for points like $q_1$, $q_2$, and $q_3$, we satisfy desideratum \textbf{D1} for all points in $B^c$ by choosing $\sigma_e^*$ large enough to satisfy $\logRL(q_1^*) < L-M$ and choosing $\sigma_s^*$ large enough to satisfy $\logRL(q_2^*) < L-M$ and $\sum_{1}^{s_z} \logRL_j(q_3^*) < L - M + n_e(\log(\sigma_e^{2v}) + 1)/2$.  In addition, because $\logRL$ has no maxima outside of $B_1$, it also has no maxima outside of $B$.  As already mentioned, if there is an $h$ line, then $q_1$ would have to be treated like $q_2$ and $q_3$.

\begin{figure}[h]
	\centering
	\includegraphics[width=.5\linewidth]{figs/boundingbox2.pdf}
	\caption{Box $B$ satisfies (\ref{eq:boundingbox})}
	\label{fig:boundingbox2}
\end{figure}

\subsection{Lines Determine Bounds within Boxes}
\begin{figure}
	\centering
	\includegraphics[width=.5\linewidth]{figs/boxb.pdf}
	\caption{For a box $b$,  the minimum and
	maximum within $b$ of $\logRL_j$
	occur at either the lower left corner, the upper right  corner, or
	on the line.
}	\label{fig:boxb}
\end{figure}
Next we consider the relationship between the $\logRL_j$'s and boxes as depicted in Figure~\ref{fig:boxb}, which shows a box $b$ and three lines.  (Though we haven't drawn $v$ or $h$ lines, the argument for them is similar to the argument for the lines we have drawn.)  For lines such as $j=2$ that lie below $b$ the partial derivatives of $\logRL_j$ are negative, so the maximum and minimum of $\logRL_j$ within $b$ are attained at the lower left and upper right corners respectively.  The situation is reversed for lines like $j=3$ that lie above $b$: the partial derivatives are positive so the maximum and minimum are attained at the upper right and lower left corners respectively.  For lines like $j=1$ that pass through $b$ the minimum is attained at either the upper right or lower left corner while the maximum is attained on the line.

For any box $b$ let $L^b_j = \inf_{p\in b} \logRL_j(p)$ and $U^b_j = \sup_{p\in b} \logRL_j(p)$.  We have just shown that $L^b_j$ and $U^b_j$ are easily computable by evaluating $\logRL_j$ at either two points (two corners for lines like $j=2,3$) or three points (two corners and one on the line for lines like $j=1$).
Armed with the $L^b_j$'s and $U^b_j$'s we can compute bounds on $\logRLss$ within $b$:
\begin{equation}
	L^b \equiv \sum_j L^b_j \le \inf_{(\sigssq,\sigesq)\in b} \logRLss \le
	\sup_{(\sigssq,\sigesq)\in b} \logRLss \le \sum_j U^b_j \equiv U^b.
\label{eq:bounds}
\end{equation}
Because $\logRL$ is continuous, $U^b - L^b \rightarrow 0$ as $b$ shrinks in both directions, thus satisfying desideratum \textbf{D2}.

\section{An Algorithm}
With \textbf{D1} and \textbf{D2} the following  algorithm will evaluate $\logRL$ to arbitrary accuracy everywhere within a box $B$.
\begin{enumerate}
\item Specify constants \maxit, $M$, $\epsilon$, $\delta_e$, and $\delta_s$.
	\begin{enumerate}[(a)]
	\item \maxit\ is the maximum number (could be $\infty$) of iterations of
		the algorithm's loop.
	\item We will not carefully analyze regions of the plane where
		$\logRLss < \logRL(\sshat) - M$.  ($M$ could be $\infty$.)
	\item We will evaluate $\logRL$ to within an accuracy of $\epsilon$ (could be 0) inside $B$
	         unless evaluation is stopped by one of the other criteria.
	\item We will not distinguish values of $\sigesq$ separated by less
		than $\delta_e$ (could be 0) nor values of $\sigssq$ separated by
		less than $\delta_s$ (could be 0).  Separation may be specified
		in either absolute or relative terms. (I.e.\ we look at either the difference
		in $\sigesq$ or $\log\sigesq$ (or $\sigssq$ or $\log\sigssq$) from one
		side of the box to the other.)
	\end{enumerate}
	\maxit, $M$, $\epsilon$, $\delta_e$, and $\delta_s$ are constants that
	control the running of the algorithm.  Their meaning may become more
	clear as we explain the algorithm.
\item With $y$, $X$, $Z$, $\Sigma_e$, and $\Sigma_s$ --- see \eqref{eq:lmm}
	and \eqref{eq:rll} for definitions --- compute $\{a_j, b_j, c_j, d_j\}$ for $j=1,
	\dots, s_z, v, h$.
\item Create two lists of boxes: active and inactive.  Both lists are initially empty.
\item Specify a box $B$ as the sole entry of the active list.  $B$ could be,
	but doesn't have to be, a box satisfying desideratum \textbf{D1}.
\item Either set $L = -\infty$ or evaluate $\logRL(p_i)$ at a few points
	$p_1, \dots p_k$ and set $L = \max_i\logRL(p_i)$.  $L$ is our current
	lower bound on $\logRL(\sshat)$.
\item While the active list is not empty or we have not reached \maxit\
	iterations
	\begin{enumerate}
	\item For each active box $b$:
		\begin{enumerate}
		\item if $U^b < L-M$ move $b$ to the inactive list.
		\item if $U^b - L^b < \epsilon$ move $b$ to the inactive list
		\item if the horizontal extent of $b$ is less than $\delta_e$,
			move $b$ to the inactive list
		\item if the vertical extent of $b$ is less than $\delta_s$,
			move $b$ to the inactive list
		\item otherwise, $b$ is still active.  Divide $b$ into four subboxes.
			Remove $b$ and add the subboxes to the active list.
		\end{enumerate}
	\item Set $L = \max ( L, \max_{b\in \text{active list}}L^b )$.  $L$ is our new
		lower bound on $\logRL(\sshat)$.
	\end{enumerate}
\item Return the active and inactive lists.
\end{enumerate}

\section{Examples}

\subsection{HMO premiums}
\subsubsection{Introduction to the Data}
Our first example is a traditional linear mixed model previously analyzed in \cite{hodges:98}, \cite{hodges:2013}, and \cite{henn&hodges:2014} who reported a bimodal log posterior density for $(\sigesq,\sigssq)$.  Quoting from \cite{henn&hodges:2014},
\begin{quote}
\itshape
\dots the HMO data set describes 341 HMOs [Health Maintenance Organizations] located in 45 states or similar political jurisdictions.  Each jurisdiction had between 1 and 31 plans with a median of 5 plans.  The data set originally was analysed to assess the cost of moving military retirees and dependents from a Department of Defense health plan to plans serving the US civil service.  \dots.  Specifically, the model is
\begin{equation*}
\begin{split}
	y_{ij} &= \alpha_i + \epsilon_{ij}\\
	\alpha_i &= \varrho_0 + \varrho_1x_{1i} + \varrho_2x_{2i} + \zeta_i,
\end{split}
\end{equation*}
where the fixed effects in $\alpha_i$ include an intercept, jurisdiction-average hospital expenses per admission ($x_{1i}$) and an indicator for plans in New England states ($x_{2i}$).

\upshape
\end{quote}
I.e., $X$ is a $341 \times 3$ matrix with columns for the intercept and two fixed effects and $Z$ is a $341 \times 45$ matrix whose columns are indicators of the 45 jurisdictions.  Because the span of $Z$'s columns contains the span of $X$'s, $s_z = 42$.

\subsubsection{A log RL Analysis}
For a $\logRLssorig$ analysis there are 43 lines, as shown in Figure~\ref{fig:hmolines}.
Running the algorithm on the box $B_1$ determined by the lines' maximum intercepts on the $\sigesq$ and $\sigssq$ with the settings
\begin{equation*}
	\maxit=10; \hspace{1cm} \epsilon=5; \hspace{1cm}
	\delta_e=\log(10); \hspace{1cm} \delta_s=\log(10); \hspace{1cm} M=5
\end{equation*}
results in the the output displayed in Table~\ref{table:hmo_HH11_run}, which shows the state of the algorithm after iterations 1 through 10: the numbers of active and inactive boxes and the current value of the lower bound $L$ on $\max\log f$.  We see that boxes are steadily transferred from the active to the inactive list and that $L$ increases monotonically.  After 10 iterations there is a total of 640 boxes, which are displayed in Figure~\ref{fig:hmo1}.

\begin{table}[h]
\centering
\begin{tabular}{|c|c|c|c|}
\hline
Iteration & $n$ active boxes & $n$ inactive boxes & $L$\\
1 & 4 & 0 & $-\infty$\\
2 & 8 & 2 & -1618.84\\
3 & 16 & 6 & -1509.24\\
4 & 32 & 14 & -1408.12\\
5 & 44 & 35 & -1325.99\\
6 & 56 & 65 & -1265.01\\
7 & 104 & 95 & -1256.35\\
8 & 184 & 153 & -1243.51\\
9 & 224 & 281 & -1240.60\\
10 & 180 & 460 & -1239.49\\
\hline
\end{tabular}
\caption{The state of the algorithm after each of 10 iterations for the HMO data.}
\label{table:hmo_HH11_run}
\end{table}

\begin{figure}
	\centering
	\includegraphics[width=.45\linewidth]{figs/hmo_HH11_lines.pdf}
	\caption{The 43 lines, $j=1, 2, \dots, 42, v$, for the $\logRLorig$ analysis of the HMO data.
	              The $v$ line is dashed.}
	\label{fig:hmolines}
\end{figure}

\begin{figure}
  \begin{subfigure}{.5\textwidth}
	\centering
	\includegraphics[width=.8\linewidth]{figs/hmo_HH11_boxes.pdf}
	\caption{Locations of the boxes.}
	\label{fig:hmoboxes}
  \end{subfigure}
  \begin{subfigure}{.5\textwidth}
	\centering
	\includegraphics[width=.8\linewidth]{figs/hmo_HH11_rll.pdf}
	\caption{Grayscale shows $L^b$ in each box.  Red dot shows $b^* \equiv\argmax L^b$.
	Boxes outlined in blue have $U^b \ge L^{b^*}$.  Axes are logarithmic.}
	\label{fig:hmorll}
  \end{subfigure}
  \caption{The 640 boxes produced in the first run  for the
	$\logRLorig$ analysis of the HMO data. $\maxit=10; \epsilon=5;
	\delta_e=\log(10); \delta_s=\log(10); \text{and\ } M=5$.}
  \label{fig:hmo1}
\end{figure}

Figure~\ref{fig:hmoboxes} shows the outlines of the 640 boxes.  The algorithm did not need to divide the boxes with large $\sigesq$ or $\sigssq$ as finely as those with small $\sigesq$ and $\sigssq$ because they more readily satisfy either $U^b < L-M$ or $U^b - L^b < \epsilon$, so become inactive.  Figure~\ref{fig:hmorll} shows the same boxes, shaded by $L^b$ in each box.  The red dot is the lower left corner of the box that maximizes $L^b$.  Boxes with $U^b \ge \max L^b$ are outlined in blue; $(\sshat)$ must lie within the blue region.  
Figure~\ref{fig:hmorll}  suggests that we can restrict attention to a box determined by $\sigesq\in(300,1000)$ and $\sigssq\in(0,1000)$.  So we rerun the algorithm on that box and with more stringent control parameters:
\begin{equation*}
	\maxit=15; \hspace{1cm} \epsilon=1; \hspace{1cm}
	\delta_e=0; \hspace{1cm} \delta_s=0; \hspace{1cm} M=10
\end{equation*}
Table~\ref{table:hmo_HH11_run2} shows the output.  The algorithm needed 12 iterations to move all boxes to the inactive list.  The resulting 37,516 boxes are shown in Figure~\ref{fig:hmo2}.  For comparison, the standard REML analysis using \R's \textcompute{lme} function yields $\sshat \approx (495, 99)$ with 95\% confidence intervals of $(421, 582)$ and $(39, 248)$.

\begin{table}[h]
\centering
\begin{tabular}{|c|c|c|c|}
\hline
Iteration & $n$ active boxes & $n$ inactive boxes & $L$\\
1 & 4 & 0 & $-1320.51$\\
2 & 16 & 0 & -1276.23\\
3 & 64 & 0 & -1252.18\\
4 & 256 & 0 & -1243.52\\
5 & 1024 & 0 & -1239.61\\
6 & 4020 & 19 & -1237.78\\
7 & 13128 & 757 & -1236.70\\
8 & 14092 & 10362 & -1236.16\\
9 & 9444 & 22093 & -1235.88\\
10 & 5716 & 30108 & -1235.88\\
11 & 2256 & 35260 & -1235.88\\
12 & 0 & 37516 & -1235.88\\
\hline
\end{tabular}
\caption{The state of the algorithm after each of 12 iterations for the HMO data.}
\label{table:hmo_HH11_run2}
\end{table}

\begin{figure}
  \begin{subfigure}{.5\textwidth}
	\centering
	\includegraphics[width=.8\linewidth]{figs/hmo_HH11_boxes2.pdf}
	\caption{Locations of the boxes.}
	\label{fig:hmoboxes2}
  \end{subfigure}
  \begin{subfigure}{.5\textwidth}
	\centering
	\includegraphics[width=.8\linewidth]{figs/hmo_HH11_rll2.pdf}
	\caption{Grayscale shows $L^b$ in each box.  Red dot shows $b^* \equiv\argmax L^b$.
	Boxes outlined in blue have $U^b \ge L^{b^*}$.  Axes are logarithmic.}
	\label{fig:hmorll2}
  \end{subfigure}
  \caption{The 37,516 boxes produced in the second run of the algorithm for the
	$\logRLorig$ analysis of the HMO data. $\maxit=15; \epsilon=1;
	\delta_e=0; \delta_s=0; \text{and\ } M=10$.}
  \label{fig:hmo2}
\end{figure}

Figure~\ref{fig:hmorll2} is sufficiently refined that we needn't run the algorithm further.  Figure~\ref{fig:hmorll2} depicts the same $\logRLorig$ as \cite{henn&hodges:2014}'s Figures~2a (MCMC draws) and~2b ($\logRLorig$ contours), but their Figure~2a was produced by MCMC whereas our Figure~\ref{fig:hmorll2} was produced by direct calculation.  Their Figure~2a shows that the MCMC sampler did not sample any values of $\sigssq$ less than about 10, whereas our Figure~\ref{fig:hmorll2} and their Figure~2b shows that there is a region of high $\logRLorig$ extending down to $\sigssq=0$.  In fact, $\logRLorig(500,0) \approx -1241.5$, only about 6 log units below $\logRLorig(\sshat) \approx -1235.5$.  Further, about their Figure~2a, Henn and Hodges say, ``No change in contour shape indicative of a local maximum could be found in the \dots region of $(500, 600) \times (10^{-3}, 1)$, regardless of contour resolution."  I.e., they cannot be sure there are no undiscovered points with large $\logRLorig$.  In contrast, our algorithm guarantees there are no undiscovered points where $\logRLorig$ is more than $\epsilon$ above $L$.

The HMO $\logRLorig$ analysis illustrates a typical workflow: start with a crudely determined box and coarse settings of the control parameters, then refine the box and the settings as suggested by the output of the first run.  Repeat as needed.

\subsubsection{A Bayesian Analysis}
 \cite{hodges:98}, \cite{wakefield:1998}, \cite{hodges:2013}, and \cite{henn&hodges:2014} report  Bayesian analyses of the HMO data.  Here we reproduce the analysis from \cite{hodges:98} which used inverse Gamma priors for $(\sigesq,\sigssq)$ with $\alpha_e = 1$; $\beta_e = 0$; $\alpha_s = 1.1$; and $\beta_s =0.1$.  (We don't defend the prior; we use it so we can compare to Hodges.)

For a Bayesian analysis there are 44 lines, as shown in Figure~\ref{fig:hmoBayeslines}.  Figure~\ref{fig:hmoBayeslines} differs from Figure~\ref{fig:hmolines} in that it includes a horizontal line and the position of the vertical line is slightly shifted.  Because there is only a slight change from the $\logRLorig$ analysis, we began by running the algorithm with the refined control parameters from the $\logRLorig$ analysis.  After 15 iterations there were still active boxes, so we ran the algorithm again with the same control parameters except $\maxit=20$.  Table~\ref{table:hmo_HH11Bayes} shows the output.  After 20 iterations there are over 3,000,000 boxes, over 2,000,000 of which are still active.
\begin{figure}
	\centering
	\includegraphics[width=.45\linewidth]{figs/hmolines_HH11_Bayes.pdf}
	\caption{The 44 lines, $j=1, 2, \dots, 44, v, h$, for the Bayesian analysis of the HMO data.
	              The $v$ and $h$ lines are dashed.}
	\label{fig:hmoBayeslines}
\end{figure}

\begin{table}[H]
\centering
\begin{tabular}{|c|c|c|c|}
\hline
Iteration & $n$ active boxes & $n$ inactive boxes & $L$\\
1 & 4 & 0 & $-1350.29$\\
2 & 16 & 0 & -1303.20\\
3 & 64 & 0 & -1279.93\\
4 & 256 & 0 & -1270.26\\
5 & 1024 & 0 & -1265.89\\
6 & 4060 & 9 & -1263.71\\
7 & 13740 & 634 & -1262.60\\
8 & 14904 & 10648 & -1262.01\\
9 & 10180 & 23007 & -1261.71\\
10 & 7364 & 31346 & -1261.71\\
11 & 6012 & 37207 & -1261.71\\
12 & 8192 & 41171 & -1261.71\\
13 & 16384 & 45267 & -1261.71\\
14 & 32768 & 53459 & -1261.71\\
15 & 65536 & 69843 & -1261.71\\
16 & 131072 & 102611 & -1261.71\\
17 & 262144 & 168147 & -1261.71\\
18 & 524288 & 299219 & -1261.71\\
19 & 1048576 & 561363 & -1261.71\\
20 & 2097152 & 1085651 & -1261.71\\
\hline
\end{tabular}
\caption{The state of the algorithm after each of 20 iterations for the Bayesian analysis of the HMO data.}
\label{table:hmo_HH11Bayes}
\end{table}

Figure~\ref{fig:hmoBayes} shows the boxes and $\log \pi(\sigesq,\sigssq)$.  Figure~\ref{fig:hmoBayes} is similar to Figure~\ref{fig:hmo2} except for the addition of the outlined blue boxes near $\sigssq = 0.04$.  In fact, those boxes all have lower boundaries $\sigssq = 0.04673004$, upper boundaries $\sigssq = 0.04768372$, $L^b \in (-1270.00, -1266.09)$, $U^b \in (-1252.877, -1248.524)$, and $U^b-L^b \in (17.12299, 18.01643)$, whereas the red dot shows where $\max L^b = -1261.705$ occurs.  The ridge in the posterior density near $\sigssq = 0.04$ is introduced by the $\IG(1.1,0.1)$ prior, which has a mean of 1, an infinite variance, and a peak at 0.04761905.  The usual analyses of the posterior density are carried out either by MCMC --- in which case one hopes that one's Markov chain happens to sample all regions of high density --- or by evaluating the posterior on a grid --- in which case one hopes one's grid is fine enough to capture all regions of high density.  With our analysis there is no need to hope; the algorithm is certain to find all regions of high density.

\begin{figure}
  \begin{subfigure}{.5\textwidth}
	\centering
	\includegraphics[width=.8\linewidth]{figs/hmo_HH11Bayes_boxes.jpg}
	\caption{Locations of the boxes.}
	\label{fig:hmoBayesboxes}
  \end{subfigure}
  \begin{subfigure}{.5\textwidth}
	\centering
	\includegraphics[width=.8\linewidth]{figs/hmo_HH11Bayes_rll.jpg}
	\caption{Grayscale shows $L^b$ in each box.  Red dot shows $b^* \equiv\argmax L^b$.
	Boxes outlined in blue have $U^b \ge L^{b^*}$.  Axes are logarithmic.}
	\label{fig:hmoBayesrll}
  \end{subfigure}
  \caption{The 3,182,803 boxes in the Bayesian analysis of the HMO data.
               $\maxit=20; \epsilon=1; \delta_e=0; \delta_s=0; \text{and\ } M=10$.}
  \label{fig:hmoBayes}
\end{figure}

\subsection{Global Mean Surface Temperature}
We reanalyze a data set in \cite{hodges:2013}, global mean surface temperatures (GMST) from 1881 through 2005, depicted in Figure~\ref{fig:gmst-scatter}.  
\begin{figure}
	\centering
	\includegraphics[width=.5\linewidth]{figs/gmst-scatter.pdf}
	\caption{Global mean surface temperature annually from 1881.
		The $y$-axis shows deviations from the overall mean in units of
		.01 degrees C.}
	\label{fig:gmst-scatter}
\end{figure}
As shown in \cite{ruppert_etal:2003}, many splines can be written as as linear mixed models.  \cite{hodges:2013} fit a piecewise quadratic spline to the GMST data, though a piecewise cubic spline would look similar.  Both splines can be formulated as linear mixed models.  We follow his lead in fitting a quadratic spline with knots at 1880, 1884, 1888, \dots, 2004.  $X$ has three columns: $\textcompute{1}, \textcompute{year}, \textcompute{year}^2$.  $Z$ is $125 \times 30$: one row for each year; one column for each knot.
\begin{equation*}
Z =	\begin{bmatrix}
		0 & 0 & \dots & 0\\
		0 & 0 & \dots & 0\\
		0 & 0 & \dots & 0\\
		0 & 0 & \dots & 0\\
		1 & 0 & \dots & 0\\
		4 & 0 & \dots & 0\\
		9 & 0 & \dots & 0\\
		16 & 0 & \dots & 0\\
		25 & 1 & \dots & 0\\
		\vdots & \vdots & \vdots & \vdots\\
		13924 & 12996 & \dots & 4\\
		14161 & 13225 & \dots & 9\\
		14400 & 13456 & \dots & 16\\
		14641 & 13689 & \dots & 25\\
	\end{bmatrix}; \hspace{.5cm} \Sigma_e = \mathbf{1}_{125}; \hspace{.5cm}
		\Sigma_s = \mathbf{1}_{30}
\end{equation*}
Because we fit a quadratic spline, the entries in $Z$ are squares.  $\Sigma_e$ and $\Sigma_s$ are identity matrices of the appropriate dimension.  See \cite{hodges:2013} for details.  Following Hodges, we center and scale the \textcompute{year} column of X, then compute the $\textcompute{year}^2$ column of $X$ and all the columns of Z from the transformed \textcompute{year}, so Z becomes
\begin{equation*}
Z =	\begin{bmatrix} 
		0 & 0 & \dots & 0\\
		\vdots & \vdots & \vdots & \vdots\\
		10.97143 & 10.2521 & \dots & 0.01219048\\
		11.15505 & 10.42971 & \dots & 0.01904762\\
	\end{bmatrix}
\end{equation*}
Centering and scaling changes only the scale on which $\sigssq$ is measured; we do it to more easily compare our result to Hodges'.

The column space of $Z$ shares no dimensions with the column space of $X$ so $s_z=30$ and, for our $\logRLorig$ analysis, there are 31 lines in all, for $j=1,\dots,31, v$, as shown in Figure~\ref{fig:gmst-lines}.
\begin{figure}
	\centering
	\includegraphics[width=.5\linewidth]{figs/gmst-lines.pdf}
	\caption{The 31 lines, $j=1, \dots, 30, v$, for the $\logRLorig$ analysis of
	              global mean surface temperatures.}
	\label{fig:gmst-lines}
\end{figure}
Our first run of the algorithm used the box determined by the largest intercepts of the 31 lines on the $\sigesq$ and $\sigssq$ axes and the control constants 
\begin{equation*}
	\maxit=15; \hspace{1cm} \epsilon=10; \hspace{1cm}
	\delta_e=1; \hspace{1cm} \delta_s=1; \hspace{1cm} M=10.
\end{equation*}
After 15 iterations there were 3740 boxes still in the active list and 52,973 in the inactive list; they are displayed in Figure~\ref{fig:gmst1}.

\begin{figure}
  \begin{subfigure}{.5\textwidth}
    \centering
    \includegraphics[width=.8\linewidth]{figs/gmst-boxes1.pdf}
    \caption{Locations of boxes}
    \label{fig:gmst-boxes1}
  \end{subfigure}
  \begin{subfigure}{.5\textwidth}
    \centering
    \includegraphics[width=.8\linewidth]{figs/gmst-rll1.pdf}
    \caption{Grayscale shows $L^b$ in each box.
                  Red dot shows $b^* \equiv\argmax L^b$.
	          Boxes outlined in blue have $U^b \ge L^{b^*}$.
	          Axes are logarithmic.}
    \label{fig:gmstrll1}
  \end{subfigure}
  \caption{The 56,713 boxes from the first, coarse, analysis of global
                mean surface temperature.}
\label{fig:gmst1}
\end{figure}

The figure shows that the algorithm needed to divide boxes near the axes more finely than boxes away from the axes and that high $\logRLorig$ is found in only a small region of the plot and therefore suggests rerunning the algorithm with a smaller initial box while imposing more stringent algorithmic parameters.  We limited the starting box to $\sigesq\in(50,500)$, $\sigssq\in(0,10^6)$ and set the control constants to 
\begin{equation*}
	\maxit=20; \hspace{1cm} \epsilon=2; \hspace{1cm}
	\delta_e=0; \hspace{1cm} \delta_s=0; \hspace{1cm} M=10.
\end{equation*}
After 20 iterations there were 1,986,000 active and 1,619,095 inactive boxes.  Results are in Figure~\ref{fig:gmst2}.  The figure agrees with Figure~15.3 in \cite{hodges:2013} and is detailed enough so no further iterations are needed.

\begin{figure}
  \begin{subfigure}{.5\textwidth}
    \centering
    \includegraphics[width=.8\linewidth]{figs/gmst-boxes2.jpg}
    \caption{Locations of boxes}
    \label{fig:gmst-boxes2}
  \end{subfigure}
  \begin{subfigure}{.5\textwidth}
    \centering
    \includegraphics[width=.8\linewidth]{figs/gmst-rll2.jpg}
    \caption{Grayscale shows $L^b$ in each box.
                  Red dot shows $b^* \equiv\argmax L^b$.
	          Boxes outlined in blue have $U^b \ge L^{b^*}$.
	          Axes are logarithmic.}
    \label{fig:gmstrll2}
  \end{subfigure}
  \caption{The 3,605,095 boxes from the second analysis of global
                mean surface temperature.}
\label{fig:gmst2}
\end{figure}

\section{Discussion}
This paper has explained and illustrated an algorithm, for linear mixed models with two variances, to find all regions where either the restricted likelihood function or the joint posterior density of the variances is high, and to evaluate the function there to arbitrary accuracy.  A natural question to ask is \emph{What about linear mixed models with more than two variances?}  A partial answer is given by \cite{hodges:2013} who shows that some models with more than two variances can be reexpressed similarly to \eqref{eq:reexpress} but others can't.  More complex models that can be reexpressed this way include but are probably not limited to models displaying general balance that are also orthogonal designs \citep[all balanced ANOVAs plus other models;][]{houtman_speed:1983}, models that are separable in a specific sense \citep[Section 17.1.5]{hodges:2013}, and miscellaneous other models \citep[Section 17.1.5]{hodges:2013}, e.g., a spatial model including random effects for heterogeneity and spatial clustering (an improper conditional autoregressive effect). We have not explored whether the reexpressible models can be analyzed by our algorithm; that's one direction for future work.

Another is to see whether the algorithm can be used to advantage even in non-reexpressible models.  If a model has, say, three variances and is now analyzed by, say, MCMC, we can create an MCMC chain that alternates between draws of $(\sigesq,\sigssq)$ and draws of the other variance.  With the aid of our algorithm we may be able to draw more accurately from $[\sigesq,\sigssq\g \sigma^2_\text{other}]$.  More generally, the conditional distribution of $(\sigesq,\sigssq)$ given other parameters can now be analyzed more accurately than in the past.  We have yet to explore how to exploit that accuracy.  A third direction is the posterior $\pi(\sigesq,\sigssq\g y)$.  We can identify a region $B^c$ where the posterior density is low relative to its maximum and it would be of at least mild interest to find an upper bound for the posterior mass of $B^c$.

As written, our algorithm moves a box $b$ to the inactive list if
\begin{enumerate}[(a)]
\item $U^b < L-M$ or
\item $U^b - L^b < \epsilon$ or
\item either the vertical or horizontal extent of $b$ is sufficiently small.
\end{enumerate}
But one could construct more elaborate rules.  One appealing example is to apply criteria (a) and (c) if $U^b \le L-\epsilon_2$ and apply criterion (b) if $U^b > L-\epsilon_2$.  Other rules are possible, too.  We don't elaborate here in order to concentrate on the main ideas.

In this paper we have taken the point of view that it is important to find all regions where $\log f$ is large without necessarily identifying all local maxima or even the global maximum, even though that point of view is at odds with common statistical estimators that maximize the likelihood, the restricted likelihood, or the posterior density.  If two local maxima are close in height it hardly matters which is slightly higher than the other.  And, as we said earlier, if there is a high plateau it hardly matters whether there are little bumps on that plateau.

\bibliographystyle{jasa}
\bibliography{paper}

\end{document}

### 2015-05-12 ms.tex
added Bray as author.  Added appendix

@@
 \usepackage{natbib}
-\usepackage{paralist}
+\usepackage{enumitem}
+%\usepackage{appendix}
 %\usepackage{setspace}
@@
 \author[1]{Michael Lavine}
-\author[2]{James Hodges}
+\author[1,2]{Andrew Bray}
+\author[3]{James Hodges}
 \affil[1]{Department of Mathematics and Statistics, University of Massachusetts, Amherst, MA 01003 USA}
-\affil[2]{Division of Biostatistics, University of Minnesota, Minneapolis, MN 55455 USA}
+\affil[2]{Mount Holyoke College}
+\affil[3]{Division of Biostatistics, University of Minnesota, Minneapolis, MN 55455 USA}
 
@@
 \end{equation}
-where $B$ is an unimportant known constant; $n_e$ is $n$ minus the dimension of the space spanned by the columns of $[X|Z]$; $\Gamma_c$ is $n \times n_e$ and spans the space orthogonal to $[X|Z]$ (so $y^t \Gamma_c \Gamma^t_c y$ is the residual sum of squares); $s_z$ is the dimension of the space spanned by the columns of $Z$ not already in the span of the columns of $X$; and the $\{a_j\}$ and $\{\hat v_j\}$ are known constants with all $a_j>0$.
+where
+\begin{enumerate}[label=(\arabic*)]
+  \item $B$ is an unimportant known constant;
+  \item $n_e$ is $n$ minus the dimension of the space spanned by the columns of $[X|Z]$;
+  \item $\Gamma_c$ is $n \times n_e$ and spans the space orthogonal to $[X|Z]$
+    (so $y^t \Gamma_c \Gamma^t_c y$ is the residual sum of squares);
+  \item $s_z$ is the dimension of the space spanned by the columns of $Z$ not already
+    in the span of the columns of $X$; and
+  \item the $\{a_j\}$ and $\{\hat v_j\}$ are known constants whose derivation is in the Appendix.  All $a_j>0$.
+\end{enumerate}
 Thus the only unknowns are $(\sigma^2_s, \sigma^2_e)$ and $\logRLssorig$ is a function of just those two arguments.
@@
 \item Specify constants \maxit, $M$, $\epsilon$, $\delta_e$, and $\delta_s$.
-	\begin{enumerate}[(a)]
+	\begin{enumerate}[label=(\alph*)]
 	\item \maxit\ is the maximum number (could be $\infty$) of iterations of
@@
 As written, our algorithm moves a box $b$ to the inactive list if
-\begin{enumerate}[(a)]
+\begin{enumerate}[label=(\alph*)]
 \item $U^b < L-M$ or
@@
 
+\section{Appendix}
+\paragraph{Derivation of $\{a_j\}$ and $\{\hat v_j\}$ in \eqref{eq:reexpress}}  Our derivation follows \cite{hodges:2013},
+which contains more details.  There are three steps.
+\begin{enumerate}
+  \item \textbf{Make the covariance matrices proportional to the identity.}  If $\Sigma_e$ is not the identity matrix,
+    transform the data to $\Sigma_e^{-.5}y$.  The transformed data, which we shall still call $y$, has covariance
+    proportional to the identity.  Similarly, if $\Sigma_s$ is not the identity matrix, re-parameterize the random
+    effects to $\Sigma_s^{-.5}u$.  The re-parameterized random effects, which we shall still call $u$, have
+    covariance proportional to the identity.
+  \item \textbf{If the column spaces of $X$ and $Z$ have a non-trivial intersection, transform them.}
+    Let $s_X=\text{rank}(X)$  and $s_Z=\text{rank}(X|Z) - s_X$.  Let $\Gamma_X$ be an $n \times s_X$ matrix whose
+    columns are an orthonormal basis for the column space of $X$.  Let $\Gamma_Z$ be an $n \times s_Z$ matrix
+    such that the columns of $\left[\Gamma_X | \Gamma_Z \right]$ are an orthonormal basis for the column space
+    of $\left[X|Z\right]$.  Let $\Gamma_c$ be an $n \times n-s_X-s_Z$ matrix such that the columns of
+    $\left[\Gamma_X | \Gamma_Z | \Gamma_c\right]$ are an orthonormal basis for $\mathbb{R}^n$.  Define the matrix
+    \begin{equation*}
+      M = \begin{bmatrix}
+               M_{XX} & M_{XZ}\\
+               0 & M_{ZZ}
+             \end{bmatrix}
+    \end{equation*}
+    by $[X|Z] = [\Gamma_X|\Gamma_Z]M$ where $M_{XX}$ is $s_X \times p$ and $M_{XZ}$ is $s_Z \times q$.
+    $\Gamma_X$ and $\Gamma_Z$ are transformed versions of $X$ and $Z$ that have non-overlapping
+    column spaces.
+  \item \textbf{Re-parameterize and diagonalize.}  Let $M_{ZZ}$ have the singular value decomposition
+    $PA^{.5}L^t$.  Now the linear mixed model \eqref{eq:lmm} can be written as
+    \begin{equation*}
+      \begin{split}
+        y &= \begin{bmatrix} X Z \end{bmatrix} \begin{bmatrix} \beta \\ u \end{bmatrix} + \epsilon\\
+        &= \begin{bmatrix} \Gamma_X  \Gamma_Z \end{bmatrix} M \begin{bmatrix} \beta \\ u \end{bmatrix} + \epsilon\\
+        &= \begin{bmatrix} \Gamma_X \Gamma_ZP \end{bmatrix} \begin{bmatrix} \beta^* \\ v \end{bmatrix} + \epsilon
+      \end{split}
+    \end{equation*}
+    where $\beta^* = M_{XX}\beta + M_{XZ}u$ and $v = A^{.5}L^t u$.  $\beta^*$ contains the re-parametrized
+    fixed effects while $v$ contains the re-parametrized random effects.  The corresponding design matrices
+    $\Gamma_X$ and $\Gamma_Z P$ are orthogonal to each other.
+\end{enumerate}
+Finally, the $\{a_j\}$ in \eqref{eq:reexpress} are the diagonal elements of $A$, all of which are strictly positive, and the $\{\hat v_j\}$ in \eqref{eq:reexpress} are given by $\hat v = (\hat v_1, \dots, \hat v_{s_Z})^t = P^t \Gamma_Z^t y$.
+
 \bibliographystyle{jasa}

### 2015-05-12 ms.tex
clarified Eq 3, 4, 5

@@
     &\qquad\qquad - \frac{2\alpha_s + 2} {2}\log(\sigssq) - \frac{2\beta_s}{2\sigssq}
-    - \frac{1}{2} \sum_{j=1}^{s_z} \left[ \log(\ass) + \frac{\hat v_j^2}{\ass}\right]\\
-    &= B - \frac{1}{2}\sum_j\left[ c_j \log(\abss) + \frac{d_j}{\abss}\right]
+    - \frac{1}{2} \sum_{j=1}^{s_z} \left[ \log(\ass) + \frac{\hat v_j^2}{\ass}\right].
   \end{split}
 \end{equation}
-for known nonnegative constants $\{a_j, b_j, c_j, d_j\}$ ---
-still a linear combination of logs and inverses of linear functions $\abss$.
-Our derivation will proceed from \eqref{eq:logpost}, though it applies also to \eqref{eq:reexpress} and other linear combinations of logs and inverses of linear functions $\abss$.
-We use $f$ to denote such functions generically.
+Equations~\eqref{eq:reexpress} and~\eqref{eq:logpost} can both be written as in \eqref{eq:sum}: a summation of multiples of logs and inverses of linear combinations $\abss$, where the summands with $j=s_Z+1$ and $j=s_Z+2$ are for the terms involving only $\sigesq$ and $\sigssq$, respectively.
+\begin{equation}
+\label{eq:sum}
+  -\frac{1}{2} \sum_{j=1}^{s_Z+2}\left[ c_j \log(\abss) + \frac{d_j}{\abss}\right]
+\end{equation}
+where
+\begin{itemize}[label=---]
+\item for $j = 1, \dots, s_Z$,
+  \begin{align*}
+    a_j &> 0\ \text{and is derived in the Appendix}\\
+    b_j &=1\\
+    c_j &=1\\
+    d_j &= \hat v_j^2\ \text{and is derived in the Appendix}
+  \end{align*}
+\item for $j = s_Z +1$,
+  \begin{align*}
+    a_j &= 0\\
+    b_j &= 1\\
+    c_j &= \begin{cases}
+                 n_e & \text{for $\logRLorig$ in \eqref{eq:reexpress}}\\
+                 n_e + 2\alpha_e + 2 & \text{for the $\log$ posterior in \eqref{eq:logpost}}
+              \end{cases}\\
+    d_j &= \begin{cases}
+                 \frac{y^t \Gamma_c \Gamma^t_c y}{\sigesq} & \text{for $\logRLorig$ in \eqref{eq:reexpress}}\\
+                 \frac{y^t \Gamma_c \Gamma^t_c y + 2\beta_e}{\sigesq} & \text{for the $\log$ posterior in \eqref{eq:logpost}}
+              \end{cases}
+  \end{align*}
+\item for $j = s_Z+2$,
+  \begin{align*}
+    a_j &= 1\\
+    b_j &= 0\\
+    c_j &= \begin{cases}
+      0 & \text{for $\logRLorig$ in \eqref{eq:reexpress}}\\
+      2 \alpha_s + 2 & \text{for the $\log$ posterior in \eqref{eq:logpost}}
+    \end{cases}\\
+    d_j &= \begin{cases}
+      0 & \text{for $\logRLorig$ in \eqref{eq:reexpress}}\\
+      2 \beta_s & \text{for the $\log$ posterior in \eqref{eq:logpost}}.
+    \end{cases}
+  \end{align*}
+\end{itemize}
+Our derivation proceeds from \eqref{eq:sum}.  We use $f$ to denote the target function generically, either~\eqref{eq:reexpress} or~\eqref{eq:logpost}.
 

### 2015-05-13 gmst.R
Initial commit.

@@ -1,0 +1,138 @@
+# global mean surface temperature
+
+library(nlme) # for the standard analysis, by REML
+library(devtools)
+devtools::install_github("andrewpbray/lmmoptim")
+library("lmmoptim")
+
+data(gmst)
+y <- gmst$temp.dev
+
+# plot the data
+pdf("gmst-scatter.pdf")
+p <- ggplot(data = gmst, aes(x = Year, y = temp.dev)) + geom_point()
+p + ylab("temperature deviation") + theme_bw()
+dev.off()
+
+# Hodges' code for centered X, Z, GamX, GamZ
+x <- 1:125
+xs <- (x - mean(x))/sqrt(var(x))
+X <- matrix(c(rep(1, 125), xs, xs^2), 125, 3)
+knots <- seq(4, 121, 4)
+ks <- (knots - mean(x))/sqrt(var(x))
+kp <- matrix(0 == 1, 125, 30)
+for (i in 1:30) {
+  for (j in 1:125) {
+    kp[j, i] <- xs[j] >= ks[i]
+    }
+  }
+Z <- kp * (matrix(xs, 125, 30) - matrix(ks, 125, 30, byrow = T))^2
+GX <- svd(X)$u
+CX <- t(svd(cbind(X,Z))$u) %*% GX
+QCX <- diag(rep(1, 33)) - CX %*% solve(t(CX) %*% CX) %*% t(CX)
+CZ <- eigen(QCX, sym = T)$vec[, 1:30]
+GZ <- svd(cbind(X, Z))$u %*% CZ
+x <- X
+z <- Z
+
+# Lavine's code for uncentered  X, Z, GamX, GamZ
+yr <- gmst$year
+nyr <- length(yr)
+knots <- 1880 + seq(4, 120, by = 4)
+x <- cbind(rep(1, nyr), yr, yr^2)
+# x <- cbind ( rep(1,nyr), scale ( x[,-1] ) ) # either use either this line or not.  it shouldn't matter
+z <- pmax(outer(yr, knots, "-" ), 0)^2
+
+# Compute H_ZZ and PDP';  Compute D and canon regressors GZ P
+
+# HZZ <- t(GZ) %*% Z
+# P <- eigen(HZZ %*% t(HZZ),sym=T)$vec
+# D <- eigen(HZZ %*% t(HZZ),sym=T)$val
+
+# D matches previous work!
+
+# > D
[…]

### 2015-05-13 hmo.R

@@ -1,0 +1,208 @@
+# HMO Example
+library(dplyr)
+library(nlme) # the standard analysis, by REML
+library(devtools)
+devtools::install_github("andrewpbray/lmmoptim")
+library("lmmoptim")
+
+data(hmo)
+
+mod <- lme(indPrem ~ 1, data = hmo, random = ~1 | state, control = list(apVar=TRUE))
+summary(mod)
+
+# random intercept model
+y <- hmo$indPrem
+n <- length(y)
+mod <- lm(indPrem ~ state, data = hmo, x = TRUE)
+x <- mod$x[, 1, drop = FALSE]
+z <- mod$x[, -1, drop = FALSE]
+SigE <- diag(n)
+SigS <- diag(44)
+
+lines <- findlines(x, z, y, SigE, SigS)
+
+pdf("hmolines.pdf")
+showlines(lines)
+dev.off()
+
+mlrebox <- with(lines,
+                makebox(lims.sigsqs = c(0, max(int.sigsqs[is.finite(int.sigsqs)])),
+	                      lims.sigsqe = c(0, max(int.sigsqe[is.finite(int.sigsqe)])),
+	                      status = rep("straddle", nrow(lines)),
+                        lines = lines))
+
+boxes <- fitlmm(lines = lines, mlrebox, eps = 1, delE = 10, delS = 10, M = 5, maxit = 10)
+p <- showboxes(boxes)
+
+pdf("manuscript/figs/hmoboxes.pdf")
+p + theme_bw()
+dev.off()
+jpeg("manuscript/figs/hmoboxes.jpg")
+p + theme_bw()
+dev.off()
+
+p <- showfunc(boxes)
+pdf("manuscript/figs/hmorll.pdf")
+p + scale_x_log10 () + scale_y_log10 () + theme_bw()
+dev.off()
+jpeg("manuscript/figs/hmorll.jpg")
+p + scale_x_log10 () + scale_y_log10 () + theme_bw()
+dev.off()
+
+# the previous figure suggests we can limit our attention to a smaller region
+box2 <- with(lines,
+             makebox(lims.sigsqe = c(0, 1000),
+                     lims.sigsqs = c(0, 2000),
+                     status = rep("straddle", nrow(lines)),
+                     lines = lines))
+
+boxes2 <- fitlmm(lines = lines, box2, eps = .5, delE = 0, delS = 0, ratio = TRUE,
[…]

### 2015-05-13 gmst.R
Update fig dir path.

@@ -85,7 +85,7 @@ SigS <- diag(length(knots))
 lines <- findlines(x, z, y, SigE, SigS)
 
 # plot the lines
-pdf("gmst-lines.pdf")
+pdf("figs/gmst-lines.pdf")
 showlines(lines)
 dev.off()
 
@@ -95,7 +95,7 @@ dev.off()
 boxes <- fitlmm(lines = lines, eps = 10, delE = 1, delS = 1, M = 10, maxit = 15) # first try
 
 # view the boxes
-pdf("gmst-boxes1.pdf")
+pdf("figs/gmst-boxes1.pdf")
 p <- showboxes(boxes)
 p + scale_x_continuous(name = expression(sigma[e]^2)) +
     scale_y_continuous(name = expression(sigma[s]^2)) +
@@ -103,7 +103,7 @@ p + scale_x_continuous(name = expression(sigma[e]^2)) +
 dev.off()
 
 p <- showfunc(boxes)
-pdf("gmst-rll1.pdf")
+pdf("figs/gmst-rll1.pdf")
 p + scale_x_log10(name = expression(sigma[e]^2),
                   breaks = c(30, 300, 3000)) +
     scale_y_log10(name = expression(sigma[s]^2),
@@ -121,14 +121,14 @@ boxes <- fitlmm(lines = lines, startbox = startbox, eps = 2, M = 10, maxit = 20,
                 delE = 0, delS = 0)
 
 # view the boxes
-jpeg("gmst-boxes2.jpg")
+jpeg("figs/gmst-boxes2.jpg")
 p <- showboxes(boxes)
 p + scale_x_continuous(name = expression(sigma[e]^2)) +
     scale_y_continuous(name = expression(sigma[s]^2)) +
     theme_bw()
 dev.off()
 
-jpeg("gmst-rll2.jpg")
+jpeg("figs/gmst-rll2.jpg")
 p <- showfunc(boxes)
 p + scale_x_log10(name = expression(sigma[e]^2),
                   breaks = c(30, 100, 300)) +

### 2015-05-13 hmo.R

@@ -34,18 +34,18 @@ mlrebox <- with(lines,
 boxes <- fitlmm(lines = lines, mlrebox, eps = 1, delE = 10, delS = 10, M = 5, maxit = 10)
 p <- showboxes(boxes)
 
-pdf("manuscript/figs/hmoboxes.pdf")
+pdf("figs/hmoboxes.pdf")
 p + theme_bw()
 dev.off()
-jpeg("manuscript/figs/hmoboxes.jpg")
+jpeg("figs/hmoboxes.jpg")
 p + theme_bw()
 dev.off()
 
 p <- showfunc(boxes)
-pdf("manuscript/figs/hmorll.pdf")
+pdf("figs/hmorll.pdf")
 p + scale_x_log10 () + scale_y_log10 () + theme_bw()
 dev.off()
-jpeg("manuscript/figs/hmorll.jpg")
+jpeg("figs/hmorll.jpg")
 p + scale_x_log10 () + scale_y_log10 () + theme_bw()
 dev.off()
 
@@ -62,18 +62,18 @@ boxes2 <- fitlmm(lines = lines, box2, eps = .5, delE = 0, delS = 0, ratio = TRUE
 mrle <- which.max(boxes2$rll.lower)
 
 p <- showboxes(boxes2)
-pdf("manuscript/figs/hmoboxes2.pdf")
+pdf("figs/hmoboxes2.pdf")
 p + theme_bw()# + scale_x_log10 () + scale_y_log10 ()
 dev.off()
-jpeg("manuscript/figs/hmoboxes2.jpeg")
+jpeg("figs/hmoboxes2.jpeg")
 p + theme_bw()# + scale_x_log10 () + scale_y_log10 ()
 dev.off()
 
 p <- showfunc(boxes2)
-pdf("manuscript/figs/hmorll2.pdf")
+pdf("figs/hmorll2.pdf")
 p + scale_x_log10 () + scale_y_log10 () + theme_bw()
 dev.off()
-jpeg("manuscript/figs/hmorll2.jpeg")
+jpeg("figs/hmorll2.jpeg")
 p + scale_x_log10 () + scale_y_log10 () + theme_bw()
 dev.off()
 
@@ -108,7 +108,7 @@ summary(mod)
 
 lines <- findlines(x, z, y, SigE, SigS)
 
-pdf("manuscript/figs/hmolines.HH11.pdf")
+pdf("figs/hmolines.HH11.pdf")
 showlines(lines)
 dev.off()
 
[…]

### 2015-05-14 gmst.R
Fix fig path.

@@ -9,7 +9,7 @@ data(gmst)
 y <- gmst$temp.dev
 
 # plot the data
-pdf("gmst-scatter.pdf")
+pdf("figs/gmst-scatter.pdf")
 p <- ggplot(data = gmst, aes(x = Year, y = temp.dev)) + geom_point()
 p + ylab("temperature deviation") + theme_bw()
 dev.off()

### 2015-05-16 ms.tex
Add subdepth package to fix vertical alignment of sub- and superscripts; change /abss to /ass; fix eqn 5, the values of dj for sz + 1.

@@
 \usepackage{enumitem}
+\usepackage{subdepth}
 %\usepackage{appendix}
@@
 Thus the only unknowns are $(\sigma^2_s, \sigma^2_e)$ and $\logRLssorig$ is a function of just those two arguments.
-Specifically, $\logRLssorig$ is a linear combination of logs and inverses of linear functions $\abss$ of the two unknowns.
+Specifically, $\logRLssorig$ is a linear combination of logs and inverses of linear functions $\ass$ of the two unknowns.
 
@@
     d_j &= \begin{cases}
-                 \frac{y^t \Gamma_c \Gamma^t_c y}{\sigesq} & \text{for $\logRLorig$ in \eqref{eq:reexpress}}\\
-                 \frac{y^t \Gamma_c \Gamma^t_c y + 2\beta_e}{\sigesq} & \text{for the $\log$ posterior in \eqref{eq:logpost}}
+                 y^t \Gamma_c \Gamma^t_c y & \text{for $\logRLorig$ in \eqref{eq:reexpress}}\\
+                 y^t \Gamma_c \Gamma^t_c y + 2\beta_e & \text{for the $\log$ posterior in \eqref{eq:logpost}}
               \end{cases}

### 2015-05-17 ms.tex
edited proof of bounding box B

--- 2015-05-12 ms.tex
@@
 Thus the only unknowns are $(\sigma^2_s, \sigma^2_e)$ and $\logRLssorig$ is a function of just those two arguments.
-Specifically, $\logRLssorig$ is a linear combination of logs and inverses of linear functions $\abss$ of the two unknowns.
 
@@
 \end{equation}
-Equations~\eqref{eq:reexpress} and~\eqref{eq:logpost} can both be written as in \eqref{eq:sum}: a summation of multiples of logs and inverses of linear combinations $\abss$, where the summands with $j=s_Z+1$ and $j=s_Z+2$ are for the terms involving only $\sigesq$ and $\sigssq$, respectively.
+Equations~\eqref{eq:reexpress} and~\eqref{eq:logpost} can both be written as a sum of multiples of logs and inverses of linear combinations $\abss$, as in \eqref{eq:sum}, where the summands with $j=s_z+1$ and $j=s_z+2$ are for the terms involving only $\sigesq$ and $\sigssq$, respectively.
 \begin{equation}
 \label{eq:sum}
-  -\frac{1}{2} \sum_{j=1}^{s_Z+2}\left[ c_j \log(\abss) + \frac{d_j}{\abss}\right]
+  -\frac{1}{2} \sum_{j=1}^{s_z+2}\left[ c_j \log(\abss) + \frac{d_j}{\abss}\right]
 \end{equation}
@@
 \begin{itemize}[label=---]
-\item for $j = 1, \dots, s_Z$,
+\item for $j = 1, \dots, s_z$,
   \begin{align*}
@@
   \end{align*}
-\item for $j = s_Z +1$,
+\item for $j = s_z +1$,
   \begin{align*}
@@
   \end{align*}
-\item for $j = s_Z+2$,
+\item for $j = s_z+2$,
   \begin{align*}
@@
 \end{itemize}
-Our derivation proceeds from \eqref{eq:sum}.  We use $f$ to denote the target function generically, either~\eqref{eq:reexpress} or~\eqref{eq:logpost}.
+Our derivation proceeds from \eqref{eq:sum}.
+We use $f$ to denote the target function generically, either~\eqref{eq:reexpress} or~\eqref{eq:logpost}.
+When the target function is~\eqref{eq:reexpress}, $c_{s_z+2} = d_{s_z+2} = 0$ and the upper limit of the sum in \eqref{eq:sum} is $s_z+1$.
 
@@
 can have multiple maxima and that existing general purpose algorithms for linear mixed models may fail to find all of them, as shown by examples in \cite{hodges:2013}, \cite{henn&hodges:2014}, and elsewhere.
-\cite{mullen:2014} examines 18 optimization functions available in \R, tests them on 48 objective functions (admittedly more complicated than $\logRLss$) and finds that even the best of them fail in over 10\% of the cases.
+\cite{mullen:2014} examines 18 optimization functions available in \R; tests them on 48 objective functions, admittedly more complicated than $\logRLss$; and finds that even the best of them fail in over 10\% of the cases.
 \cite{henn&hodges:2014} examine conditions under which multiple maxima occur in posterior densities and conclude ``\dots second maxima in posterior distributions therefore may be more common than reports in the literature would suggest."  Thus, failure to find local and global maxima may be common.
@@
 \end{subequations}
+
 We work with one term in (\ref{eq:partials})'s summations at a time; that is, one $j$ at a time.
-The $j$'th terms in (\ref{eq:partials}) differ only by a multiplicative constant $a_j/b_j$; they have the same sign as each other and the same sign as $(a_jc_j\sigssq + b_jc_j\sigesq  - d_j)$, which determines a line $\sigssq = d_j/a_j c_j - b_j \sigesq/a_j$ --- call it the $j$'th line --- in the first quadrant of the $(\sigssq,\sigesq)$ plane.
-The line has positive intercept and negative slope because all the constants are nonnegative.
-Both partial derivatives of the $j$'th term are positive below the $j$'th line, 0 on the line, and negative above the line, as indicated in Figure~\ref{fig:oneline}.
+For $j=1, \dots, s_z$, the $j$'th terms in (\ref{eq:partials}) differ  by a multiplicative constant $a_j/b_j = a_j$; they have the same sign as each other and the same sign as $(a_jc_j\sigssq + b_jc_j\sigesq  - d_j) = (a_j\sigssq + \sigesq  - d_j)$, which determines a line $\sigssq = d_j/a_j - \sigesq/a_j$ --- call it the $j$'th line --- in the first quadrant of the $(\sigssq,\sigesq)$ plane.
+The $j$'th line has positive intercept $ d_j/a_j$ and negative slope $-1/a_j$.
+For $j=s_z+1$, $a_j=0$, so $\eqref{eq:drlds} = 0$ and \eqref{eq:partials} determines a vertical line at $\sigesq = \sigma_e^{2*} \equiv d_{s_z+1} / c_{s_z+1}$.
+For $j=s_z+2$,  $b_j=0$, so $\eqref{eq:drlde} = 0$ and, if the target function is \eqref{eq:logpost}, \eqref{eq:partials} determines a horizontal line at $\sigssq = \sigma_s^{2*} \equiv d_{s_z+2} / c_{s_z+2}$.
+
+For all $j$, both partial derivatives of the $j$'th term are nonnegative below or to the left of the $j$'th line, 0 on the line, and nonpositive above or to the right of the line, as indicated in Figure~\ref{fig:oneline}.
 The $j$'th term is constant on the $j$'th line and attains its maximum there.
-Both \eqref{eq:reexpress} and \eqref{eq:logpost} contain a term, call it the $v$'th (for vertical) term,  with $a_j=0$.
-The corresponding line is vertical at $\sigesq = \sigma_e^{2v} \equiv d_v / b_v c_v$.
+%Both \eqref{eq:reexpress} and \eqref{eq:logpost} contain a term, call it the $v$'th (for vertical) term,  with $a_j=0$.
 The $v$'th term in (\ref{eq:partials}) is negative to the right of the $v$'th line, positive to the left, and 0 on the line; hence the $v$'th term in \eqref{eq:logpost} is maximized on the $v$'th line.
@@
 	\includegraphics[width=.5\linewidth]{figs/oneline.pdf}
-	\caption{For a single summand, i.e., a fixed $j$, in (\ref{eq:logpost}),
+	\caption{For a single summand, i.e., a fixed $j$, in \eqref{eq:sum},
 	              the partial derivatives (\ref{eq:drlds}) and (\ref{eq:drlde})
-	              are 0 on the line and positive and negative where indicated
-	              by ``\textbf{+}" and ``\textbf{-}".}
+	              are 0 on the line and nonnegative and nonpositive
+	              where indicated by ``\textbf{+}" and ``\textbf{-}".}
 	\label{fig:oneline}
@@
 
-Figure~(\ref{fig:bigboundingbox}) shows five lines --- realistic data sets may have more --- labelled $j=1, 2, 3, v, h$.
+Figure~(\ref{fig:bigboundingbox}) shows five lines --- realistic data sets may have more --- labelled $j=1, 2, 3, s_z+1, s_z+2$.
 The largest intercept of those lines on the $\sigssq$ axis and the largest intercept on the $\sigesq$ axis determine a rectangle $B_1$ whose outline is marked in bold.
-At any point above $B_1$, e.g., $p_1$, the partial derivatives \eqref{eq:partials}, for all $j$, are negative or, for $j=v$, zero.
-Therefore, $\logRL(p_1^*) \ge \logRL(p_1)$.
-Similarly, for points like $p_2$, $\logRL(p_2) \ge \logRL(p_2^*)$ and, for points like $p_3$, $\logRL(p_3^*) \ge \logRL(p_3)$.
+At any point above $B_1$, e.g., $p_1$, the partial derivatives \eqref{eq:partials} are either negative or, for $j=s_z+1$, zero.
+Therefore, $\logRL(\widetilde{p_1}) \ge \logRL(p_1)$.
+Similarly, $\logRL(\widetilde{p_2}) \ge \logRL(p_2)$ and $\logRL(\widetilde{p_3}) \ge \logRL(p_3)$.
 That is, for points above and/or to the right of $B_1$, $\logRLss$ can be increased by moving down and/or to the left as far as $B_1$'s boundary.
 Therefore \mrle\ must lie on or inside $B_1$ and there can be no maxima of $\logRLss$ in $B_1^c$.
-$B_1$ could be passed to an optimizer such as \R's \textsf{optim} or \textsf{nlminb}, with potentially better results than using those functions without bounds.
+$B_1$ could be passed to an optimizer such as \R's \textsf{optim} or \textsf{nlminb} with potentially better results than using those functions without bounds.
 However, even with known bounds, general purpose optimizers may still miss \mrle\ and regions of high $\logRL$.
@@
 	\includegraphics[width=.8\linewidth]{figs/boundingbox.pdf}
-	\caption{A rectangular region containing all maxima.}
+	\caption{A rectangular region $B_1$ containing all maxima.}
 	\label{fig:bigboundingbox}
@@
 
-To find $B$, choose an arbitrary constant $M>0$ and an arbitrary point inside $B_1$, say $p_4$, as illustrated in Figure~(\ref{fig:boundingbox2}) and define $L \equiv \logRL(p_4)$.
+To find $B$, choose an arbitrary constant $M>0$ and an arbitrary point inside $B_1$, say $p$, as illustrated in Figure~(\ref{fig:boundingbox2}) and define $L \equiv \logRL(p)$.
 We will find a box $B\supset B_1$, as shown in Figure~(\ref{fig:boundingbox2}), outside of which $\logRLss \le L-M$, thus satisfying \textbf{D1}.
-We have drawn a figure without an $h$ line so the reader can see  the derivation without it, which would apply to $\logRLorig$.
-If there were an $h$ line, it would influence the derivation for Figure~(\ref{fig:boundingbox2})'s $q_1$ in a manner analogous to how the $v$ line influences the argument for $q_2$ and $q_3$.
 
-$B$ is determined by its intercepts $\sigma_e^{2*}$ and $\sigma_s^{2*}$ on the $\sigesq$ and $\sigssq$ axes respectively.
-Let $q_1^* = (\sigma_e^{2*},0)$ be the intercept of $B$ and the $\sigesq$ axis.
-Because each term of the partial derivatives \eqref{eq:partials} is nonpositive to the right of $B_1$, for any point $q_1$ to the right of $B$, every term in the summation of \eqref{eq:logpost} (with the possible exception of an $h$ term) is larger at $q_1^*$ than at $q_1$.
-Therefore, $\logRL(\sshat) - \logRL(q_1) \ge L - \logRL(q_1^*)$.
-The latter expression is greater than $M$ iff $\logRL(q_1^*) < L-M$.
-But examination of (\ref{eq:logpost}) shows that for any fixed $\sigssq$, in particular for $\sigssq=0$, $\lim_{\sigesq \rightarrow \infty} \logRLss = -\infty$.
-Thus by choosing $\sigma_e^{2*}$ large enough so $\logRL(q_1^*) < L-M$, we satisfy (\ref{eq:boundingbox}) for all points with $\sigesq \ge \sigma_e^{2*}$.
+$B$ is determined by its intercepts $\widetilde{\sigma_e^2}$ and $\widetilde{\sigma_s^2}$ on the $\sigesq$ and $\sigssq$ axes respectively.
+Let $\widetilde{q_1} = (0,\widetilde{\sigma_s^2})$ be the intercept of $B$ and the $\sigssq$ axis and let $\widetilde{q_2} = (\sigma_e^{2*},\widetilde{\sigma_s^2})$.
+Points such as $q_1$ which satisfy $\sigesq \le \sigma_e^{2*}$ and $\sigssq \ge \widetilde{\sigma_s^2}$ --- above $B$ and to the left of the $s_z+1$ line --- are in a region where the partial derivatives (\ref{eq:drlds}, \ref{eq:drlde}) are negative for $j \in \{1, \dots, s_z, s_z+2\}$ but nonnegative for $j=s_z+1$.
+Let $\logRL_j(\cdot)$ denote the $j$'th term of \eqref{eq:sum} evaluated at $(\cdot)$.
+\begin{equation}
+	\logRL(\sshat) - \logRL(q_1) \ge L - \sum_{j \ne s_z+1} \logRL_j(\widetilde{q_1}) - \logRL_{s_z+1}(\widetilde{q_2})
+\end{equation}
+which is larger than $M$ if
+\begin{equation}
+\label{eq:q1}
+  \begin{split}
+    \sum_{j \ne s_z+1} \logRL_j(\widetilde{q_1})
+      &< L - M - \logRL_{s_z+1}(\widetilde{q_2})\\
+      &= L - M + \frac{1}{2} \left[ c_{s_z+1}\log\sigma_e^{2*} + \frac{d_{s_z+1}}{\sigma_e^{2*}}\right].
+  \end{split}
+\end{equation}
+Examination of (\ref{eq:sum}) shows that for any fixed $\sigesq$, in particular for $\sigesq=0$, and for $j \ne s_z+1$,\\ $\lim_{\sigssq \rightarrow \infty} \logRL_j(\sigesq,\sigssq) = -\infty$.
+Thus $\widetilde{\sigma_s^2}$ can be chosen large enough so that the summation on the l.h.s.~of \eqref{eq:q1} is less than the r.h.s.~of \eqref{eq:q1}, and  \textbf{D1} is satisfied.
 
-Let $q_2^* = (\sigma_e^{2v},\sigma_s^{2*})$.
-For any point $q_2$ having $\sigssq \ge \sigma_s^{2*}$ and $\sigesq \in [\sigma_e^{2v}, \sigma_e^{2*}]$ --- i.e., above $B$ and to the right of the $v$'th line --- the partial derivatives are nonpositive so $\logRL(\sshat) - \logRL(q_2) \ge L - \logRL(q_2^*)$, which is larger than $M$ if $\logRL(q_2^*) < L-M$.
-But for any fixed $\sigesq$, in particular for $\sigesq=\sigma_e^{2v}$, $\lim_{\sigssq \rightarrow \infty} \logRLss = -\infty$.
-So by choosing $\sigma_s^{2*}$ large enough we satisfy (\ref{eq:boundingbox}) for all points with $\sigssq \ge \sigma_s^{2*}$ and $\sigesq \in [\sigma_e^{2v}, \sigma_e^{2*}]$.
+Points such as $q_2$ which satisfy $\sigesq \ge \sigma_e^{2*}$ and $\sigssq \ge \widetilde{\sigma_s^2}$ are in a region where the partial derivatives (\ref{eq:drlds}, \ref{eq:drlde}) are nonpositive for $j \in \{1, \dots, s_z+2\}$ and therefore, $\logRL(\sshat) - \logRL(q_2) \ge L - \logRL(\widetilde{q_2})$.
+The latter expression is greater than $M$ iff $\logRL(\widetilde{q_2}) < L-M$.
+But examination of (\ref{eq:sum}) shows that for any fixed $\sigesq$, $\lim_{\sigssq \rightarrow \infty} \logRLss = -\infty$ and therefore $\widetilde{\sigma_s^2}$ can be chosen large enough so that \textbf{D1} is satisfied.
 
-Finally, Let $q_3^* = (0,\sigma_s^{2*})$ be the intercept of $B$ and the $\sigssq$ axis.
-Points such as $q_3$ which satisfy $\sigssq \ge \sigma_s^{2*}$ and $\sigesq \le \sigma_e^{2v}$ --- above $B$ and to the left of the $v$'th line --- are in a region where the partial derivatives (\ref{eq:drlds}, \ref{eq:drlde}) are negative for $j \in \{1, \dots, s_z\}$ but nonnegative for $j=v$.
-Let $\logRL_j(\cdot)$ be the $j$'th term of (\ref{eq:reexpress}) evaluated at $(\cdot)$.
-For $j\ge1$ we compare $\logRL_j(q_3)$ to $\logRL_j(q_3^*)$ but we compare $\logRL_v(q_3)$ to $\logRL_v(q_2^*)$.
+Let $\widetilde{q_3} = (\widetilde{\sigma_e^2}, \sigma_s^{2*})$.
+Points such as $q_3$ which satisfy $\sigesq \ge \widetilde{\sigma_e^2}$ and $\sigssq \ge \sigma_s^{2*}$ are in a region where the partial derivatives (\ref{eq:drlds}, \ref{eq:drlde}) are nonpositive for $j \in \{1, \dots, s_{z+2}\}$ and therefore, $\logRL(\sshat) - \logRL(q_3) \ge L - \logRL(\widetilde{q_3})$.
+The latter expression is greater than $M$ iff $\logRL(\widetilde{q_3}) < L-M$.
+But examination of (\ref{eq:sum}) shows that for any fixed $\sigssq$, $\lim_{\sigesq \rightarrow \infty} \logRLss = -\infty$ and therefore $\widetilde{\sigma_e^2}$ can be chosen large enough so that \textbf{D1} is satisfied.
+
+Let $\widetilde{q_4} = (\widetilde{\sigma_e^2},0)$.
+For any point $q_4$ having $\sigssq \le \sigma_s^{2*}$ and $\sigesq \ge \widetilde{q_4}$, the partial derivatives (\ref{eq:drlds}, \ref{eq:drlde}) are nonpositive for $j \in \{1, \dots, s_z+1\}$  but nonnegative for $j=s_z+2$.
+\begin{equation*}
+	\logRL(\sshat) - \logRL(q_4) \ge L - \sum_{j \ne s_z+2} \logRL_j(\widetilde{q_4}) - \logRL_{s_z+2}(\widetilde{q_3})
+\end{equation*}
+which is larger than $M$ if
 \begin{equation}
-	\logRL(\sshat) - \logRL(q_2) \ge L - \sum_{j=1}^{s_z} \logRL_j(q_3^*) - \logRL_v(q_2^*)
+\label{eq:q4}
+  \begin{split}
+    \sum_{j \ne s_z+2} \logRL_j(\widetilde{q_4})
+      &< L - M - \logRL_{s_z+2}(\widetilde{q_3})\\
+      &= L - M + \frac{1}{2} \left[ c_{s_z+2}\log\sigma_s^{2*} + \frac{d_{s_z+2}}{\sigma_s^{2*}}\right].
+  \end{split}
 \end{equation}
-which is larger than $M$ if $\sum_{1}^{s_z} \logRL_j(q_3^*) < L - M - \logRL_v(q_2^*) = L - M + n_e(\log(\sigma_e^{2v}) + 1)/2$.
+Examination of (\ref{eq:sum}) shows that for any fixed $\sigssq$, in particular for $\sigssq=0$, and for $j \ne s_z+2$,\\ $\lim_{\sigesq \rightarrow \infty} \logRL_j(\sigesq,\sigssq) = -\infty$.
+Thus $\widetilde{\sigma_e^2}$ can be chosen large enough so that the summation on the l.h.s.~of \eqref{eq:q4} is less than the r.h.s.~of \eqref{eq:q4}, and  \textbf{D1} is satisfied.
 
-Combining the requirements for points like $q_1$, $q_2$, and $q_3$, we satisfy desideratum \textbf{D1} for all points in $B^c$ by choosing $\sigma_e^*$ large enough to satisfy $\logRL(q_1^*) < L-M$ and choosing $\sigma_s^*$ large enough to satisfy $\logRL(q_2^*) < L-M$ and $\sum_{1}^{s_z} \logRL_j(q_3^*) < L - M + n_e(\log(\sigma_e^{2v}) + 1)/2$.
+Combining the requirements for points like $q_1$, $q_2$, $q_3$, and $q_4$, we satisfy desideratum \textbf{D1} for all points in $B^c$ by choosing $\widetilde{\sigma_e^2}$ and $\widetilde{\sigma_s^2}$ sufficiently large.
+%enough to satisfy $\logRL(q_1^*) < L-M$ and choosing $\sigma_s^*$ large enough to satisfy $\logRL(q_2^*) < L-M$ and $\sum_{1}^{s_z} \logRL_j(q_3^*) < L - M + n_e(\log(\sigma_e^{2v}) + 1)/2$.
 In addition, because $\logRL$ has no maxima outside of $B_1$, it also has no maxima outside of $B$.

### 2015-05-17 ms.R

@@ -1,3 +1,5 @@
+library(ggplot2)
+
 # dinvgamma: density function of the inverse-gamma distribution
 dinvgamma = function(x, shape = 1, rate = 1, scale = 1/rate, log = FALSE) {
     # return( rate^shape / gamma(shape) * x^(-shape-1) * exp(-rate/x) )
@@ -25,7 +27,7 @@ rinvgamma = function(n, shape = 1, rate = 1, scale = 1/rate) {
     return( 1 / rgamma(n,shape,rate,scale) )
 }
 
-
+# base graphics
 baseplot <- function(...){
   plot ( x=c(0,1.3), y=c(0,1.3), type="n", bty="l",
          xlab = expression(sigma[e]^2), ylab = expression(sigma[s]^2),
@@ -78,12 +80,40 @@ boxb <- function(){
   text ( x=.325, y=.125, expression(italic(b)), cex=1.5)
 }
 
+# ggplot2 version
+df <- data.frame ( x=c(0,1.3), y=c(0,1.3) )
+p <- ggplot ( df, aes(x=x,y=y) ) +
+  scale_x_continuous ( limits = c(0,1.3),
+                       name = expression(sigma[e]^2),
+                       breaks = NULL
+                     ) +
+  scale_y_continuous ( limits = c(0,1.3),
+                       name = expression(sigma[s]^2),
+                       breaks = NULL
+                     ) +
+  theme_bw()
+x1 <- c(0,.4); x1ann <- 0.23; y1 <- c(.7,0); y1ann <- 0.3
+x2 <- c(0,.1); x2ann <- 0.1; y2 <- c(1,0); y2ann <- 0.1
+x3 <- c(0,1); x3ann <- 0.5; y3 <- c(.6,0); y3ann <- 0.3
+xv <- c(.6,.6); xvann <- 0.61; yv <- c(0,1); yvann <- 0.4
+xh <- c(0,1); xhann <- 0.3; yh <- c(.6,.6); yhann <- 0.62
+B1 <- data.frame ( x=c(0,1,1,0,0), y=c(0,0,1,1,0) )
+B1annx <- c (.2,.97); B1anny <- c(.97,.2)
+B1small <- data.frame ( x=c(1,1,.6,.6,0,0,2/47,.1,1),
[…]

### 2015-05-19 ms.tex
edits to Sec 2.1, 2.2

@@
 \end{equation}
-Equations~\eqref{eq:reexpress} and~\eqref{eq:logpost} can both be written as a sum of multiples of logs and inverses of linear combinations $\abss$, as in \eqref{eq:sum}, where the summands with $j=s_z+1$ and $j=s_z+2$ are for the terms involving only $\sigesq$ and $\sigssq$, respectively.
+Equations~\eqref{eq:reexpress} and~\eqref{eq:logpost} can both be written as a sum of multiples of logs and inverses of linear combinations $\abss$, as in \eqref{eq:sum}, where the summands with $j=s_z+1$ and $j=s_z+2$ are for the terms involving only $\sigesq$ and $\sigssq$, respectively and where we have dropped the irrelevant constant $B$.
 \begin{equation}
 \label{eq:sum}
-  -\frac{1}{2} \sum_{j=1}^{s_z+2}\left[ c_j \log(\abss) + \frac{d_j}{\abss}\right]
+  \logRLss = -\frac{1}{2} \sum_{j=1}^{s_z+2}\left[ c_j \log(\abss) + \frac{d_j}{\abss}\right]
 \end{equation}
@@
 Our derivation proceeds from \eqref{eq:sum}.
-We use $f$ to denote the target function generically, either~\eqref{eq:reexpress} or~\eqref{eq:logpost}.
-When the target function is~\eqref{eq:reexpress}, $c_{s_z+2} = d_{s_z+2} = 0$ and the upper limit of the sum in \eqref{eq:sum} is $s_z+1$.
+We use $f$ to denote the target function generically, either $\RLorig(\sigesq,\sigssq)$ or $\pi(\sigesq,\sigssq\g y)$.
+When the target function is $\RLorig(\sigesq,\sigssq)$, $c_{s_z+2} = d_{s_z+2} = 0$ and the upper limit of the sum in \eqref{eq:sum} is effectively $s_z+1$.
 
@@
 \subsection{Partial Derivatives Determine Lines}
-The partial derivatives of $\logRLss$ can be calculated from (\ref{eq:logpost}):
+The partial derivatives of $\logRLss$ can be calculated from (\ref{eq:sum}):
 \begin{subequations}
@@
 For $j=s_z+1$, $a_j=0$, so $\eqref{eq:drlds} = 0$ and \eqref{eq:partials} determines a vertical line at $\sigesq = \sigma_e^{2*} \equiv d_{s_z+1} / c_{s_z+1}$.
-For $j=s_z+2$,  $b_j=0$, so $\eqref{eq:drlde} = 0$ and, if the target function is \eqref{eq:logpost}, \eqref{eq:partials} determines a horizontal line at $\sigssq = \sigma_s^{2*} \equiv d_{s_z+2} / c_{s_z+2}$.
+For $j=s_z+2$,  $b_j=0$, so $\eqref{eq:drlde} = 0$ and, if the target function is \eqref{eq:logpost}, \eqref{eq:partials} determines a horizontal line at $\sigssq = \sigma_s^{2*} \equiv d_{s_z+2} / c_{s_z+2}$, while if the target function is \eqref{eq:reexpress}, $c_{s_z+2} = d_{s_z+2} = 0$ and the term for $j=s_z+2$ effectively vanishes.
 
@@
 The largest intercept of those lines on the $\sigssq$ axis and the largest intercept on the $\sigesq$ axis determine a rectangle $B_1$ whose outline is marked in bold.
-At any point above $B_1$, e.g., $p_1$, the partial derivatives \eqref{eq:partials} are either negative or, for $j=s_z+1$, zero.
+At any point above $B_1$, e.g., $p_1$, the partial derivatives \eqref{eq:partials} are either negative or, for $j=s_z+1$, the partial derivative \eqref{eq:drlds} is zero.
 Therefore, $\logRL(\widetilde{p_1}) \ge \logRL(p_1)$.
@@
 \begin{equation}
-	\logRL(\sshat) - \logRL(q_1) \ge L - \sum_{j \ne s_z+1} \logRL_j(\widetilde{q_1}) - \logRL_{s_z+1}(\widetilde{q_2})
+	\logRL(\sshat) - \logRL(q_1) \ge L - \sum_{j \ne s_z+1} \logRL_j(\widetilde{q_1}) - \logRL_{s_z+1}(\widetilde{q_2}),
 \end{equation}
@@
 For any point $q_4$ having $\sigssq \le \sigma_s^{2*}$ and $\sigesq \ge \widetilde{q_4}$, the partial derivatives (\ref{eq:drlds}, \ref{eq:drlde}) are nonpositive for $j \in \{1, \dots, s_z+1\}$  but nonnegative for $j=s_z+2$.
+Therefore,
 \begin{equation*}
@@
 
-Combining the requirements for points like $q_1$, $q_2$, $q_3$, and $q_4$, we satisfy desideratum \textbf{D1} for all points in $B^c$ by choosing $\widetilde{\sigma_e^2}$ and $\widetilde{\sigma_s^2}$ sufficiently large.
+Combining the requirements for points like $q_1$, $q_2$, $q_3$, and $q_4$, desideratum \textbf{D1} is satisfied for all points in $B^c$ by choosing $\widetilde{\sigma_e^2}$ and $\widetilde{\sigma_s^2}$ sufficiently large.
 %enough to satisfy $\logRL(q_1^*) < L-M$ and choosing $\sigma_s^*$ large enough to satisfy $\logRL(q_2^*) < L-M$ and $\sum_{1}^{s_z} \logRL_j(q_3^*) < L - M + n_e(\log(\sigma_e^{2v}) + 1)/2$.
@@
 Next we consider the relationship between the $\logRL_j$'s and boxes as depicted in Figure~\ref{fig:boxb}, which shows a box $b$ and three lines.
-(Though we haven't drawn $v$ or $h$ lines, the argument for them is similar to the argument for the lines we have drawn.)  For lines such as $j=2$ that lie below $b$ the partial derivatives of $\logRL_j$ are negative, so the maximum and minimum of $\logRL_j$ within $b$ are attained at the lower left and upper right corners respectively.
-The situation is reversed for lines like $j=3$ that lie above $b$: the partial derivatives are positive so the maximum and minimum are attained at the upper right and lower left corners respectively.
-For lines like $j=1$ that pass through $b$ the minimum is attained at either the upper right or lower left corner while the maximum is attained on the line.
+For lines such as $j=2$ that lie below or to the left of $b$, the partial derivatives (\ref{eq:drlds}, \ref{eq:drlde}) are nonpositive, so the maximum and minimum of $\logRL_j$ within $b$ are attained at the lower left and upper right corners respectively.
+The situation is reversed for lines like $j=3$ that lie above or to the right of $b$: the partial derivatives are nonnegative so the maximum and minimum are attained at the upper right and lower left corners respectively.
+For lines like $j=1$ that pass through $b$, the minimum is attained at either the upper right or lower left corner while the maximum is attained on the line.
 
@@
 \item With $y$, $X$, $Z$, $\Sigma_e$, and $\Sigma_s$ --- see \eqref{eq:lmm}
-	and \eqref{eq:rll} for definitions --- compute $\{a_j, b_j, c_j, d_j\}$ for $j=1,
-	\dots, s_z, v, h$.
+	and \eqref{eq:rll} for definitions and the Appendix for details --- compute $\{a_j, b_j, c_j, d_j\}$ for $j=1,
+	\dots, s_z+2$.
 \item Create two lists of boxes: active and inactive.  Both lists are initially empty.
@@
 For a $\logRLssorig$ analysis there are 43 lines, as shown in Figure~\ref{fig:hmolines}.
-Running the algorithm on the box $B_1$ determined by the lines' maximum intercepts on the $\sigesq$ and $\sigssq$ with the settings
+Running the algorithm on the box $B_1$ determined by the lines' maximum intercepts on the $\sigesq$ and $\sigssq$ axes with the settings
 \begin{equation*}
@@
 	\includegraphics[width=.45\linewidth]{figs/hmo_HH11_lines.pdf}
-	\caption{The 43 lines, $j=1, 2, \dots, 42, v$, for the $\logRLorig$ analysis of the HMO data.
-	              The $v$ line is dashed.}
+	\caption{The 43 lines, $j=1, 2, \dots, 43$, for the $\logRLorig$ analysis of the HMO data.
+	              The line for $j = s_z+1 = 43$ is dashed.}
 	\label{fig:hmolines}
@@
 The red dot is the lower left corner of the box that maximizes $L^b$.
-Boxes with $U^b \ge \max L^b$ are outlined in blue; $(\sshat)$ must lie within the blue region.
+Boxes with $U^b \ge L \equiv \max L^b$ are outlined in blue; $(\sshat)$ must lie within the blue region.
 Figure~\ref{fig:hmorll}  suggests that we can restrict attention to a box determined by $\sigesq\in(300,1000)$ and $\sigssq\in(0,1000)$.
@@
 For a Bayesian analysis there are 44 lines, as shown in Figure~\ref{fig:hmoBayeslines}.
-Figure~\ref{fig:hmoBayeslines} differs from Figure~\ref{fig:hmolines} in that it includes a horizontal line and the position of the vertical line is slightly shifted.
+Figure~\ref{fig:hmoBayeslines} differs from Figure~\ref{fig:hmolines} in that it includes a horizontal line for $j = s_z+2$ and the position of the vertical line for $j = s_z+1$ is slightly shifted.
 Because there is only a slight change from the $\logRLorig$ analysis, we began by running the algorithm with the refined control parameters from the $\logRLorig$ analysis.
@@
 	\includegraphics[width=.45\linewidth]{figs/hmolines_HH11_Bayes.pdf}
-	\caption{The 44 lines, $j=1, 2, \dots, 44, v, h$, for the Bayesian analysis of the HMO data.
-	              The $v$ and $h$ lines are dashed.}
+	\caption{The 44 lines, $j=1, 2, \dots, 44$, for the Bayesian analysis of the HMO data.
+	              Lines 43 and 44 are dashed.}
 	\label{fig:hmoBayeslines}
@@
 In fact, those boxes all have lower boundaries $\sigssq = 0.04673004$, upper boundaries $\sigssq = 0.04768372$, $L^b \in (-1270.00, -1266.09)$, $U^b \in (-1252.877, -1248.524)$, and $U^b-L^b \in (17.12299, 18.01643)$, whereas the red dot shows where $\max L^b = -1261.705$ occurs.
-The ridge in the posterior density near $\sigssq = 0.04$ is introduced by the $\IG(1.1,0.1)$ prior, which has a mean of 1, an infinite variance, and a peak at 0.04761905.
+The blue boxes near $\sigssq = 0.04$ are introduced by the $\IG(1.1,0.1)$ prior, which has a mean of 1, an infinite variance, and a peak at 0.04761905.
 The usual analyses of the posterior density are carried out either by MCMC --- in which case one hopes that one's Markov chain happens to sample all regions of high density --- or by evaluating the posterior on a grid --- in which case one hopes one's grid is fine enough to capture all regions of high density.
@@
 Because we fit a quadratic spline, the entries in $Z$ are squares.
-$\Sigma_e$ and $\Sigma_s$ are identity matrices of the appropriate dimension.
-See \cite{hodges:2013} for details.
+$\Sigma_e$ and $\Sigma_s$ are identity matrices of the appropriate dimension; see \cite{hodges:2013} for details.
 Following Hodges, we center and scale the \textcompute{year} column of X, then compute the $\textcompute{year}^2$ column of $X$ and all the columns of Z from the transformed \textcompute{year}, so Z becomes
@@
 
-The column space of $Z$ shares no dimensions with the column space of $X$ so $s_z=30$ and, for our $\logRLorig$ analysis, there are 31 lines in all, for $j=1,\dots,31, v$, as shown in Figure~\ref{fig:gmst-lines}.
+The column space of $Z$ shares no dimensions with the column space of $X$ so $s_z=30$ and, for our $\logRLorig$ analysis, there are 31 lines in all, as shown in Figure~\ref{fig:gmst-lines}.
 \begin{figure}
@@
 	\includegraphics[width=.5\linewidth]{figs/gmst-lines.pdf}
-	\caption{The 31 lines, $j=1, \dots, 30, v$, for the $\logRLorig$ analysis of
+	\caption{The 31 lines for the $\logRLorig$ analysis of
 	              global mean surface temperatures.}
@@
 A natural question to ask is \emph{What about linear mixed models with more than two variances?}  A partial answer is given by \cite{hodges:2013} who shows that some models with more than two variances can be reexpressed similarly to \eqref{eq:reexpress} but others can't.
-More complex models that can be reexpressed this way include but are probably not limited to models displaying general balance that are also orthogonal designs \citep[all balanced ANOVAs plus other models;][]{houtman_speed:1983}, models that are separable in a specific sense \citep[Section 17.1.5]{hodges:2013}, and miscellaneous other models \citep[Section 17.1.5]{hodges:2013}, e.g., a spatial model including random effects for heterogeneity and spatial clustering (an improper conditional autoregressive effect).
+More complex models that can be reexpressed this way include, but are probably not limited to, models displaying general balance that are also orthogonal designs \citep[all balanced ANOVAs plus other models;][]{houtman_speed:1983}, models that are separable in a specific sense \citep[Section 17.1.5]{hodges:2013}, and miscellaneous other models \citep[Section 17.1.5]{hodges:2013}, e.g., a spatial model including random effects for heterogeneity and spatial clustering (an improper conditional autoregressive effect).
 We have not explored whether the reexpressible models can be analyzed by our algorithm; that's one direction for future work.
@@
 \section{Appendix}
-\paragraph{Derivation of $\{a_j\}$ and $\{\hat v_j\}$ in \eqref{eq:reexpress}}  Our derivation follows \cite{hodges:2013},
+\paragraph{Derivation of $\{a_j\}$ and $\{\hat v_j\}$ in \eqref{eq:reexpress}, \eqref{eq:logpost}, and \eqref{eq:sum}}  Our derivation follows \cite{hodges:2013},
 which contains more details.  There are three steps.

### 2015-05-19 ms.R
figs in ggplot.  minor edits to ms

@@ -99,8 +99,11 @@ xv <- c(.6,.6); xvann <- 0.61; yv <- c(0,1); yvann <- 0.4
 xh <- c(0,1); xhann <- 0.3; yh <- c(.6,.6); yhann <- 0.62
 B1 <- data.frame ( x=c(0,1,1,0,0), y=c(0,0,1,1,0) )
 B1annx <- c (.2,.97); B1anny <- c(.97,.2)
-B1small <- data.frame ( x=c(1,1,.6,.6,0,0,2/47,.1,1),
-                        y=c(0,.6,.6,1,1,.6,27/47,0,0)
+# B1small <- data.frame ( x=c(1,.6,.6,0,0,2/47,.1,1),
+#                         y=c(0,.24,1,1,.6,27/47,0,0)
+#                       )
+B1small <- data.frame ( x=c(1,.6,.6,.4/7,4/110,0,0,.4/9.4,.1,1),
+                        y=c(0,.24,.6,.6,7/11,1,.6,5.4/9.4,0,0)
                       )
 B <- data.frame ( x=c(0,1.2,1.2,0,0), y=c(1.2,1.2,0,0,1.2) )
 Bannx <- c (.2,1.17); Banny <- c(1.17,.2)

### 2015-05-20 ms.tex
incidence of multimodality

@@
 
-It is known (e.g.\ \citealt{henn&hodges:2014}) that $\logRLss$\  
-can have multiple maxima and that existing general purpose algorithms for linear mixed models may fail to find all of them, as shown by examples in \cite{hodges:2013}, \cite{henn&hodges:2014}, and elsewhere.
-\cite{mullen:2014} examines 18 optimization functions available in \R; tests them on 48 objective functions, admittedly more complicated than $\logRLss$; and finds that even the best of them fail in over 10\% of the cases.
+%It is known (e.g.\ \citealt{henn&hodges:2014}) that $\logRLss$\  
+%can have multiple maxima and that existing general purpose algorithms for linear mixed models may fail to find all of them, as shown by examples in \cite{hodges:2013}, \cite{henn&hodges:2014}, and elsewhere.
+%\cite{mullen:2014} examines 18 optimization functions available in \R; tests them on 48 objective functions, admittedly more complicated than $\logRLss$; and finds that even the best of them fail in over 10\% of the cases.
 \cite{henn&hodges:2014} examine conditions under which multiple maxima occur in posterior densities and conclude ``\dots second maxima in posterior distributions therefore may be more common than reports in the literature would suggest."  Thus, failure to find local and global maxima may be common.
+It is known (e.g.\ \citealt{henn&hodges:2014}) that $\logRLss$
+can have multiple maxima,
+though the incidence of multiple maxima is unknown.
+Multi-modal posterior distributions arise readily from conflict between the likelihood and prior;  \cite{Liu&Hodges:2003} explores an important such conflict in detail and \cite{wakefield:1998} gives a naturally-occurring example explored further in \cite{henn&hodges:2014}.
+Multiple maxima in restricted likelihoods have received far less attention and any statement about their incidence would be speculation.
+To our knowledge, two naturally-occurring cases have have been reported, in \cite{welham_thompson:2009} and \cite{reiss_etal:2014};  \cite{henn&hodges:2014} report an artificial case and give a recipe for manufacturing examples.
+
+As for numerical optimizers, it is known that existing general purpose algorithms for linear mixed models may fail to find all of the local maxima of $\logRLss$, as shown by examples in \cite{hodges:2013}, \cite{henn&hodges:2014}, and elsewhere.
+\cite{mullen:2014} examines 18 optimization functions available in \R, tests them on 48 objective functions (admittedly more complicated than $\logRLss$) and finds that even the best of them fail in over 10\% of the cases.
+\cite{henn&hodges:2014} examine conditions under which multiple maxima occur in posterior densities and conclude ``\dots second maxima in posterior distributions therefore may be more common than reports in the literature would suggest."  Thus, failure to find local and global maxima may be common, though with available tools it is extremely laborious to determine whether multiple maxima are present and where they are.
 

### 2015-05-20 ms.tex
add 'no horizontal line'

@@
 For $j=s_z+1$, $a_j=0$, so $\eqref{eq:drlds} = 0$ and \eqref{eq:partials} determines a vertical line at $\sigesq = \sigma_e^{2*} \equiv d_{s_z+1} / c_{s_z+1}$.
-For $j=s_z+2$,  $b_j=0$, so $\eqref{eq:drlde} = 0$ and, if the target function is \eqref{eq:logpost}, \eqref{eq:partials} determines a horizontal line at $\sigssq = \sigma_s^{2*} \equiv d_{s_z+2} / c_{s_z+2}$, while if the target function is \eqref{eq:reexpress}, $c_{s_z+2} = d_{s_z+2} = 0$ and the term for $j=s_z+2$ effectively vanishes.
+For $j=s_z+2$,  $b_j=0$, so $\eqref{eq:drlde} = 0$ and, if the target function is \eqref{eq:logpost}, \eqref{eq:partials} determines a horizontal line at $\sigssq = \sigma_s^{2*} \equiv d_{s_z+2} / c_{s_z+2}$, while if the target function is \eqref{eq:reexpress}, $c_{s_z+2} = d_{s_z+2} = 0$; the term for $j=s_z+2$ effectively vanishes; and there is no horizontal line.
 

### 2015-05-20 ms.tex
minor edits

@@
 
-\item[D2] For any box $b$ we can quickly compute lower and upper bounds $(L^b,U^b)$ satisfying
+\item[D2] For any other box $b$ with sides parallel to the axes we can quickly compute lower and upper bounds $(L^b,U^b)$ satisfying
 \begin{equation*}
@@
 
-To find $B$, choose an arbitrary constant $M>0$ and an arbitrary point inside $B_1$, say $p$, as illustrated in Figure~(\ref{fig:boundingbox2}) and define $L \equiv \logRL(p)$.
+To find $B$, choose an arbitrary constant $M>0$ and an arbitrary point inside $B_1$, say $p$, as illustrated in Figure~(\ref{fig:boundingbox2}), and define $L \equiv \logRL(p)$.
 We will find a box $B\supset B_1$, as shown in Figure~(\ref{fig:boundingbox2}), outside of which $\logRLss \le L-M$, thus satisfying \textbf{D1}.

### 2015-05-20 ms.tex
beginning of Section 3

@@
 \usepackage{natbib}
-\usepackage{enumitem}
+\usepackage[inline]{enumitem}
 \usepackage{subdepth}
@@
 
-\item[D2] For any other box $b$ with sides parallel to the axes we can quickly compute lower and upper bounds $(L^b,U^b)$ satisfying
+\item[D2] For any box $b$ with sides parallel to the axes we can quickly compute lower and upper bounds $(L^b,U^b)$ satisfying
 \begin{equation*}
@@
 \end{description}
+\textbf{D1} and \textbf{D2} will allow us to divide the $(\sigesq,\sigssq)$ plane into two parts: one where $\logRL$ is low and another where we know $\logRL$ to within a pre-specified $\epsilon$.
 The next section shows how the partial derivatives are used to satisfy \textbf{D1} and \textbf{D2}.
@@
 \subsection{Lines Determine Bounds within Boxes}
+\label{sec:boundsinboxes}
 \begin{figure}
@@
 \section{An Algorithm}
+Section~\ref{sec:boundingbox} shows how to determine a box $B_1$ containing all local maxima of $\logRLss$; and it shows that we can find a bigger box $B\supseteq B_1$ satisfying \textbf{D1}.
+We are about to present an algorithm that uses Section~\ref{sec:boundsinboxes} to show that any bounded box $B^0$ can be partitioned into finitely many smaller boxes $B^0_1, \dots B^0_p$  such that, for each $B^0_i$ in the partition we know, for pre-specified constants $M, \epsilon >0$, either
+\begin{enumerate*}[label=(\alph*)]
+\item $\sup_{B^0_i} \logRL < \sup_{B_0} \logRL -M$ or
+\item $\sup_{B^0_i} \logRL - \inf_{B^0_i} \logRL < \epsilon$.
+\end{enumerate*}
+If $B^0$ satisfies \textbf{D1} and $B^0\supseteq B_1$ then we have accomplished our goal of dividing the plane into one region where we know $\logRL$ to be low and another where we know $\logRL$ to within $\epsilon$.
+One strategy would be to find $B$ as outlined in Section~\ref{sec:boundingbox} and set $B^0 \equiv B$.
+However, we have found that $B_1$ satisfies \textbf{D1} in all the examples we have tried and we therefore follow a different strategy.
+We set $B^0 \equiv B_1$ and run the following algorithm, during which we learn whether $B_1$ satisfies \textbf{D1}.
+If it does, we're done.
+If it doesn't, then we would expand $B_1$ by a factor of 2 in both dimensions; set $B^0$ to be the expanded box; rerun the algorithm; and continue to expand and rerun until we find a $B^0$ that does satisfy \textbf{D1}.
+Section~\ref{sec:boundingbox} shows that such a bounded $B^0$ exists, and thus we are sure to find it in finite time.
+
 With \textbf{D1} and \textbf{D2} the following  algorithm will evaluate $\logRL$ to arbitrary accuracy everywhere within a box $B$.

### 2015-05-21 ms.tex
justify regions of large f, rather than mle

@@
 
-%It is known (e.g.\ \citealt{henn&hodges:2014}) that $\logRLss$\  
-%can have multiple maxima and that existing general purpose algorithms for linear mixed models may fail to find all of them, as shown by examples in \cite{hodges:2013}, \cite{henn&hodges:2014}, and elsewhere.
-%\cite{mullen:2014} examines 18 optimization functions available in \R; tests them on 48 objective functions, admittedly more complicated than $\logRLss$; and finds that even the best of them fail in over 10\% of the cases.
-\cite{henn&hodges:2014} examine conditions under which multiple maxima occur in posterior densities and conclude ``\dots second maxima in posterior distributions therefore may be more common than reports in the literature would suggest."  Thus, failure to find local and global maxima may be common.
 It is known (e.g.\ \citealt{henn&hodges:2014}) that $\logRLss$
@@
 
-Our point of view is that it is important to find regions where $f$ or $\log f$ is large relative to its maximum regardless of whether the regions contain local maxima.
-(E.g., if $f$ or $\log f$ is relatively flat and large over a region, it matters little whether the region contains small bumps that are, technically, local maxima.)  This paper shows how to find $(\sshat) \equiv \argsup_{\sigesq, \sigssq}\logRLss$ globally (typically either the maximum restricted log likelihood estimate or the maximum \textit{a posteriori} estimate) and to evaluate $\logRLss$ everywhere, each to within a prespecified tolerance (hence ``approximately exact''), without fear of missing regions of high $f$ or $\logRL$.
-The technique relies on the partial derivatives of $\logRLss$.
-Analysis of the partial derivatives allows us to satisfy two desiderata.
+Our point of view is that it is important to find regions where $f$ or $\log f$ is large relative to its maximum regardless of whether those regions contain local maxima.
+Points with large $\logRLss$ are those that describe the data, and possibly also prior information, well, at least compared to points with low $\logRLss$.
+If $f$ or $\log f$ is relatively flat and large over a region, it matters little whether the region contains small bumps that are, technically, local maxima.
+A good analysis should strive to find all  points with large $\logRL$.
+Therefore, the purpose of this paper is to introduce an algorithm that will, in finite time, divide the $(\sigesq,\sigssq)$ plane into two parts: one where we know $\logRL$ is small relative to its maximum and another where we know $\logRL$ to within a pre-specified $\epsilon$ (hence the term ``approximately exact'').
+The algorithm can also be used to find $(\sshat) \equiv \argsup_{\sigesq, \sigssq}\logRLss$  (typically either the maximum restricted log likelihood estimate or the maximum \textit{a posteriori} estimate), to within a pre-specified tolerance without fear of missing regions of high $f$ or $\logRL$.
+
+The technique relies on the partial derivatives of $\logRLss$.  Analysis of the partial derivatives allows us to satisfy two desiderata.
 \begin{description}
@@
 \end{description}
-\textbf{D1} and \textbf{D2} will allow us to divide the $(\sigesq,\sigssq)$ plane into two parts: one where $\logRL$ is low and another where we know $\logRL$ to within a pre-specified $\epsilon$.
+\textbf{D1} and \textbf{D2} allow us to divide the $(\sigesq,\sigssq)$ plane into two parts: one where $\logRL$ is low and another where we know $\logRL$ to within a pre-specified $\epsilon$.
 The next section shows how the partial derivatives are used to satisfy \textbf{D1} and \textbf{D2}.

### 2015-05-21 ms.tex
minor edit

@@
 \end{description}
-\textbf{D1} and \textbf{D2} allow us to divide the $(\sigesq,\sigssq)$ plane into two parts: one where $\logRL$ is low and another where we know $\logRL$ to within a pre-specified $\epsilon$.
+\textbf{D1} and \textbf{D2} allow us to divide the $(\sigesq,\sigssq)$ plane into two parts: one where $\logRL$ is at least $M$ below its maximum and another where we know $\logRL$ to within a pre-specified $\epsilon$.
 The next section shows how the partial derivatives are used to satisfy \textbf{D1} and \textbf{D2}.

### 2015-05-21 ms.tex
wrote pseudocode

@@
 \subsection{Partial Derivatives Determine Lines}
+\label{subsec:lines}
 The partial derivatives of $\logRLss$ can be calculated from (\ref{eq:sum}):
@@
 \begin{equation}
+\label{eq:bounds}
 	L^b \equiv \sum_j L^b_j \le \inf_{(\sigssq,\sigesq)\in b} \logRLss \le
 	\sup_{(\sigssq,\sigesq)\in b} \logRLss \le \sum_j U^b_j \equiv U^b.
-\label{eq:bounds}
 \end{equation}
@@
 \section{An Algorithm}
+\label{sec:algorithm}
 Section~\ref{sec:boundingbox} shows how to determine a box $B_1$ containing all local maxima of $\logRLss$; and it shows that we can find a bigger box $B\supseteq B_1$ satisfying \textbf{D1}.
@@
 
-With \textbf{D1} and \textbf{D2} the following  algorithm will evaluate $\logRL$ to arbitrary accuracy everywhere within a box $B$.
+With \textbf{D1} and \textbf{D2} the \textcompute{R} function \textcompute{findf}, sketched below, will evaluate, within a box $B^0$, $\logRL$ to arbitrary accuracy everywhere $\logRL$ is large.
+These notes explain some parts of the function.
 \begin{enumerate}
-\item Specify constants \maxit, $M$, $\epsilon$, $\delta_e$, and $\delta_s$.
+\item A dataframe \textcompute{lines} is an input to \textcompute{findf}.  \textcompute{lines} contains the slopes
+  and the intercepts on both axes of the $s_z+2$ lines discussed in Section~\ref{subsec:lines}.
+  Details for computing them from $y$, $X$, $Z$, $\Sigma_e$, and $\Sigma_s$ are in the Appendix.
+\item \textcompute{startbox}, or $B^0$, is an input to \textcompute{findf}.  As explained at the beginning of
+  Section~\ref{sec:algorithm}, \textcompute{startbox} could be $B_1$, $B$, or any other box the user chooses.
+\item Constants \maxit, $M$, $\epsilon$, $\delta_e$, and $\delta_s$ are inputs to \textcompute{findf}.
 	\begin{enumerate}[label=(\alph*)]
 	\item \maxit\ is the maximum number (could be $\infty$) of iterations of
-		the algorithm's loop.
-	\item We will not carefully analyze regions of the plane where
+		\textcompute{findf}'s loop.
+	\item We will not further analyze regions of the plane where
 		$\logRLss < \logRL(\sshat) - M$.  ($M$ could be $\infty$.)
-	\item We will evaluate $\logRL$ to within an accuracy of $\epsilon$ (could be 0) inside $B$
+	\item We will evaluate $\logRL$ to within an accuracy of $\epsilon$ (could be 0) inside $B^0$
 	         unless evaluation is stopped by one of the other criteria.
@@
 	\end{enumerate}
-	\maxit, $M$, $\epsilon$, $\delta_e$, and $\delta_s$ are constants that
-	control the running of the algorithm.  Their meaning may become more
-	clear as we explain the algorithm.
-\item With $y$, $X$, $Z$, $\Sigma_e$, and $\Sigma_s$ --- see \eqref{eq:lmm}
-	and \eqref{eq:rll} for definitions and the Appendix for details --- compute $\{a_j, b_j, c_j, d_j\}$ for $j=1,
-	\dots, s_z+2$.
-\item Create two lists of boxes: active and inactive.  Both lists are initially empty.
-\item Specify a box $B$ as the sole entry of the active list.  $B$ could be,
-	but doesn't have to be, a box satisfying desideratum \textbf{D1}.
-\item Either set $L = -\infty$ or evaluate $\logRL(p_i)$ at a few points
-	$p_1, \dots p_k$ and set $L = \max_i\logRL(p_i)$.  $L$ is our current
-	lower bound on $\logRL(\sshat)$.
-\item While the active list is not empty or we have not reached \maxit\
-	iterations
-	\begin{enumerate}
-	\item For each active box $b$:
-		\begin{enumerate}
-		\item if $U^b < L-M$ move $b$ to the inactive list.
-		\item if $U^b - L^b < \epsilon$ move $b$ to the inactive list
-		\item if the horizontal extent of $b$ is less than $\delta_e$,
-			move $b$ to the inactive list
-		\item if the vertical extent of $b$ is less than $\delta_s$,
-			move $b$ to the inactive list
-		\item otherwise, $b$ is still active.  Divide $b$ into four subboxes.
-			Remove $b$ and add the subboxes to the active list.
+\item A \textcompute{box} is a list consisting of
+	 \begin{itemize}
+	 \item upper and lower limits on $\sigesq$,
+	 \item upper and lower limits on $\sigssq$,
+	 \item upper and lower bounds on $\logRL$, $U^b$ and $L^b$ from \eqref{eq:bounds},
+	 \item indicators for whether this box lies above (or to the right of), below (or to the left of),
+	   or straddles each of the $s_z+2$ lines.
+	 \end{itemize}
+\item \textcompute{killfunc} is a function, shown below \textcompute{findf}, that determines whether
+	a box $b$ should be divided more finely.
+\item \textcompute{splitbox} is a function that takes a box $b =
+	[\sigesq{}_\text{low}, \sigesq{}_\text{high}] \times [\sigssq{}_\text{low}, \sigssq{}_\text{high}]$ as input 
+	and returns the four boxes created by dividing each side of $b$ at its midpoint.
+%\item With $y$, $X$, $Z$, $\Sigma_e$, and $\Sigma_s$ --- see \eqref{eq:lmm}
+%	and \eqref{eq:rll} for definitions and the Appendix for details --- compute $\{a_j, b_j, c_j, d_j\}$ for $j=1,
+%	\dots, s_z+2$.
+%\item Create two lists of boxes: active and inactive.  Both lists are initially empty.
+%\item Specify a box $B$ as the sole entry of the active list.  $B$ could be,
+%	but doesn't have to be, a box satisfying desideratum \textbf{D1}.
+%\item Either set $L = -\infty$ or evaluate $\logRL(p_i)$ at a few points
+%	$p_1, \dots p_k$ and set $L = \max_i\logRL(p_i)$.  $L$ is our current
+%	lower bound on $\logRL(\sshat)$.
+%\item While the active list is not empty or we have not reached \maxit\
+%	iterations
+%	\begin{enumerate}
+%	\item For each active box $b$:
+%		\begin{enumerate}
+%		\item if $U^b < L-M$ move $b$ to the inactive list.
+%		\item if $U^b - L^b < \epsilon$ move $b$ to the inactive list
+%		\item if the horizontal extent of $b$ is less than $\delta_e$,
+%			move $b$ to the inactive list
+%		\item if the vertical extent of $b$ is less than $\delta_s$,
+%			move $b$ to the inactive list
+%		\item otherwise, $b$ is still active.  Divide $b$ into four subboxes.
+%			Remove $b$ and add the subboxes to the active list.
+%		\end{enumerate}
+%	\item Set $L = \max ( L, \max_{b\in \text{active list}}L^b )$.  $L$ is our new
+%		lower bound on $\logRL(\sshat)$.
+%	\end{enumerate}
+%\item Return the active and inactive lists.
 \end{enumerate}
-	\item Set $L = \max ( L, \max_{b\in \text{active list}}L^b )$.  $L$ is our new
-		lower bound on $\logRL(\sshat)$.
-	\end{enumerate}
-\item Return the active and inactive lists.
-\end{enumerate}
+\begin{verbatim}
+findf <- function(lines, startbox, eps = 0, delE = 0, delS = 0, M = Inf, maxit = 10,
+                  ratio = FALSE, lognote = "summary") {
+
+    inactive <- list()  # a list of inactive boxes
+    ninact <- 0  # the number of inactive boxes
+
+    active <- list(startbox)  # a list of active boxes
+    nact <- 0  # the number of active boxes
+
+    lowbound <- -Inf  # lowerbound on max(log(f))
+    iter <- 0  # iteration number
+
+    while (nact > 0 && iter < maxit) {
+        # Find the lower bound of each box and the maximum of the lower bounds.  For
+        # each active box, either make it inactive or divide it.
+        low.act <- max(vapply(X = active, FUN = function(box) {
+            box$bounds[1]
+        }, FUN.VALUE = 0.1))
+        lowbound <- max(lowbound, low.act)
+        kill <- vapply(X = active, FUN = killfunc, FUN.VALUE = TRUE, lb = lowbound,
+            M = M, eps = eps, delE - delE, delS = delS, ratio = ratio)
+            # which boxes become inactive?
+        nkill <- sum(kill)
+        if (nkill > 0) {
+            inactive[(ninact + 1):(ninact + nkill)] <- active[kill] 
+            # add those boxes to the inactive list
+        }
+        ninact <- length(inactive)
+        kids <- list()  # subdivisions of active boxes
+        nkids <- 0
+        for (i in which(!kill)) {  # split active boxes into 4 parts
+            kids[(nkids + 1):(nkids + 4)] <- splitbox(active[[i]], lines)
+            nkids <- nkids + 4
+        }
+        active <- kids
+        nact <- length(active)
+
+        iter <- iter + 1
+        write(c("iteration", iter, "nact", nact, "ninact", ninact, "lowbound",
+            lowbound), file = "log.out", ncolumns = 8, append = TRUE)
+    }
+
+    tmp <- t(vapply(X = c(active, inactive), FUN = function(box) {
+        c(box$lims.sigsqs, box$lims.sigsqe, box$bounds)
+    }, FUN.VALUE = c(sigsqs.lo = 0.1, sigsqs.hi = 0.1, sigsqe.lo = 0.1, sigsqe.hi = 0.1,
+        rll.lower = 0.1, rll.upper = 0.1)))
+    end_time <- Sys.time()
+    write(paste(lognote, "runtime: ", round(end_time - start_time, digits = 3)),
+          file = "log.out", ncolumns = 1, append = TRUE)
+    return(data.frame(tmp))
+}
+
+# conditions under which a box becomes inactive
+killfunc <- function(box, lb, M, eps, delE, delS, ratio) {
+    # lb is a global (within startbox) lower bound on max(log(f));
+    #      it changes at each iteration.
+    #  M, eps, delE, delS stay constant throughout the iterations.
+    cond.low <- box$bounds[2] < lb - M
+    cond.eps <- diff(box$bounds) < eps
+    cond.E <- ifelse(ratio, diff(log(box$lims.sigsqe)) < delE,
+                            diff(box$lims.sigsqe) < delE)
+    cond.S <- ifelse(ratio, diff(log(box$lims.sigsqs)) < delS,
+                            diff(box$lims.sigsqs) < delS)
+    return(cond.low || cond.eps || cond.E || cond.S)  # stop dividing this box?
+}
+\end{verbatim}
 

### 2015-05-22 ms.tex
implication of setting maxit = infty

@@
 
+\subsection*{Alt: Lines Determine a Bounding Box}
+%\label{sec:boundingbox}
+
+Let $\sigesq{}^M$ and $\sigssq{}^M$ be the largest intercepts of the $s_z+2$ lines on the $\sigesq$ and $\sigssq$ axes, respectively.  I.e.,
+\begin{align*}
+  \sigesq{}^M &= \max \left\{ \frac{d_1}{c_1}, \cdots, \frac{d_{s_z+1}}{c_{s_z+1}} \right\}\\
+  \sigssq{}^M &= \begin{cases}
+                             \max \left\{ \frac{d_1}{a_1}, \cdots, \frac{d_{s_z}}{a_{s_z}} \right\} &
+                                  \text{if $\logRL$ is \eqref{eq:reexpress}}\\
+                             \max \left\{ \frac{d_1}{a_1}, \cdots, \frac{d_{s_z}}{a_{s_z}}, \frac{d_{s_z+2}}{c_{s_z+2}} \right\} &
+                                  \text{if $\logRL$ is \eqref{eq:logpost}}.
+                           \end{cases}
+\end{align*}
+Let $B_1$ be the box whose lower-left and upper-right corners are $(0,0)$ and $(\sigesq{}^M,\sigssq{}^M)$, respectively.
+Figure~(\ref{fig:bigboundingbox}) illustrates with five lines --- realistic data sets may have more --- labelled $j=1, 2, 3, s_z+1, s_z+2$.\\
+\textbf{Claim} \emph{All local maxima of $\logRL$ lie within $B_1$.}\\
+\textbf{Proof} Let $p_1 = (\sigesq{}_1, \sigssq{}_1)$ be a point such that $\sigssq{}_1 > \sigssq{}^M$ and let $\widetilde{p_1} = (\sigesq{}_1, \sigssq{}^M)$, as illustrated in Figure~(\ref{fig:bigboundingbox}).
+For $j=1, \dots, s_z, s_z+2$, the partial derivatives \eqref{eq:drlde} are negative; for $j=s_z+1$, the partial derivative is zero.
+Therefore, $\logRL(\widetilde{p_1}) \ge \logRL(p_1)$ and there can be no local maxima of $\logRL$ above the line $\sigssq = \sigssq{}^M$.
+Let $p_2 = (\sigesq{}_2, \sigssq{}_2)$ be a point such that $\sigesq{}_2 > \sigesq{}^M$ and let $\widetilde{p_2} = (\sigesq{}^M, \sigssq{}_2)$.
+For $j=1, \dots, s_z, s_z+1$, the partial derivatives \eqref{eq:drlds} are negative; for $j=s_z+2$, the partial derivative is zero.
+Therefore, $\logRL(\widetilde{p_2}) \ge \logRL(p_2)$ and there can be no local maxima of $\logRL$ to the right of the line $\sigesq = \sigesq{}^M$.
+\textbf{QED}\\
+
+By the claim, \mrle\ must lie on or inside $B_1$, so $B_1$ could be passed to an optimizer such as \R's \textsf{optim} or \textsf{nlminb} with potentially better results than using those functions without bounds.
+However, even with known bounds, general purpose optimizers may still miss \mrle\ and regions of high $\logRL$.
+We will next find a box $B \supset B_1$ satisfying desideratum \textbf{D1}.
+First, though, we pause to note that the region outlined in bold in Figure~(\ref{fig:smallboundingbox}) is a subset of $B_1$ that must also contain \mrle\ by the same reasoning used to prove the previous claim.
+But it is not rectangular, hence less convenient than $B_1$, so we don't pursue it further.
+
+%\begin{figure}[h]
+%  \begin{subfigure}{.5\textwidth}
+%	\centering
+%	\includegraphics[width=.8\linewidth]{figs/boundingbox.pdf}
+%	\caption{A rectangular region $B_1$ containing all maxima.}
+%	\label{fig:bigboundingbox}
+%  \end{subfigure}
+%  \begin{subfigure}{.5\textwidth}
+%	\centering
+%	\includegraphics[width=.8\linewidth]{figs/smallboundingregion.pdf}
+%	\caption{A smaller region containing all maxima.}
+%	\label{fig:smallboundingbox}
+%  \end{subfigure}
+%  \caption{all local and global maxima lie in the regions bounded by the solid
+%                dark line.
+%               }
+%  \label{fig:boundingbox}
+%\end{figure}
+
+\noindent\textbf{Claim} \emph{For any positive number $M$, there exist positive numbers $\widetilde{\sigma_e^2} > \sigesq{}^M$ and $\widetilde{\sigma_s^2} > \sigssq{}^M$ that determine a box $B \supset B_1$ whose lower-left and upper-right corners are $(0,0)$ and $(\widetilde{\sigma_e^2}, \widetilde{\sigma_s^2})$, respectively, such that
+\begin{equation*}
+  \sup_{B_1^c} \logRLss \le \logRL(\sshat) - M.
+\end{equation*}}
+I.e., \textbf{D1} is satisfied.\\
+\noindent\textbf{Proof}
+Let $p$ be an arbitrary point inside $B_1$, as illustrated in Figure~(\ref{fig:boundingbox2}), and define $L \equiv \logRL(p)$.
+Let $\widetilde{q_1} = (0,\widetilde{\sigma_s^2})$ be the intercept of $B$ and the $\sigssq$ axis and let $\widetilde{q_2} = (\sigma_e^{2*},\widetilde{\sigma_s^2})$.
+Let $q_1 = (\sigesq{}_1, \sigssq{}_1)$ be a point satisfying $\sigesq{}_1 \le \sigma_e^{2*}$ and $\sigssq{}_1 \ge \widetilde{\sigma_s^2}$ --- above $B$ and to the left of the $s_z+1$ line.
+The partial derivatives (\ref{eq:drlds}), evaluated at $q_1$ are negative for $j \in \{1, \dots, s_z, s_z+2\}$ but zero for $j=s_z+1$.
+The partial derivatives (\ref{eq:drlde}), evaluated at $q_1$ are negative for $j \in \{1, \dots, s_z, s_z+1\}$ but nonnegative for $j=s_z+2$.
+Let $\logRL_j(\cdot)$ denote the $j$'th term of \eqref{eq:sum} evaluated at $(\cdot)$.
+\begin{equation}
+	\logRL(\sshat) - \logRL(q_1) \ge L - \sum_{j \ne s_z+1} \logRL_j(\widetilde{q_1}) - \logRL_{s_z+1}(\widetilde{q_2}),
+\end{equation}
+which is larger than $M$ if
+\begin{equation}
+\label{eq:q1}
+  \begin{split}
+    \sum_{j \ne s_z+1} \logRL_j(\widetilde{q_1})
+      &< L - M - \logRL_{s_z+1}(\widetilde{q_2})\\
+      &= L - M + \frac{1}{2} \left[ c_{s_z+1}\log\sigma_e^{2*} + \frac{d_{s_z+1}}{\sigma_e^{2*}}\right].
+  \end{split}
+\end{equation}
+Examination of (\ref{eq:sum}) shows that for any fixed $\sigesq$, in particular for $\sigesq=0$, and for $j \ne s_z+1$,\\ $\lim_{\sigssq \rightarrow \infty} \logRL_j(\sigesq,\sigssq) = -\infty$.
+Thus $\widetilde{\sigma_s^2}$ can be chosen large enough so that the summation on the l.h.s.~of \eqref{eq:q1} is less than the r.h.s.~of \eqref{eq:q1}, and  \textbf{D1} is satisfied.
+
+Let $q_2 = (\sigesq{}_2, \sigssq{}_2)$ satisfy $\sigesq{}_2 \ge \sigma_e^{2*}$ and $\sigssq{}_2 \ge \widetilde{\sigma_s^2}$ --- above $B$ and to the right of the $s_z+1$ line.
+(Points $q_2$ and $q_3$ in Figure~(\ref{fig:boundingbox2}) both satisfy this criterion.)  The partial derivatives (\ref{eq:drlds}, \ref{eq:drlde}) evaluated at $q_2$ are nonpositive for $j \in \{1, \dots, s_z+2\}$; therefore, $\logRL(\sshat) - \logRL(q_2) \ge L - \logRL(\widetilde{q_2})$.
+The latter expression is greater than $M$ iff $\logRL(\widetilde{q_2}) < L-M$.
+But examination of (\ref{eq:sum}) shows that for any fixed $\sigesq$, $\lim_{\sigssq \rightarrow \infty} \logRLss = -\infty$ and therefore $\widetilde{\sigma_s^2}$ can be chosen large enough so that \textbf{D1} is satisfied.
+
+Let $\widetilde{q_3} = (\widetilde{\sigma_e^2}, \sigma_s^{2*})$ and let $q_3 = (\sigesq{}_3, \sigssq{}_3)$ satisfy $\sigesq{}_3 \ge \widetilde{\sigma_e^2}$ and $\sigssq{}_3 \ge \sigma_s^{2*}$.
+The partial derivatives (\ref{eq:drlds}, \ref{eq:drlde}) evaluated at $q_3$ are nonpositive for $j \in \{1, \dots, s_{z+2}\}$; therefore, $\logRL(\sshat) - \logRL(q_3) \ge L - \logRL(\widetilde{q_3})$.
+The latter expression is greater than $M$ iff $\logRL(\widetilde{q_3}) < L-M$.
+But examination of (\ref{eq:sum}) shows that for any fixed $\sigssq$, $\lim_{\sigesq \rightarrow \infty} \logRLss = -\infty$ and therefore $\widetilde{\sigma_e^2}$ can be chosen large enough so that \textbf{D1} is satisfied.
+
+Let $\widetilde{q_4} = (\widetilde{\sigma_e^2},0)$ and let $q_4 = (\sigesq{}_4, \sigssq{}_4)$ satisfy $\sigssq{}_4 \le \sigma_s^{2*}$ and $\sigesq{}_4 \ge \widetilde{q_4}$.
+The partial derivatives \eqref{eq:drlds} evaluated at $q_4$ are nonpositive for $j \in \{1, \dots, s_z+1\}$  but nonnegative for $j=s_z+2$.
+The partial derivatives \eqref{eq:drlde} evaluated at $q_4$ are nonpositive for $j \in \{1, \dots, s_z, s_z+2\}$  but nonnegative for $j=s_z+1$.
+Therefore,
+\begin{equation*}
+	\logRL(\sshat) - \logRL(q_4) \ge L - \sum_{j \ne s_z+2} \logRL_j(\widetilde{q_4}) - \logRL_{s_z+2}(\widetilde{q_3})
+\end{equation*}
+which is larger than $M$ if
+\begin{equation}
+\label{eq:q4}
+  \begin{split}
+    \sum_{j \ne s_z+2} \logRL_j(\widetilde{q_4})
+      &< L - M - \logRL_{s_z+2}(\widetilde{q_3})\\
+      &= L - M + \frac{1}{2} \left[ c_{s_z+2}\log\sigma_s^{2*} + \frac{d_{s_z+2}}{\sigma_s^{2*}}\right].
+  \end{split}
+\end{equation}
+Examination of (\ref{eq:sum}) shows that for any fixed $\sigssq$, in particular for $\sigssq=0$, and for $j \ne s_z+2$,\\ $\lim_{\sigesq \rightarrow \infty} \logRL_j(\sigesq,\sigssq) = -\infty$.
+Thus $\widetilde{\sigma_e^2}$ can be chosen large enough so that the summation on the l.h.s.~of \eqref{eq:q4} is less than the r.h.s.~of \eqref{eq:q4}, and  \textbf{D1} is satisfied.
+
+Combining the requirements for points like $q_1$, $q_2$, $q_3$, and $q_4$, desideratum \textbf{D1} is satisfied for all points in $B^c$ by choosing $\widetilde{\sigma_e^2}$ and $\widetilde{\sigma_s^2}$ sufficiently large.
+\textbf{QED}\\
+%enough to satisfy $\logRL(q_1^*) < L-M$ and choosing $\sigma_s^*$ large enough to satisfy $\logRL(q_2^*) < L-M$ and $\sum_{1}^{s_z} \logRL_j(q_3^*) < L - M + n_e(\log(\sigma_e^{2v}) + 1)/2$.
+In addition, because $\logRL$ has no maxima outside of $B_1$, it also has no maxima outside of $B$.
+As already mentioned, if there is an $h$ line, then $q_1$ would have to be treated like $q_2$ and $q_3$.
+
+%\begin{figure}[h]
+%	\centering
+%	\includegraphics[width=.5\linewidth]{figs/boundingbox2.pdf}
+%	\caption{Box $B$ satisfies (\ref{eq:boundingbox})}
+%	\label{fig:boundingbox2}
+%\end{figure}
+
 \subsection{Lines Determine Bounds within Boxes}
@@
 	\end{enumerate}
+  Setting $\maxit = \infty$ and $\delta_e = \delta_s = 0$ implies that $B^0$ will be partitioned so that either
+  $\logRL < \logRL(\sshat) - M$ or $\logRL$ is known to within $\epsilon$, for every $B^0_i$ in the partition.
+
 \item A \textcompute{box} is a list consisting of

### 2015-05-29 ms.tex
First paragraph of Section 3

@@
 Section~\ref{sec:boundingbox} shows how to determine a box $B_1$ containing all local maxima of $\logRLss$; and it shows that we can find a bigger box $B\supseteq B_1$ satisfying \textbf{D1}.
-We are about to present an algorithm that uses Section~\ref{sec:boundsinboxes} to show that any bounded box $B^0$ can be partitioned into finitely many smaller boxes $B^0_1, \dots B^0_p$  such that, for each $B^0_i$ in the partition we know, for pre-specified constants $M, \epsilon >0$, either
+We are about to present an algorithm that uses Section~\ref{sec:boundsinboxes} to show that any bounded box $B^0$ can be partitioned into finitely many smaller boxes $B^0_1, \dots B^0_p$  such that, for each $B^0_i$ in the partition, we know for pre-specified constants $M, \epsilon >0$, either
 \begin{enumerate*}[label=(\alph*)]
-\item $\sup_{B^0_i} \logRL < \sup_{B_0} \logRL -M$ or
+\item $\sup_{B^0_i} \logRL < \sup_{B^0} \logRL -M$ or
 \item $\sup_{B^0_i} \logRL - \inf_{B^0_i} \logRL < \epsilon$.
@@
 
-With \textbf{D1} and \textbf{D2} the \textcompute{R} function \textcompute{findf}, sketched below, will evaluate, within a box $B^0$, $\logRL$ to arbitrary accuracy everywhere $\logRL$ is large.
-These notes explain some parts of the function.
+With \textbf{D1} and \textbf{D2} the \textcompute{R} function \textcompute{findf}, sketched below, will evaluate, within a box $B^0$, $\logRL$ to arbitrary accuracy everywhere $\logRL$ is large by performing the following tasks:
+\begin{enumerate*}[label=(\alph*)]
+\item accept as input $\{a_j, b_j, c_j, d_j\}, B^0$ and some tuning constants;
+\item create a list of active boxes, initially consisting of just $B^0$;
+\item create a list of inactive boxes, initially empty;
+\item for each active box $B_j$, find $(L^b_j, U^b_j)$ and determine whether $B_j$ needs to be subdivided.
+\end{enumerate*}
+ These notes explain some parts of the function in more detail.
 \begin{enumerate}

### 2015-05-29 ms.tex
RLL analysis of HMO

@@
 \begin{equation*}
-	\maxit=10; \hspace{1cm} \epsilon=5; \hspace{1cm}
-	\delta_e=\log(10); \hspace{1cm} \delta_s=\log(10); \hspace{1cm} M=5
-\end{equation*}
-results in the the output displayed in Table~\ref{table:hmo_HH11_run}, which shows the state of the algorithm after iterations 1 through 10: the numbers of active and inactive boxes and the current value of the lower bound $L$ on $\max\log f$.
+	\maxit=15; \hspace{1cm} \epsilon=1; \hspace{1cm}
+	\delta_e=\log(10); \hspace{1cm} \delta_s=\log(10); \hspace{1cm} \text{ratio $=$ TRUE}; \hspace{1cm} M=5
+\end{equation*}
+results in the the output displayed in Table~\ref{table:hmo_HH11_run}, which shows the state of the algorithm after iterations 1 through 15: the numbers of active and inactive boxes and the current value of the lower bound $L$ on $\max\log f$.
 We see that boxes are steadily transferred from the active to the inactive list and that $L$ increases monotonically.
-After 10 iterations there is a total of 640 boxes, which are displayed in Figure~\ref{fig:hmo1}.
+After 15 iterations there is a total of 895 boxes, which are displayed in Figure~\ref{fig:hmo1}.
+
+%\begin{table}[h]
+%\centering
+%\begin{tabular}{|c|c|c|c|}
+%\hline
+%Iteration & $n$ active boxes & $n$ inactive boxes & $L$\\
+%1 & 4 & 0 & $-\infty$\\
+%2 & 8 & 2 & -1618.84\\
+%3 & 16 & 6 & -1509.24\\
+%4 & 32 & 14 & -1408.12\\
+%5 & 44 & 35 & -1325.99\\
+%6 & 56 & 65 & -1265.01\\
+%7 & 104 & 95 & -1256.35\\
+%8 & 184 & 153 & -1243.51\\
+%9 & 224 & 281 & -1240.60\\
+%10 & 180 & 460 & -1239.49\\
+%\hline
+%\end{tabular}
+%\caption{The state of the algorithm after each of 10 iterations for the HMO data.}
+%\label{table:hmo_HH11_run}
+%\end{table}
 
@@
 2 & 8 & 2 & -1618.84\\
-3 & 16 & 6 & -1509.24\\
-4 & 32 & 14 & -1408.12\\
-5 & 44 & 35 & -1325.99\\
-6 & 56 & 65 & -1265.01\\
-7 & 104 & 95 & -1256.35\\
-8 & 184 & 153 & -1243.51\\
-9 & 224 & 281 & -1240.60\\
-10 & 180 & 460 & -1239.49\\
+3 & 12 & 7 & -1509.24\\
+4 & 12 & 16 & -1408.12\\
+5 & 12 & 25 & -1325.99\\
+6 & 16 & 33 & -1265.01\\
+7 & 20 & 44 & -1256.35\\
+8 & 28 & 57 & -1243.51\\
+9 & 48 & 73 & -1240.60\\
+10 & 76 & 102 & -1239.49\\
+11 & 132 & 145 & -1239.40\\
+12 & 232 & 219 & -1239.40\\
+13 & 420 & 346 & -1239.40\\
+14 & 172 & 723 & -1239.40\\
+15 & 0 & 895 & -1239.40\\
 \hline
 \end{tabular}
-\caption{The state of the algorithm after each of 10 iterations for the HMO data.}
+\caption{The state of the algorithm after each of 15 iterations for the HMO data.}
 \label{table:hmo_HH11_run}
@@
 
-Figure~\ref{fig:hmoboxes} shows the outlines of the 640 boxes.
+Figure~\ref{fig:hmoboxes} shows the outlines of the 895 boxes.
 The algorithm did not need to divide the boxes with large $\sigesq$ or $\sigssq$ as finely as those with small $\sigesq$ and $\sigssq$ because they more readily satisfy either $U^b < L-M$ or $U^b - L^b < \epsilon$, so become inactive.
@@
 Boxes with $U^b \ge L \equiv \max L^b$ are outlined in blue; $(\sshat)$ must lie within the blue region.
-Figure~\ref{fig:hmorll}  suggests that we can restrict attention to a box determined by $\sigesq\in(300,1000)$ and $\sigssq\in(0,1000)$.
+%Figure~\ref{fig:hmorll}  suggests that we can restrict attention to a box determined by $\sigesq\in(300,1000)$ and $\sigssq\in(0,1000)$.
 So we rerun the algorithm on that box and with more stringent control parameters:
-\begin{equation*}
-	\maxit=15; \hspace{1cm} \epsilon=1; \hspace{1cm}
-	\delta_e=0; \hspace{1cm} \delta_s=0; \hspace{1cm} M=10
-\end{equation*}
-Table~\ref{table:hmo_HH11_run2} shows the output.
-The algorithm needed 12 iterations to move all boxes to the inactive list.
-The resulting 37,516 boxes are shown in Figure~\ref{fig:hmo2}.
+%\begin{equation*}
+%	\maxit=15; \hspace{1cm} \epsilon=1; \hspace{1cm}
+%	\delta_e=0; \hspace{1cm} \delta_s=0; \hspace{1cm} M=10
+%\end{equation*}
+%Table~\ref{table:hmo_HH11_run2} shows the output.  The algorithm needed 12 iterations to move all boxes to the inactive list.  The resulting 37,516 boxes are shown in Figure~\ref{fig:hmo2}.
 For comparison, the standard REML analysis using \R's \textcompute{lme} function yields $\sshat \approx (495, 99)$ with 95\% confidence intervals of $(421, 582)$ and $(39, 248)$.
 
-\begin{table}[h]
-\centering
-\begin{tabular}{|c|c|c|c|}
-\hline
-Iteration & $n$ active boxes & $n$ inactive boxes & $L$\\
-1 & 4 & 0 & $-1320.51$\\
-2 & 16 & 0 & -1276.23\\
-3 & 64 & 0 & -1252.18\\
-4 & 256 & 0 & -1243.52\\
-5 & 1024 & 0 & -1239.61\\
-6 & 4020 & 19 & -1237.78\\
-7 & 13128 & 757 & -1236.70\\
-8 & 14092 & 10362 & -1236.16\\
-9 & 9444 & 22093 & -1235.88\\
-10 & 5716 & 30108 & -1235.88\\
-11 & 2256 & 35260 & -1235.88\\
-12 & 0 & 37516 & -1235.88\\
-\hline
-\end{tabular}
-\caption{The state of the algorithm after each of 12 iterations for the HMO data.}
-\label{table:hmo_HH11_run2}
-\end{table}
-
-\begin{figure}
-  \begin{subfigure}{.5\textwidth}
-	\centering
-	\includegraphics[width=.8\linewidth]{figs/hmo_HH11_boxes2.pdf}
-	\caption{Locations of the boxes.}
-	\label{fig:hmoboxes2}
-  \end{subfigure}
-  \begin{subfigure}{.5\textwidth}
-	\centering
-	\includegraphics[width=.8\linewidth]{figs/hmo_HH11_rll2.pdf}
-	\caption{Grayscale shows $L^b$ in each box.  Red dot shows $b^* \equiv\argmax L^b$.
-	Boxes outlined in blue have $U^b \ge L^{b^*}$.  Axes are logarithmic.}
-	\label{fig:hmorll2}
-  \end{subfigure}
-  \caption{The 37,516 boxes produced in the second run of the algorithm for the
-	$\logRLorig$ analysis of the HMO data. $\maxit=15; \epsilon=1;
-	\delta_e=0; \delta_s=0; \text{and\ } M=10$.}
-  \label{fig:hmo2}
-\end{figure}
-
-Figure~\ref{fig:hmorll2} is sufficiently refined that we needn't run the algorithm further.
-Figure~\ref{fig:hmorll2} depicts the same $\logRLorig$ as \cite{henn&hodges:2014}'s Figures~2a (MCMC draws) and~2b ($\logRLorig$ contours), but their Figure~2a was produced by MCMC whereas our Figure~\ref{fig:hmorll2} was produced by direct calculation.
-Their Figure~2a shows that the MCMC sampler did not sample any values of $\sigssq$ less than about 10, whereas our Figure~\ref{fig:hmorll2} and their Figure~2b shows that there is a region of high $\logRLorig$ extending down to $\sigssq=0$.
+%\begin{table}[h]
+%\centering
+%\begin{tabular}{|c|c|c|c|}
+%\hline
+%Iteration & $n$ active boxes & $n$ inactive boxes & $L$\\
+%1 & 4 & 0 & $-1320.51$\\
+%2 & 16 & 0 & -1276.23\\
+%3 & 64 & 0 & -1252.18\\
+%4 & 256 & 0 & -1243.52\\
+%5 & 1024 & 0 & -1239.61\\
+%6 & 4020 & 19 & -1237.78\\
+%7 & 13128 & 757 & -1236.70\\
+%8 & 14092 & 10362 & -1236.16\\
+%9 & 9444 & 22093 & -1235.88\\
+%10 & 5716 & 30108 & -1235.88\\
+%11 & 2256 & 35260 & -1235.88\\
+%12 & 0 & 37516 & -1235.88\\
+%\hline
+%\end{tabular}
+%\caption{The state of the algorithm after each of 12 iterations for the HMO data.}
+%\label{table:hmo_HH11_run2}
+%\end{table}
+
+%\begin{figure}
+%  \begin{subfigure}{.5\textwidth}
+%	\centering
+%	\includegraphics[width=.8\linewidth]{figs/hmo_HH11_boxes2.pdf}
+%	\caption{Locations of the boxes.}
+%	\label{fig:hmoboxes2}
+%  \end{subfigure}
+%  \begin{subfigure}{.5\textwidth}
+%	\centering
+%	\includegraphics[width=.8\linewidth]{figs/hmo_HH11_rll2.pdf}
+%	\caption{Grayscale shows $L^b$ in each box.  Red dot shows $b^* \equiv\argmax L^b$.
+%	Boxes outlined in blue have $U^b \ge L^{b^*}$.  Axes are logarithmic.}
+%	\label{fig:hmorll2}
+%  \end{subfigure}
+%  \caption{The 37,516 boxes produced in the second run of the algorithm for the
+%	$\logRLorig$ analysis of the HMO data. $\maxit=15; \epsilon=1;
+%	\delta_e=0; \delta_s=0; \text{and\ } M=10$.}
+%  \label{fig:hmo2}
+%\end{figure}
+
+%Figure~\ref{fig:hmorll2} is sufficiently refined that we needn't run the algorithm further.
+Figure~\ref{fig:hmorll} depicts the same $\logRLorig$ as \cite{henn&hodges:2014}'s Figures~2a (MCMC draws) and~2b ($\logRLorig$ contours), but their Figure~2a was produced by MCMC whereas our Figure~\ref{fig:hmorll} was produced by direct calculation.
+Their Figure~2a shows that the MCMC sampler did not sample any values of $\sigssq$ less than about 10, whereas our Figure~\ref{fig:hmorll} and their Figure~2b shows that there is a region of high $\logRLorig$ extending down to $\sigssq=0$.
 In fact, $\logRLorig(500,0) \approx -1241.5$, only about 6 log units below $\logRLorig(\sshat) \approx -1235.5$.
@@
 
-The HMO $\logRLorig$ analysis illustrates a typical workflow: start with a crudely determined box and coarse settings of the control parameters, then refine the box and the settings as suggested by the output of the first run.
+%The HMO $\logRLorig$ analysis illustrates a typical workflow: start with a crudely determined box and coarse settings of the control parameters, then refine the box and the settings as suggested by the output of the first run.
 Repeat as needed.

### 2015-05-29 ms.tex
minor edit
guidance on M, epsilon

@@
 
-\subsection{Lines Determine a Bounding Box}
-\label{sec:boundingbox}
-
-Figure~(\ref{fig:bigboundingbox}) shows five lines --- realistic data sets may have more --- labelled $j=1, 2, 3, s_z+1, s_z+2$.
+%\subsection{Lines Determine a Bounding Box}
+%\label{sec:boundingbox}
+%
+%Figure~(\ref{fig:bigboundingbox}) shows five lines --- realistic data sets may have more --- labelled $j=1, 2, 3, s_z+1, s_z+2$.
 The largest intercept of those lines on the $\sigssq$ axis and the largest intercept on the $\sigesq$ axis determine a rectangle $B_1$ whose outline is marked in bold.
@@
 But it is not rectangular, hence less convenient than $B_1$, so we don't pursue it further.
-
-\begin{figure}[h]
-  \begin{subfigure}{.5\textwidth}
-	\centering
-	\includegraphics[width=.8\linewidth]{figs/boundingbox.pdf}
-	\caption{A rectangular region $B_1$ containing all maxima.}
-	\label{fig:bigboundingbox}
-  \end{subfigure}
-  \begin{subfigure}{.5\textwidth}
-	\centering
-	\includegraphics[width=.8\linewidth]{figs/smallboundingregion.pdf}
-	\caption{A smaller region containing all maxima.}
-	\label{fig:smallboundingbox}
-  \end{subfigure}
-  \caption{all local and global maxima lie in the regions bounded by the solid
-                dark line.
-               }
-  \label{fig:boundingbox}
-\end{figure}
-
-To find $B$, choose an arbitrary constant $M>0$ and an arbitrary point inside $B_1$, say $p$, as illustrated in Figure~(\ref{fig:boundingbox2}), and define $L \equiv \logRL(p)$.
+%
+%\begin{figure}[h]
+%  \begin{subfigure}{.5\textwidth}
+%	\centering
+%	\includegraphics[width=.8\linewidth]{figs/boundingbox.pdf}
+%	\caption{A rectangular region $B_1$ containing all maxima.}
+%	\label{fig:bigboundingbox}
+%  \end{subfigure}
+%  \begin{subfigure}{.5\textwidth}
+%	\centering
+%	\includegraphics[width=.8\linewidth]{figs/smallboundingregion.pdf}
+%	\caption{A smaller region containing all maxima.}
+%	\label{fig:smallboundingbox}
+%  \end{subfigure}
+%  \caption{all local and global maxima lie in the regions bounded by the solid
+%                dark line.
+%               }
+%  \label{fig:boundingbox}
+%\end{figure}
+%
+%To find $B$, choose an arbitrary constant $M>0$ and an arbitrary point inside $B_1$, say $p$, as illustrated in Figure~(\ref{fig:boundingbox2}), and define $L \equiv \logRL(p)$.
 We will find a box $B\supset B_1$, as shown in Figure~(\ref{fig:boundingbox2}), outside of which $\logRLss \le L-M$, thus satisfying \textbf{D1}.
-
-$B$ is determined by its intercepts $\widetilde{\sigma_e^2}$ and $\widetilde{\sigma_s^2}$ on the $\sigesq$ and $\sigssq$ axes respectively.
+%
+%$B$ is determined by its intercepts $\widetilde{\sigma_e^2}$ and $\widetilde{\sigma_s^2}$ on the $\sigesq$ and $\sigssq$ axes respectively.
 Let $\widetilde{q_1} = (0,\widetilde{\sigma_s^2})$ be the intercept of $B$ and the $\sigssq$ axis and let $\widetilde{q_2} = (\sigma_e^{2*},\widetilde{\sigma_s^2})$.
@@
 Let $\logRL_j(\cdot)$ denote the $j$'th term of \eqref{eq:sum} evaluated at $(\cdot)$.
-\begin{equation}
-	\logRL(\sshat) - \logRL(q_1) \ge L - \sum_{j \ne s_z+1} \logRL_j(\widetilde{q_1}) - \logRL_{s_z+1}(\widetilde{q_2}),
-\end{equation}
-which is larger than $M$ if
-\begin{equation}
-\label{eq:q1}
-  \begin{split}
-    \sum_{j \ne s_z+1} \logRL_j(\widetilde{q_1})
-      &< L - M - \logRL_{s_z+1}(\widetilde{q_2})\\
-      &= L - M + \frac{1}{2} \left[ c_{s_z+1}\log\sigma_e^{2*} + \frac{d_{s_z+1}}{\sigma_e^{2*}}\right].
-  \end{split}
-\end{equation}
-Examination of (\ref{eq:sum}) shows that for any fixed $\sigesq$, in particular for $\sigesq=0$, and for $j \ne s_z+1$,\\ $\lim_{\sigssq \rightarrow \infty} \logRL_j(\sigesq,\sigssq) = -\infty$.
+%\begin{equation}
+%	\logRL(\sshat) - \logRL(q_1) \ge L - \sum_{j \ne s_z+1} \logRL_j(\widetilde{q_1}) - \logRL_{s_z+1}(\widetilde{q_2}),
+%\end{equation}
+%which is larger than $M$ if
+%\begin{equation}
+%\label{eq:q1}
+%  \begin{split}
+%    \sum_{j \ne s_z+1} \logRL_j(\widetilde{q_1})
+%      &< L - M - \logRL_{s_z+1}(\widetilde{q_2})\\
+%      &= L - M + \frac{1}{2} \left[ c_{s_z+1}\log\sigma_e^{2*} + \frac{d_{s_z+1}}{\sigma_e^{2*}}\right].
+%  \end{split}
+%\end{equation}
+%Examination of (\ref{eq:sum}) shows that for any fixed $\sigesq$, in particular for $\sigesq=0$, and for $j \ne s_z+1$,\\ $\lim_{\sigssq \rightarrow \infty} \logRL_j(\sigesq,\sigssq) = -\infty$.
 Thus $\widetilde{\sigma_s^2}$ can be chosen large enough so that the summation on the l.h.s.~of \eqref{eq:q1} is less than the r.h.s.~of \eqref{eq:q1}, and  \textbf{D1} is satisfied.
-
-Points such as $q_2$ which satisfy $\sigesq \ge \sigma_e^{2*}$ and $\sigssq \ge \widetilde{\sigma_s^2}$ are in a region where the partial derivatives (\ref{eq:drlds}, \ref{eq:drlde}) are nonpositive for $j \in \{1, \dots, s_z+2\}$ and therefore, $\logRL(\sshat) - \logRL(q_2) \ge L - \logRL(\widetilde{q_2})$.
+%
+%Points such as $q_2$ which satisfy $\sigesq \ge \sigma_e^{2*}$ and $\sigssq \ge \widetilde{\sigma_s^2}$ are in a region where the partial derivatives (\ref{eq:drlds}, \ref{eq:drlde}) are nonpositive for $j \in \{1, \dots, s_z+2\}$ and therefore, $\logRL(\sshat) - \logRL(q_2) \ge L - \logRL(\widetilde{q_2})$.
 The latter expression is greater than $M$ iff $\logRL(\widetilde{q_2}) < L-M$.
 But examination of (\ref{eq:sum}) shows that for any fixed $\sigesq$, $\lim_{\sigssq \rightarrow \infty} \logRLss = -\infty$ and therefore $\widetilde{\sigma_s^2}$ can be chosen large enough so that \textbf{D1} is satisfied.
-
-Let $\widetilde{q_3} = (\widetilde{\sigma_e^2}, \sigma_s^{2*})$.
+%
+%Let $\widetilde{q_3} = (\widetilde{\sigma_e^2}, \sigma_s^{2*})$.
 Points such as $q_3$ which satisfy $\sigesq \ge \widetilde{\sigma_e^2}$ and $\sigssq \ge \sigma_s^{2*}$ are in a region where the partial derivatives (\ref{eq:drlds}, \ref{eq:drlde}) are nonpositive for $j \in \{1, \dots, s_{z+2}\}$ and therefore, $\logRL(\sshat) - \logRL(q_3) \ge L - \logRL(\widetilde{q_3})$.
@@
 But examination of (\ref{eq:sum}) shows that for any fixed $\sigssq$, $\lim_{\sigesq \rightarrow \infty} \logRLss = -\infty$ and therefore $\widetilde{\sigma_e^2}$ can be chosen large enough so that \textbf{D1} is satisfied.
-
-Let $\widetilde{q_4} = (\widetilde{\sigma_e^2},0)$.
+%
+%Let $\widetilde{q_4} = (\widetilde{\sigma_e^2},0)$.
 For any point $q_4$ having $\sigssq \le \sigma_s^{2*}$ and $\sigesq \ge \widetilde{q_4}$, the partial derivatives (\ref{eq:drlds}, \ref{eq:drlde}) are nonpositive for $j \in \{1, \dots, s_z+1\}$  but nonnegative for $j=s_z+2$.
 Therefore,
-\begin{equation*}
-	\logRL(\sshat) - \logRL(q_4) \ge L - \sum_{j \ne s_z+2} \logRL_j(\widetilde{q_4}) - \logRL_{s_z+2}(\widetilde{q_3})
-\end{equation*}
-which is larger than $M$ if
-\begin{equation}
-\label{eq:q4}
-  \begin{split}
-    \sum_{j \ne s_z+2} \logRL_j(\widetilde{q_4})
-      &< L - M - \logRL_{s_z+2}(\widetilde{q_3})\\
-      &= L - M + \frac{1}{2} \left[ c_{s_z+2}\log\sigma_s^{2*} + \frac{d_{s_z+2}}{\sigma_s^{2*}}\right].
-  \end{split}
-\end{equation}
-Examination of (\ref{eq:sum}) shows that for any fixed $\sigssq$, in particular for $\sigssq=0$, and for $j \ne s_z+2$,\\ $\lim_{\sigesq \rightarrow \infty} \logRL_j(\sigesq,\sigssq) = -\infty$.
+%\begin{equation*}
+%	\logRL(\sshat) - \logRL(q_4) \ge L - \sum_{j \ne s_z+2} \logRL_j(\widetilde{q_4}) - \logRL_{s_z+2}(\widetilde{q_3})
+%\end{equation*}
+%which is larger than $M$ if
+%\begin{equation}
+%\label{eq:q4}
+%  \begin{split}
+%    \sum_{j \ne s_z+2} \logRL_j(\widetilde{q_4})
+%      &< L - M - \logRL_{s_z+2}(\widetilde{q_3})\\
+%      &= L - M + \frac{1}{2} \left[ c_{s_z+2}\log\sigma_s^{2*} + \frac{d_{s_z+2}}{\sigma_s^{2*}}\right].
+%  \end{split}
+%\end{equation}
+%Examination of (\ref{eq:sum}) shows that for any fixed $\sigssq$, in particular for $\sigssq=0$, and for $j \ne s_z+2$,\\ $\lim_{\sigesq \rightarrow \infty} \logRL_j(\sigesq,\sigssq) = -\infty$.
 Thus $\widetilde{\sigma_e^2}$ can be chosen large enough so that the summation on the l.h.s.~of \eqref{eq:q4} is less than the r.h.s.~of \eqref{eq:q4}, and  \textbf{D1} is satisfied.
-
-Combining the requirements for points like $q_1$, $q_2$, $q_3$, and $q_4$, desideratum \textbf{D1} is satisfied for all points in $B^c$ by choosing $\widetilde{\sigma_e^2}$ and $\widetilde{\sigma_s^2}$ sufficiently large.
-%enough to satisfy $\logRL(q_1^*) < L-M$ and choosing $\sigma_s^*$ large enough to satisfy $\logRL(q_2^*) < L-M$ and $\sum_{1}^{s_z} \logRL_j(q_3^*) < L - M + n_e(\log(\sigma_e^{2v}) + 1)/2$.
+%
+%Combining the requirements for points like $q_1$, $q_2$, $q_3$, and $q_4$, desideratum \textbf{D1} is satisfied for all points in $B^c$ by choosing $\widetilde{\sigma_e^2}$ and $\widetilde{\sigma_s^2}$ sufficiently large.
+%%enough to satisfy $\logRL(q_1^*) < L-M$ and choosing $\sigma_s^*$ large enough to satisfy $\logRL(q_2^*) < L-M$ and $\sum_{1}^{s_z} \logRL_j(q_3^*) < L - M + n_e(\log(\sigma_e^{2v}) + 1)/2$.
 In addition, because $\logRL$ has no maxima outside of $B_1$, it also has no maxima outside of $B$.
@@
 \begin{figure}[h]
+	\label{fig:boundingbox2}
 	\centering
@@
 	\caption{Box $B$ satisfies (\ref{eq:boundingbox})}
-	\label{fig:boundingbox2}
-\end{figure}
-
-\subsection*{Alt: Lines Determine a Bounding Box}
-%\label{sec:boundingbox}
+\end{figure}
+
+\subsection{Lines Determine a Bounding Box}
+\label{sec:boundingbox}
 
@@
 
-%\begin{figure}[h]
-%  \begin{subfigure}{.5\textwidth}
-%	\centering
-%	\includegraphics[width=.8\linewidth]{figs/boundingbox.pdf}
-%	\caption{A rectangular region $B_1$ containing all maxima.}
-%	\label{fig:bigboundingbox}
-%  \end{subfigure}
-%  \begin{subfigure}{.5\textwidth}
-%	\centering
-%	\includegraphics[width=.8\linewidth]{figs/smallboundingregion.pdf}
-%	\caption{A smaller region containing all maxima.}
-%	\label{fig:smallboundingbox}
-%  \end{subfigure}
-%  \caption{all local and global maxima lie in the regions bounded by the solid
-%                dark line.
-%               }
-%  \label{fig:boundingbox}
-%\end{figure}
+\begin{figure}[h]
+  \begin{subfigure}{.5\textwidth}
+	\centering
+	\includegraphics[width=.8\linewidth]{figs/boundingbox.pdf}
+	\caption{A rectangular region $B_1$ containing all maxima.}
+	\label{fig:bigboundingbox}
+  \end{subfigure}
+  \begin{subfigure}{.5\textwidth}
+	\centering
+	\includegraphics[width=.8\linewidth]{figs/smallboundingregion.pdf}
+	\caption{A smaller region containing all maxima.}
+	\label{fig:smallboundingbox}
+  \end{subfigure}
+  \caption{all local and global maxima lie in the regions bounded by the solid
+                dark line.
+               }
+  \label{fig:boundingbox}
+\end{figure}
 
@@
 \noindent\textbf{Proof}
-Let $p$ be an arbitrary point inside $B_1$, as illustrated in Figure~(\ref{fig:boundingbox2}), and define $L \equiv \logRL(p)$.
+Let $p$ be an arbitrary point inside $B_1$, as illustrated in Figure~(\ref{fig:boundingbox2}), and define $L \equiv \logRL(p)$; we use $L$ as a lower bound on $\logRL(\sshat)$.
+The proof considers several regions outside $B_1$ in turn, as exemplified by the points $q_1$, $q_2$, $q_3$, and $q_4$ in Figure~\ref{fig:boundingbox2}.
 Let $\widetilde{q_1} = (0,\widetilde{\sigma_s^2})$ be the intercept of $B$ and the $\sigssq$ axis and let $\widetilde{q_2} = (\sigma_e^{2*},\widetilde{\sigma_s^2})$.
@@
 
-%\begin{figure}[h]
-%	\centering
-%	\includegraphics[width=.5\linewidth]{figs/boundingbox2.pdf}
-%	\caption{Box $B$ satisfies (\ref{eq:boundingbox})}
-%	\label{fig:boundingbox2}
-%\end{figure}
+\begin{figure}[h]
+	\centering
+	\includegraphics[width=.5\linewidth]{figs/boundingbox2.pdf}
+	\caption{Box $B$ satisfies (\ref{eq:boundingbox})}
+	\label{fig:boundingbox2}
+\end{figure}
 
@@
 \begin{enumerate}
-\item A dataframe \textcompute{lines} is an input to \textcompute{findf}.  \textcompute{lines} contains the slopes
-  and the intercepts on both axes of the $s_z+2$ lines discussed in Section~\ref{subsec:lines}.
+\item A dataframe \textcompute{lines} is an input to \textcompute{findf}.  \textcompute{lines} contains the
+  $\{a_j, b_j, c_j, d_j \}$ from Section~\ref{subsec:lines}.
   Details for computing them from $y$, $X$, $Z$, $\Sigma_e$, and $\Sigma_s$ are in the Appendix.
@@
 
+Before giving examples, we mention some considerations in setting the tuning constants $M$ and $\epsilon$, which have to do with the interpretation of likelihood ratios.
+We imagine a reference experiment in which a coin is chosen, either fair or two-headed, and tossed.
+If the coin is tossed twice and lands Heads both times, it produces a likelihood ratio of 4 in favor of the two-headed coin.
+That's only weak evidence, so we don't see the need to resolve likelihood functions much beyond a factor of four, or loglikelihood functions much beyond a factor of $\log(4) \approx 1.4$.
+In practice, we use $\epsilon = 1$ as a convenient default.
+If the coin were tossed 10 times and yielded 10 Heads, the likelihood ratio would be $2^{10} = 1024$.
+That's strong evidence, so we don't see the need to resolve $\logRL$ where it is less than .001 times its maximum.
+$\log(1000) \approx 7$, so we use $M = 7$ as a convenient default.
+
 \section{Examples}

### 2015-05-30 ms.tex
HMO Bayes

@@
 \begin{equation*}
-	\maxit=15; \hspace{1cm} \epsilon=1; \hspace{1cm}
-	\delta_e=\log(10); \hspace{1cm} \delta_s=\log(10); \hspace{1cm} \text{ratio $=$ TRUE}; \hspace{1cm} M=5
+	\maxit=\infty; \hspace{1cm} \epsilon=1; \hspace{1cm}
+	\delta_e=0; \hspace{1cm} \delta_s=0; \hspace{1cm} M=7
 \end{equation*}
@@
 Figure~\ref{fig:hmoBayeslines} differs from Figure~\ref{fig:hmolines} in that it includes a horizontal line for $j = s_z+2$ and the position of the vertical line for $j = s_z+1$ is slightly shifted.
-Because there is only a slight change from the $\logRLorig$ analysis, we began by running the algorithm with the refined control parameters from the $\logRLorig$ analysis.
-After 15 iterations there were still active boxes, so we ran the algorithm again with the same control parameters except $\maxit=20$.
-Table~\ref{table:hmo_HH11Bayes} shows the output.
-After 20 iterations there are over 3,000,000 boxes, over 2,000,000 of which are still active.
 \begin{figure}
@@
 \end{figure}
-
+Because the $\logRLorig$ analysis in Figure~\ref{fig:hmorll} shows low values for $\sigesq, \sigssq > 1000$, we run the Bayesian analysis on the box $B^0 = (0,1000) \times (0,1000)$, again with the control parameters
+\begin{equation*}
+	\maxit=\infty; \hspace{1cm} \epsilon=1; \hspace{1cm}
+	\delta_e=0; \hspace{1cm} \delta_s=0; \hspace{1cm} M=7.
+\end{equation*}
+The algorithm finished in 22 iterations.  Table~\ref{table:hmo_HH11Bayes} shows the output.
 \begin{table}[H]
@@
 Iteration & $n$ active boxes & $n$ inactive boxes & $L$\\
-1 & 4 & 0 & $-1350.29$\\
-2 & 16 & 0 & -1303.20\\
-3 & 64 & 0 & -1279.93\\
-4 & 256 & 0 & -1270.26\\
-5 & 1024 & 0 & -1265.89\\
-6 & 4060 & 9 & -1263.71\\
-7 & 13740 & 634 & -1262.60\\
-8 & 14904 & 10648 & -1262.01\\
-9 & 10180 & 23007 & -1261.71\\
-10 & 7364 & 31346 & -1261.71\\
-11 & 6012 & 37207 & -1261.71\\
-12 & 8192 & 41171 & -1261.71\\
-13 & 16384 & 45267 & -1261.71\\
-14 & 32768 & 53459 & -1261.71\\
-15 & 65536 & 69843 & -1261.71\\
-16 & 131072 & 102611 & -1261.71\\
-17 & 262144 & 168147 & -1261.71\\
-18 & 524288 & 299219 & -1261.71\\
-19 & 1048576 & 561363 & -1261.71\\
-20 & 2097152 & 1085651 & -1261.71\\
+1 & 4 & 0 & $-\infty$\\
+2 & 16 & 0 & -1313.42\\
+3 & 52 & 3 & -1285.05\\
+4 & 132 & 22 & -1270.76\\
+5 & 264 & 88 & -1263.98\\
+6 & 628 & 195 & -1260.47\\
+7 & 1780 & 378 & -1258.71\\
+8 & 4188 & 1111 & -1257.70\\
+9 & 5676 & 3880 & -1256.88\\
+10 & 5136 & 8272 & -1255.56\\
+11 & 5528 & 12026 & -1254.20\\
+12 & 8236 & 15495 & -1252.84\\
+13 & 12488 & 20609 & -1251.50\\
+14 & 23852 & 27134 & -1250.26\\
+15 & 22740 & 45301 & -1249.22\\
+16 & 43228 & 57234 & -1248.77\\
+17 & 160208 & 60410 & -1248.77\\
+18 & 331900 & 137643 & -1248.77\\
+19 & 575196 & 325744 & -1248.77\\
+20 & 942860 & 665225 & -1248.77\\
+21 & 722412 & 1427482 & -1248.77\\
+22 & 0 & 2149894 & -1248.77\\
 \hline
 \end{tabular}
-\caption{The state of the algorithm after each of 20 iterations for the Bayesian analysis of the HMO data.}
+\caption{The state of the algorithm after each of 22 iterations for the Bayesian analysis of the HMO data.}
 \label{table:hmo_HH11Bayes}
 \end{table}
-
-Figure~\ref{fig:hmoBayes} shows the boxes and $\log \pi(\sigesq,\sigssq)$.
 Figure~\ref{fig:hmoBayes} is similar to Figure~\ref{fig:hmo2} except for the addition of the outlined blue boxes near $\sigssq = 0.04$.
@@
   \end{subfigure}
-  \caption{The 3,182,803 boxes in the Bayesian analysis of the HMO data.
-               $\maxit=20; \epsilon=1; \delta_e=0; \delta_s=0; \text{and\ } M=10$.}
+  \caption{The 2,149,894 boxes in the Bayesian analysis of the HMO data.}
   \label{fig:hmoBayes}

### 2015-05-31 ms.tex
HMO rll

@@
 We see that boxes are steadily transferred from the active to the inactive list and that $L$ increases monotonically.
-After 15 iterations there is a total of 895 boxes, which are displayed in Figure~\ref{fig:hmo1}.
-
-%\begin{table}[h]
-%\centering
-%\begin{tabular}{|c|c|c|c|}
-%\hline
-%Iteration & $n$ active boxes & $n$ inactive boxes & $L$\\
-%1 & 4 & 0 & $-\infty$\\
-%2 & 8 & 2 & -1618.84\\
-%3 & 16 & 6 & -1509.24\\
-%4 & 32 & 14 & -1408.12\\
-%5 & 44 & 35 & -1325.99\\
-%6 & 56 & 65 & -1265.01\\
-%7 & 104 & 95 & -1256.35\\
-%8 & 184 & 153 & -1243.51\\
-%9 & 224 & 281 & -1240.60\\
-%10 & 180 & 460 & -1239.49\\
-%\hline
-%\end{tabular}
-%\caption{The state of the algorithm after each of 10 iterations for the HMO data.}
-%\label{table:hmo_HH11_run}
-%\end{table}
+After 15 iterations there is a total of 9490 boxes, which are displayed in Figure~\ref{fig:hmo1}.
 
@@
 1 & 4 & 0 & $-\infty$\\
-2 & 8 & 2 & -1618.84\\
-3 & 12 & 7 & -1509.24\\
-4 & 12 & 16 & -1408.12\\
-5 & 12 & 25 & -1325.99\\
-6 & 16 & 33 & -1265.01\\
-7 & 20 & 44 & -1256.35\\
-8 & 28 & 57 & -1243.51\\
-9 & 48 & 73 & -1240.60\\
-10 & 76 & 102 & -1239.49\\
-11 & 132 & 145 & -1239.40\\
-12 & 232 & 219 & -1239.40\\
-13 & 420 & 346 & -1239.40\\
-14 & 172 & 723 & -1239.40\\
-15 & 0 & 895 & -1239.40\\
+2 & 16 & 0 & -1618.84\\
+3 & 40 & 6 & -1509.24\\
+4 & 72 & 28 & -1408.12\\
+5 & 144 & 64 & -1325.99\\
+6 & 68 & 191 & -1265.01\\
+7 & 116 & 230 & -1256.35\\
+8 & 144 & 310 & -1243.51\\
+9 & 340 & 369 & -1240.05\\
+10 & 920 & 479 & -1237.87\\
+11 & 2540 & 764 & -1236.79\\
+12 & 4380 & 2209 & -1236.20\\
+13 & 3524 & 5708 & -1235.90\\
+14 & 344 & 9146 & -1235.90\\
+15 & 0 & 9490 & -1235.90\\
 \hline
@@
   \end{subfigure}
-  \caption{The 640 boxes produced in the first run  for the
-	$\logRLorig$ analysis of the HMO data. $\maxit=10; \epsilon=5;
-	\delta_e=\log(10); \delta_s=\log(10); \text{and\ } M=5$.}
+  \caption{The 9490 boxes produced in the
+	$\logRLorig$ analysis of the HMO data. $\maxit=\infty; \epsilon=1;
+	\delta_e= \delta_s= 0; \text{and\ } M=7$.}
   \label{fig:hmo1}
@@
 
-Figure~\ref{fig:hmoboxes} shows the outlines of the 895 boxes.
+Figure~\ref{fig:hmoboxes} shows the outlines of the 9490 boxes.
 The algorithm did not need to divide the boxes with large $\sigesq$ or $\sigssq$ as finely as those with small $\sigesq$ and $\sigssq$ because they more readily satisfy either $U^b < L-M$ or $U^b - L^b < \epsilon$, so become inactive.
@@
 
-%\begin{table}[h]
-%\centering
-%\begin{tabular}{|c|c|c|c|}
-%\hline
-%Iteration & $n$ active boxes & $n$ inactive boxes & $L$\\
-%1 & 4 & 0 & $-1320.51$\\
-%2 & 16 & 0 & -1276.23\\
-%3 & 64 & 0 & -1252.18\\
-%4 & 256 & 0 & -1243.52\\
-%5 & 1024 & 0 & -1239.61\\
-%6 & 4020 & 19 & -1237.78\\
-%7 & 13128 & 757 & -1236.70\\
-%8 & 14092 & 10362 & -1236.16\\
-%9 & 9444 & 22093 & -1235.88\\
-%10 & 5716 & 30108 & -1235.88\\
-%11 & 2256 & 35260 & -1235.88\\
-%12 & 0 & 37516 & -1235.88\\
-%\hline
-%\end{tabular}
-%\caption{The state of the algorithm after each of 12 iterations for the HMO data.}
-%\label{table:hmo_HH11_run2}
-%\end{table}
-
-%\begin{figure}
-%  \begin{subfigure}{.5\textwidth}
-%	\centering
-%	\includegraphics[width=.8\linewidth]{figs/hmo_HH11_boxes2.pdf}
-%	\caption{Locations of the boxes.}
-%	\label{fig:hmoboxes2}
-%  \end{subfigure}
-%  \begin{subfigure}{.5\textwidth}
-%	\centering
-%	\includegraphics[width=.8\linewidth]{figs/hmo_HH11_rll2.pdf}
-%	\caption{Grayscale shows $L^b$ in each box.  Red dot shows $b^* \equiv\argmax L^b$.
-%	Boxes outlined in blue have $U^b \ge L^{b^*}$.  Axes are logarithmic.}
-%	\label{fig:hmorll2}
-%  \end{subfigure}
-%  \caption{The 37,516 boxes produced in the second run of the algorithm for the
-%	$\logRLorig$ analysis of the HMO data. $\maxit=15; \epsilon=1;
-%	\delta_e=0; \delta_s=0; \text{and\ } M=10$.}
-%  \label{fig:hmo2}
-%\end{figure}
-
-%Figure~\ref{fig:hmorll2} is sufficiently refined that we needn't run the algorithm further.
-Figure~\ref{fig:hmorll} depicts the same $\logRLorig$ as \cite{henn&hodges:2014}'s Figures~2a (MCMC draws) and~2b ($\logRLorig$ contours), but their Figure~2a was produced by MCMC whereas our Figure~\ref{fig:hmorll} was produced by direct calculation.
-Their Figure~2a shows that the MCMC sampler did not sample any values of $\sigssq$ less than about 10, whereas our Figure~\ref{fig:hmorll} and their Figure~2b shows that there is a region of high $\logRLorig$ extending down to $\sigssq=0$.
-In fact, $\logRLorig(500,0) \approx -1241.5$, only about 6 log units below $\logRLorig(\sshat) \approx -1235.5$.
+Figure~\ref{fig:hmorll} depicts the same $\logRLorig$ as \cite{henn&hodges:2014}'s Figures~2a (MCMC draws) and~2b ($\logRLorig$ contours), but their Figure~2a was produced by MCMC whereas our Figure~\ref{fig:hmorll} was produced by direct calculation in under 10 seconds.
+Their Figure~2a shows that the MCMC sampler did not sample any values of $\sigssq$ less than about 10, whereas our Figure~\ref{fig:hmorll} and their Figure~2b shows a region of high $\logRLorig$ extending down to $\sigssq=0$.
+In fact, $\logRLorig(500,0) \approx -1241.5$, only about 6 log units below $\logRLorig(\sshat) \approx -1235.7$.
 Further, about their Figure~2a, Henn and Hodges say, ``No change in contour shape indicative of a local maximum could be found in the \dots region of $(500, 600) \times (10^{-3}, 1)$, regardless of contour resolution."  I.e., they cannot be sure there are no undiscovered points with large $\logRLorig$.
@@
 \end{table}
-Figure~\ref{fig:hmoBayes} is similar to Figure~\ref{fig:hmo2} except for the addition of the outlined blue boxes near $\sigssq = 0.04$.
-In fact, those boxes all have lower boundaries $\sigssq = 0.04673004$, upper boundaries $\sigssq = 0.04768372$, $L^b \in (-1270.00, -1266.09)$, $U^b \in (-1252.877, -1248.524)$, and $U^b-L^b \in (17.12299, 18.01643)$, whereas the red dot shows where $\max L^b = -1261.705$ occurs.
-The blue boxes near $\sigssq = 0.04$ are introduced by the $\IG(1.1,0.1)$ prior, which has a mean of 1, an infinite variance, and a peak at 0.04761905.
-The usual analyses of the posterior density are carried out either by MCMC --- in which case one hopes that one's Markov chain happens to sample all regions of high density --- or by evaluating the posterior on a grid --- in which case one hopes one's grid is fine enough to capture all regions of high density.
+The Bayesian analysis in Figure~\ref{fig:hmoBayes} can be compared to the $\logRLorig$ analysis in Figure~\ref{fig:hmo1}.
+The peak of the log posterior is near $(600, .05)$, very far from $(\sshat)$.
+The posterior peak is due to the $\IG(1.1,0.1)$ prior for $\sigssq$, which has a mean of 1, an infinite variance, and a peak at 0.04761905.
+The region near $(\sshat)$, though having a lower posterior density, is a nearly flat plateau covering a large area, and has significant posterior mass, as shown by the MCMC draws in \cite{henn&hodges:2014} Figure~2c.
+Usual analyses of the posterior density are carried out either by MCMC, in which case one hopes that one's Markov chain happens to sample all regions of high enough posterior density to matter, or by evaluating the posterior on a grid --- in which case one hopes one's grid is fine enough.
 With our analysis there is no need to hope; the algorithm is certain to find all regions of high density.

### 2015-05-31 ms.tex
minor edits

@@
 \end{figure}
-Because the $\logRLorig$ analysis in Figure~\ref{fig:hmorll} shows low values for $\sigesq, \sigssq > 1000$, we run the Bayesian analysis on the box $B^0 = (0,1000) \times (0,1000)$, again with the control parameters
+Because the $\logRLorig$ analysis in Figure~\ref{fig:hmorll} shows low values for $\sigesq, \sigssq > 1000$, we run the Bayesian analysis on the box $B^0 = (0,1000) \times (0,1000)$, again with control parameters
 \begin{equation*}
@@
 The Bayesian analysis in Figure~\ref{fig:hmoBayes} can be compared to the $\logRLorig$ analysis in Figure~\ref{fig:hmo1}.
-The peak of the log posterior is near $(600, .05)$, very far from $(\sshat)$.
-The posterior peak is due to the $\IG(1.1,0.1)$ prior for $\sigssq$, which has a mean of 1, an infinite variance, and a peak at 0.04761905.
-The region near $(\sshat)$, though having a lower posterior density, is a nearly flat plateau covering a large area, and has significant posterior mass, as shown by the MCMC draws in \cite{henn&hodges:2014} Figure~2c.
-Usual analyses of the posterior density are carried out either by MCMC, in which case one hopes that one's Markov chain happens to sample all regions of high enough posterior density to matter, or by evaluating the posterior on a grid --- in which case one hopes one's grid is fine enough.
+The peak of the log posterior is near $(600, .05)$, very far from the peak of the $\logRLorig$ around $(500,100)$.
+The posterior peak is due to the $\IG(1.1,0.1)$ prior for $\sigssq$, which has a mean of 1, an infinite variance, and a peak at 0.048.
+The region near $(500,100)$, though having a lower posterior density, is part of a nearly flat plateau covering a large area.
+The MCMC draws depicted in \cite{henn&hodges:2014} Figure~2c show that the plateau has significant posterior mass.
+%Usual analyses of the posterior density are carried out either by MCMC, in which case one hopes that one's Markov chain happens to sample all regions of high enough posterior density to matter, or by evaluating the posterior on a grid --- in which case one hopes one's grid is fine enough.
 With our analysis there is no need to hope; the algorithm is certain to find all regions of high density.

### 2015-05-31 ms.tex
Added run times for code

@@
 results in the the output displayed in Table~\ref{table:hmo_HH11_run}, which shows the state of the algorithm after iterations 1 through 15: the numbers of active and inactive boxes and the current value of the lower bound $L$ on $\max\log f$.
+The run finished in less than 10 seconds.
 We see that boxes are steadily transferred from the active to the inactive list and that $L$ increases monotonically.
@@
 
-Figure~\ref{fig:hmorll} depicts the same $\logRLorig$ as \cite{henn&hodges:2014}'s Figures~2a (MCMC draws) and~2b ($\logRLorig$ contours), but their Figure~2a was produced by MCMC whereas our Figure~\ref{fig:hmorll} was produced by direct calculation in under 10 seconds.
+Figure~\ref{fig:hmorll} depicts the same $\logRLorig$ as \cite{henn&hodges:2014}'s Figures~2a (MCMC draws) and~2b ($\logRLorig$ contours), but their Figure~2a was produced by MCMC whereas our Figure~\ref{fig:hmorll} was produced by direct calculation.
 Their Figure~2a shows that the MCMC sampler did not sample any values of $\sigssq$ less than about 10, whereas our Figure~\ref{fig:hmorll} and their Figure~2b shows a region of high $\logRLorig$ extending down to $\sigssq=0$.
@@
 \end{equation*}
-The algorithm finished in 22 iterations.  Table~\ref{table:hmo_HH11Bayes} shows the output.
+The algorithm finished in 22 iterations and took a little under 3 hours.  Table~\ref{table:hmo_HH11Bayes} shows the output.
 \begin{table}[H]

### 2015-05-31 ms.tex
Discussion

@@
 \section{Discussion}
-This paper has explained and illustrated an algorithm, for linear mixed models with two variances, to find all regions where either the restricted likelihood function or the joint posterior density of the variances is high, and to evaluate the function there to arbitrary accuracy.
+This paper has explained and illustrated an algorithm for linear mixed models with two variances, to find all regions where either the restricted likelihood function or the joint posterior density of the variances is high, and to evaluate the function there to arbitrary accuracy.
 A natural question to ask is \emph{What about linear mixed models with more than two variances?}  A partial answer is given by \cite{hodges:2013} who shows that some models with more than two variances can be reexpressed similarly to \eqref{eq:reexpress} but others can't.
@@
 We don't elaborate here in order to concentrate on the main ideas.
+
+More generally, our algorithm differs from typical optimization algorithms in that it has a different goal: learning $\logRL$ to specified accuracy wherever $\logRL$ is high.
+The algorithm works by exploiting a re-expression of $\logRL$ as a sum of simpler, easily analyzed functions.
+But there may be many other statistically interesting functions that can be so re-expressed.
+For example, many likelihood functions are products of terms from conditionally independent parts of the data.
+Posterior densities have the same terms, plus a term from the prior.
+We have not yet explored whether our algorithm, and more generally the idea of learning $\logRL$ to specified accuracy, is useful outside the family of linear mixed models; that's another direction for future work.
 

### 2015-05-31 ms.tex
gmst and minor edits

@@
 \end{figure}
-Our first run of the algorithm used the box determined by the largest intercepts of the 31 lines on the $\sigesq$ and $\sigssq$ axes and the control constants 
-\begin{equation*}
-	\maxit=15; \hspace{1cm} \epsilon=10; \hspace{1cm}
-	\delta_e=1; \hspace{1cm} \delta_s=1; \hspace{1cm} M=10.
-\end{equation*}
-After 15 iterations there were 3740 boxes still in the active list and 52,973 in the inactive list; they are displayed in Figure~\ref{fig:gmst1}.
+As usual, we use $B^0 = B_1$, the box determined by the largest intercepts of the 31 lines on the $\sigesq$ and $\sigssq$ axes and the control constants 
+\begin{equation*}
+	\maxit=\infty; \hspace{1cm} \epsilon=1; \hspace{1cm}
+	\delta_e= \delta_s=0; \hspace{1cm} M=7.
+\end{equation*}
+After 25 iterations and about 35 minutes of computing time, all boxes became inactive.  Output is in Table~\ref{table:gmst_output}; boxes are displayed in Figure~\ref{fig:gmst1}.
+\begin{table}[H]
+\centering
+\begin{tabular}{|c|c|c|c|}
+\hline
+Iteration & $n$ active boxes & $n$ inactive boxes & $L$\\
+1 & 4 & 0 & $-\infty$\\
+2 & 16 & 0 & -560.82\\
+3 & 48 & 4 & -519.98\\
+4 & 84 & 31 & -480.58\\
+5 & 152 & 77 & -446.67\\
+6 & 184 & 183 & -419.47\\
+7 & 176 & 323 & -406.31\\
+8 & 164 & 458 & -395.36\\
+9 & 284 & 551 & -390.24\\
+10 & 496 & 711 & -385.84\\
+11 & 984 & 961 & -382.45\\
+12 & 1792 & 1497 & -379.65\\
+13 & 2880 & 2569 & -377.36\\
+14 & 4668 & 4282 & -375.37\\
+15 & 7312 & 7122 & -373.66\\
+16 & 10336 & 11850 & -372.29\\
+17 & 10108 & 19659 & -371.21\\
+18 & 18596 & 25118 & -370.82\\
+19 & 62488 & 28092 & -370.82\\
+20 & 124240 & 59520 & -370.70\\
+21 & 287596 & 111861 & -370.70\\
+22 & 494292 & 275884 & -370.70\\
+23 & 646808 & 608474 & -370.70\\
+24 & 291944 & 1182296 & -370.70\\
+25 & 0 & 1474240 & -370.70\\
+\hline
+\end{tabular}
+\caption{The state of the algorithm after each of 25 iterations for the GMST data.}
+\label{table:gmst_output}
+\end{table}
 
@@
     \centering
-    \includegraphics[width=.8\linewidth]{figs/gmst-boxes1.pdf}
+    \includegraphics[width=.8\linewidth]{figs/gmst-boxes1.jpg}
     \caption{Locations of boxes}
@@
     \centering
-    \includegraphics[width=.8\linewidth]{figs/gmst-rll1.pdf}
+    \includegraphics[width=.8\linewidth]{figs/gmst-rll1.jpg}
     \caption{Grayscale shows $L^b$ in each box.
@@
   \end{subfigure}
-  \caption{The 56,713 boxes from the first, coarse, analysis of global
-                mean surface temperature.}
+  \caption{The 1,474,240 boxes from the analysis of global mean surface temperature.}
 \label{fig:gmst1}
 \end{figure}
-
-The figure shows that the algorithm needed to divide boxes near the axes more finely than boxes away from the axes and that high $\logRLorig$ is found in only a small region of the plot and therefore suggests rerunning the algorithm with a smaller initial box while imposing more stringent algorithmic parameters.
-We limited the starting box to $\sigesq\in(50,500)$, $\sigssq\in(0,10^6)$ and set the control constants to
-\begin{equation*}
-	\maxit=20; \hspace{1cm} \epsilon=2; \hspace{1cm}
-	\delta_e=0; \hspace{1cm} \delta_s=0; \hspace{1cm} M=10.
-\end{equation*}
-After 20 iterations there were 1,986,000 active and 1,619,095 inactive boxes.
-Results are in Figure~\ref{fig:gmst2}.
-The figure agrees with Figure~15.3 in \cite{hodges:2013} and is detailed enough so no further iterations are needed.
-
-\begin{figure}
-  \begin{subfigure}{.5\textwidth}
-    \centering
-    \includegraphics[width=.8\linewidth]{figs/gmst-boxes2.jpg}
-    \caption{Locations of boxes}
-    \label{fig:gmst-boxes2}
-  \end{subfigure}
-  \begin{subfigure}{.5\textwidth}
-    \centering
-    \includegraphics[width=.8\linewidth]{figs/gmst-rll2.jpg}
-    \caption{Grayscale shows $L^b$ in each box.
-                  Red dot shows $b^* \equiv\argmax L^b$.
-	          Boxes outlined in blue have $U^b \ge L^{b^*}$.
-	          Axes are logarithmic.}
-    \label{fig:gmstrll2}
-  \end{subfigure}
-  \caption{The 3,605,095 boxes from the second analysis of global
-                mean surface temperature.}
-\label{fig:gmst2}
-\end{figure}
+The figure shows that the algorithm needed to divide boxes near the axes more finely than boxes away from the axes and that high $\logRLorig$ is found near $(200, 1000)$.
+The figure agrees with Figure~15.3 in \cite{hodges:2013}.
+
+%\begin{figure}
+%  \begin{subfigure}{.5\textwidth}
+%    \centering
+%    \includegraphics[width=.8\linewidth]{figs/gmst-boxes2.jpg}
+%    \caption{Locations of boxes}
+%    \label{fig:gmst-boxes2}
+%  \end{subfigure}
+%  \begin{subfigure}{.5\textwidth}
+%    \centering
+%    \includegraphics[width=.8\linewidth]{figs/gmst-rll2.jpg}
+%    \caption{Grayscale shows $L^b$ in each box.
+%                  Red dot shows $b^* \equiv\argmax L^b$.
+%	          Boxes outlined in blue have $U^b \ge L^{b^*}$.
+%	          Axes are logarithmic.}
+%    \label{fig:gmstrll2}
+%  \end{subfigure}
+%  \caption{The 3,605,095 boxes from the second analysis of global
+%                mean surface temperature.}
+%\label{fig:gmst2}
+%\end{figure}
 
 \section{Discussion}
-This paper has explained and illustrated an algorithm for linear mixed models with two variances, to find all regions where either the restricted likelihood function or the joint posterior density of the variances is high, and to evaluate the function there to arbitrary accuracy.
-A natural question to ask is \emph{What about linear mixed models with more than two variances?}  A partial answer is given by \cite{hodges:2013} who shows that some models with more than two variances can be reexpressed similarly to \eqref{eq:reexpress} but others can't.
-More complex models that can be reexpressed this way include, but are probably not limited to, models displaying general balance that are also orthogonal designs \citep[all balanced ANOVAs plus other models;][]{houtman_speed:1983}, models that are separable in a specific sense \citep[Section 17.1.5]{hodges:2013}, and miscellaneous other models \citep[Section 17.1.5]{hodges:2013}, e.g., a spatial model including random effects for heterogeneity and spatial clustering (an improper conditional autoregressive effect).
-We have not explored whether the reexpressible models can be analyzed by our algorithm; that's one direction for future work.
-
-Another is to see whether the algorithm can be used to advantage even in non-reexpressible models.
+This paper explains and illustrates an algorithm for linear mixed models with two variances, to find all regions where either the restricted likelihood function or the joint posterior density of the variances is high, and to evaluate the function there to arbitrary accuracy.
+A natural question to ask is \emph{What about linear mixed models with more than two variances?}  A partial answer is given by \cite{hodges:2013} who shows that some models with more than two variances can be re-expressed similarly to \eqref{eq:reexpress} but others can't.
+More complex models that can be re-expressed this way include, but are probably not limited to, models displaying general balance that are also orthogonal designs \citep[all balanced ANOVAs plus other models;][]{houtman_speed:1983}, models that are separable in a specific sense \citep[Section 17.1.5]{hodges:2013}, and miscellaneous other models \citep[Section 17.1.5]{hodges:2013}, e.g., a spatial model including random effects for heterogeneity and spatial clustering (an improper conditional autoregressive effect).
+We have not explored whether the re-expressible models can be analyzed by our algorithm; that's one direction for future work.
+
+Another is to see whether the algorithm can be used to advantage even in non-re-expressible models.
 If a model has, say, three variances and is now analyzed by, say, MCMC, we can create an MCMC chain that alternates between draws of $(\sigesq,\sigssq)$ and draws of the other variance.

### 2015-06-01 ms.tex
minor edits

@@
 Equations~\eqref{eq:reexpress} and~\eqref{eq:logpost} can both be written as a sum of multiples of logs and inverses of linear combinations $\abss$, as in \eqref{eq:sum}, where the summands with $j=s_z+1$ and $j=s_z+2$ are for the terms involving only $\sigesq$ and $\sigssq$, respectively and where we have dropped the irrelevant constant $B$.
+I.e.,
 \begin{equation}
@@
 The $j$'th term is constant on the $j$'th line and attains its maximum there.
-%Both \eqref{eq:reexpress} and \eqref{eq:logpost} contain a term, call it the $v$'th (for vertical) term,  with $a_j=0$.
-The $v$'th term in (\ref{eq:partials}) is negative to the right of the $v$'th line, positive to the left, and 0 on the line; hence the $v$'th term in \eqref{eq:logpost} is maximized on the $v$'th line.
-\eqref{eq:logpost} also contains a term, call it the $h$'th (for horizontal) term, with $b_j=0$.
-The corresponding line is horizontal at $\sigssq = \sigma_e^{2h} \equiv d_h / a_h c_h$.
-The $h$'th term in \eqref{eq:logpost} is maximized on the $h$'th line.
 
@@
 \end{figure}
-
-%\subsection{Lines Determine a Bounding Box}
-%\label{sec:boundingbox}
-%
-%Figure~(\ref{fig:bigboundingbox}) shows five lines --- realistic data sets may have more --- labelled $j=1, 2, 3, s_z+1, s_z+2$.
-The largest intercept of those lines on the $\sigssq$ axis and the largest intercept on the $\sigesq$ axis determine a rectangle $B_1$ whose outline is marked in bold.
-At any point above $B_1$, e.g., $p_1$, the partial derivatives \eqref{eq:partials} are either negative or, for $j=s_z+1$, the partial derivative \eqref{eq:drlds} is zero.
-Therefore, $\logRL(\widetilde{p_1}) \ge \logRL(p_1)$.
-Similarly, $\logRL(\widetilde{p_2}) \ge \logRL(p_2)$ and $\logRL(\widetilde{p_3}) \ge \logRL(p_3)$.
-That is, for points above and/or to the right of $B_1$, $\logRLss$ can be increased by moving down and/or to the left as far as $B_1$'s boundary.
-Therefore \mrle\ must lie on or inside $B_1$ and there can be no maxima of $\logRLss$ in $B_1^c$.
-$B_1$ could be passed to an optimizer such as \R's \textsf{optim} or \textsf{nlminb} with potentially better results than using those functions without bounds.
-However, even with known bounds, general purpose optimizers may still miss \mrle\ and regions of high $\logRL$.
-This paper presents a computational method guaranteed not to miss \mrle\ and which evaluates $\logRLss$ everywhere to within a specified tolerance inside a box $B$ satisfying desideratum \textbf{D1}.
-First, though, we pause to note that the region outlined in bold in Figure~(\ref{fig:smallboundingbox}) is a subset of $B_1$ that must also contain \mrle\ for the reasons given above.
-But it is not rectangular, hence less convenient than $B_1$, so we don't pursue it further.
-%
-%\begin{figure}[h]
-%  \begin{subfigure}{.5\textwidth}
-%	\centering
-%	\includegraphics[width=.8\linewidth]{figs/boundingbox.pdf}
-%	\caption{A rectangular region $B_1$ containing all maxima.}
-%	\label{fig:bigboundingbox}
-%  \end{subfigure}
-%  \begin{subfigure}{.5\textwidth}
-%	\centering
-%	\includegraphics[width=.8\linewidth]{figs/smallboundingregion.pdf}
-%	\caption{A smaller region containing all maxima.}
-%	\label{fig:smallboundingbox}
-%  \end{subfigure}
-%  \caption{all local and global maxima lie in the regions bounded by the solid
-%                dark line.
-%               }
-%  \label{fig:boundingbox}
-%\end{figure}
-%
-%To find $B$, choose an arbitrary constant $M>0$ and an arbitrary point inside $B_1$, say $p$, as illustrated in Figure~(\ref{fig:boundingbox2}), and define $L \equiv \logRL(p)$.
-We will find a box $B\supset B_1$, as shown in Figure~(\ref{fig:boundingbox2}), outside of which $\logRLss \le L-M$, thus satisfying \textbf{D1}.
-%
-%$B$ is determined by its intercepts $\widetilde{\sigma_e^2}$ and $\widetilde{\sigma_s^2}$ on the $\sigesq$ and $\sigssq$ axes respectively.
-Let $\widetilde{q_1} = (0,\widetilde{\sigma_s^2})$ be the intercept of $B$ and the $\sigssq$ axis and let $\widetilde{q_2} = (\sigma_e^{2*},\widetilde{\sigma_s^2})$.
-Points such as $q_1$ which satisfy $\sigesq \le \sigma_e^{2*}$ and $\sigssq \ge \widetilde{\sigma_s^2}$ --- above $B$ and to the left of the $s_z+1$ line --- are in a region where the partial derivatives (\ref{eq:drlds}, \ref{eq:drlde}) are negative for $j \in \{1, \dots, s_z, s_z+2\}$ but nonnegative for $j=s_z+1$.
-Let $\logRL_j(\cdot)$ denote the $j$'th term of \eqref{eq:sum} evaluated at $(\cdot)$.
-%\begin{equation}
-%	\logRL(\sshat) - \logRL(q_1) \ge L - \sum_{j \ne s_z+1} \logRL_j(\widetilde{q_1}) - \logRL_{s_z+1}(\widetilde{q_2}),
-%\end{equation}
-%which is larger than $M$ if
-%\begin{equation}
-%\label{eq:q1}
-%  \begin{split}
-%    \sum_{j \ne s_z+1} \logRL_j(\widetilde{q_1})
-%      &< L - M - \logRL_{s_z+1}(\widetilde{q_2})\\
-%      &= L - M + \frac{1}{2} \left[ c_{s_z+1}\log\sigma_e^{2*} + \frac{d_{s_z+1}}{\sigma_e^{2*}}\right].
-%  \end{split}
-%\end{equation}
-%Examination of (\ref{eq:sum}) shows that for any fixed $\sigesq$, in particular for $\sigesq=0$, and for $j \ne s_z+1$,\\ $\lim_{\sigssq \rightarrow \infty} \logRL_j(\sigesq,\sigssq) = -\infty$.
-Thus $\widetilde{\sigma_s^2}$ can be chosen large enough so that the summation on the l.h.s.~of \eqref{eq:q1} is less than the r.h.s.~of \eqref{eq:q1}, and  \textbf{D1} is satisfied.
-%
-%Points such as $q_2$ which satisfy $\sigesq \ge \sigma_e^{2*}$ and $\sigssq \ge \widetilde{\sigma_s^2}$ are in a region where the partial derivatives (\ref{eq:drlds}, \ref{eq:drlde}) are nonpositive for $j \in \{1, \dots, s_z+2\}$ and therefore, $\logRL(\sshat) - \logRL(q_2) \ge L - \logRL(\widetilde{q_2})$.
-The latter expression is greater than $M$ iff $\logRL(\widetilde{q_2}) < L-M$.
-But examination of (\ref{eq:sum}) shows that for any fixed $\sigesq$, $\lim_{\sigssq \rightarrow \infty} \logRLss = -\infty$ and therefore $\widetilde{\sigma_s^2}$ can be chosen large enough so that \textbf{D1} is satisfied.
-%
-%Let $\widetilde{q_3} = (\widetilde{\sigma_e^2}, \sigma_s^{2*})$.
-Points such as $q_3$ which satisfy $\sigesq \ge \widetilde{\sigma_e^2}$ and $\sigssq \ge \sigma_s^{2*}$ are in a region where the partial derivatives (\ref{eq:drlds}, \ref{eq:drlde}) are nonpositive for $j \in \{1, \dots, s_{z+2}\}$ and therefore, $\logRL(\sshat) - \logRL(q_3) \ge L - \logRL(\widetilde{q_3})$.
-The latter expression is greater than $M$ iff $\logRL(\widetilde{q_3}) < L-M$.
-But examination of (\ref{eq:sum}) shows that for any fixed $\sigssq$, $\lim_{\sigesq \rightarrow \infty} \logRLss = -\infty$ and therefore $\widetilde{\sigma_e^2}$ can be chosen large enough so that \textbf{D1} is satisfied.
-%
-%Let $\widetilde{q_4} = (\widetilde{\sigma_e^2},0)$.
-For any point $q_4$ having $\sigssq \le \sigma_s^{2*}$ and $\sigesq \ge \widetilde{q_4}$, the partial derivatives (\ref{eq:drlds}, \ref{eq:drlde}) are nonpositive for $j \in \{1, \dots, s_z+1\}$  but nonnegative for $j=s_z+2$.
-Therefore,
-%\begin{equation*}
-%	\logRL(\sshat) - \logRL(q_4) \ge L - \sum_{j \ne s_z+2} \logRL_j(\widetilde{q_4}) - \logRL_{s_z+2}(\widetilde{q_3})
-%\end{equation*}
-%which is larger than $M$ if
-%\begin{equation}
-%\label{eq:q4}
-%  \begin{split}
-%    \sum_{j \ne s_z+2} \logRL_j(\widetilde{q_4})
-%      &< L - M - \logRL_{s_z+2}(\widetilde{q_3})\\
-%      &= L - M + \frac{1}{2} \left[ c_{s_z+2}\log\sigma_s^{2*} + \frac{d_{s_z+2}}{\sigma_s^{2*}}\right].
-%  \end{split}
-%\end{equation}
-%Examination of (\ref{eq:sum}) shows that for any fixed $\sigssq$, in particular for $\sigssq=0$, and for $j \ne s_z+2$,\\ $\lim_{\sigesq \rightarrow \infty} \logRL_j(\sigesq,\sigssq) = -\infty$.
-Thus $\widetilde{\sigma_e^2}$ can be chosen large enough so that the summation on the l.h.s.~of \eqref{eq:q4} is less than the r.h.s.~of \eqref{eq:q4}, and  \textbf{D1} is satisfied.
-%
-%Combining the requirements for points like $q_1$, $q_2$, $q_3$, and $q_4$, desideratum \textbf{D1} is satisfied for all points in $B^c$ by choosing $\widetilde{\sigma_e^2}$ and $\widetilde{\sigma_s^2}$ sufficiently large.
-%%enough to satisfy $\logRL(q_1^*) < L-M$ and choosing $\sigma_s^*$ large enough to satisfy $\logRL(q_2^*) < L-M$ and $\sum_{1}^{s_z} \logRL_j(q_3^*) < L - M + n_e(\log(\sigma_e^{2v}) + 1)/2$.
-In addition, because $\logRL$ has no maxima outside of $B_1$, it also has no maxima outside of $B$.
-As already mentioned, if there is an $h$ line, then $q_1$ would have to be treated like $q_2$ and $q_3$.
 
@@
 \textbf{QED}\\
-%enough to satisfy $\logRL(q_1^*) < L-M$ and choosing $\sigma_s^*$ large enough to satisfy $\logRL(q_2^*) < L-M$ and $\sum_{1}^{s_z} \logRL_j(q_3^*) < L - M + n_e(\log(\sigma_e^{2v}) + 1)/2$.
-In addition, because $\logRL$ has no maxima outside of $B_1$, it also has no maxima outside of $B$.
-As already mentioned, if there is an $h$ line, then $q_1$ would have to be treated like $q_2$ and $q_3$.
 
@@
 \item create a list of active boxes, initially consisting of just $B^0$;
-\item create a list of inactive boxes, initially empty;
-\item for each active box $B_j$, find $(L^b_j, U^b_j)$ and determine whether $B_j$ needs to be subdivided.
+\item create a list of inactive boxes, initially empty; and
+\item for each active box $B_j$, find $(L^{B_j}, U^{B_j})$ and determine whether $B_j$ needs to be subdivided.
 \end{enumerate*}
@@
 
-\item A \textcompute{box} is a list consisting of
+\item A \textcompute{box} $b$ is a list consisting of
 	 \begin{itemize}
@@
 \end{equation*}
-The algorithm finished in 22 iterations and took a little under 3 hours.  Table~\ref{table:hmo_HH11Bayes} shows the output.
+The algorithm finished in 22 iterations and took a little under 3 hours.  Table~\ref{table:hmo_HH11Bayes} shows the output and Figure~\ref{fig:hmoBayes} shows the boxes.
 \begin{table}[H]

### 2015-06-02 ms.tex
some edits.  delete p3 from Fig 3

@@
 Books are written about them (e.g.\ \citealt{bryk_raudenbush:1992, verbeke&molenberghs:2000, hodges:2013, west&welch&galecki:2014}),  courses are taught about them, and they have many applications.
-Examples include random intercept models (including balanced and unbalanced one-way random effect models), additive models with one penalized spline, spatial models with one intrinsic conditional autoregression (ICAR) random effect, dynamic linear models with one system-level variance, and some multiple membership models (e.g.\ \citealt{browne_etal:2001, mccaffrey_etal:2004}).
 Typical notation, which we adopt, is
@@
 The key for this manuscript is that $\theta$ contains only those two unknown variances and no others.
-\cite{hodges:2013} gives examples and explains the importance of this special case.
+Examples include random intercept models (including balanced and unbalanced one-way random effect models), additive models with one penalized spline, spatial models with one intrinsic conditional autoregression (ICAR) random effect, dynamic linear models with one system-level variance, and some multiple membership models (e.g.\ \citealt{browne_etal:2001, mccaffrey_etal:2004}).
+\cite{hodges:2013} examines these and other examples and explains the importance of this special case.
 
@@
 \end{equation}
-Equations~\eqref{eq:reexpress} and~\eqref{eq:logpost} can both be written as a sum of multiples of logs and inverses of linear combinations $\abss$, as in \eqref{eq:sum}, where the summands with $j=s_z+1$ and $j=s_z+2$ are for the terms involving only $\sigesq$ and $\sigssq$, respectively and where we have dropped the irrelevant constant $B$.
+Equations~\eqref{eq:reexpress} and~\eqref{eq:logpost} can both be written as a sum of multiples of logs and inverses of linear combinations $\abss$, as in \eqref{eq:sum}, where the summands with $j=s_z+1$ and $j=s_z+2$ are for the terms involving only $\sigesq$ and $\sigssq$, respectively, and where we have dropped the irrelevant constant $B$.
 I.e.,
@@
     c_j &=1\\
-    d_j &= \hat v_j^2\ \text{and is derived in the Appendix}
+    d_j &= \hat v_j^2\ \text{is a known function of $y$ derived in the Appendix}
   \end{align*}
@@
 
-Our point of view is that it is important to find regions where $f$ or $\log f$ is large relative to its maximum regardless of whether those regions contain local maxima.
-Points with large $\logRLss$ are those that describe the data, and possibly also prior information, well, at least compared to points with low $\logRLss$.
+Our view is that it is important to find regions where $f$ or $\log f$ is large relative to its maximum regardless of whether those regions contain local maxima.
+Points with large $\logRLss$ are those that describe the data, or possibly  prior information, well, at least compared to points with low $\logRLss$.
 If $f$ or $\log f$ is relatively flat and large over a region, it matters little whether the region contains small bumps that are, technically, local maxima.
@@
 \end{equation}
-In practice we will take $M$ to be large enough to interpret (\ref{eq:boundingbox}) as meaning that we can restrict attention to $B$ because values of $(\sigesq, \sigssq) \in B^c$ have $\logRLss$ too low to be of further interest.
+In practice we will take $M$ large enough to interpret (\ref{eq:boundingbox}) as meaning that we can restrict attention to $B$ because values of $(\sigesq, \sigssq) \in B^c$ have $\logRLss$ too low to be of further interest.
 
@@
 \textbf{D1} and \textbf{D2} allow us to divide the $(\sigesq,\sigssq)$ plane into two parts: one where $\logRL$ is at least $M$ below its maximum and another where we know $\logRL$ to within a pre-specified $\epsilon$.
-The next section shows how the partial derivatives are used to satisfy \textbf{D1} and \textbf{D2}.
+The next section shows how the partial derivatives are used to satisfy \textbf{D1} and \textbf{D2}: the partial derivatives determine lines in the $(\sigesq,\sigssq)$ plane; those lines determine a box $B_1$ containing all local maxima and a larger box $B \supset B_1$ satisfying \textbf{D1}; and those lines also determine upper and lower bounds on $\logRL$ within any box $b$.
 
@@
 For $j=s_z+1$, $a_j=0$, so $\eqref{eq:drlds} = 0$ and \eqref{eq:partials} determines a vertical line at $\sigesq = \sigma_e^{2*} \equiv d_{s_z+1} / c_{s_z+1}$.
-For $j=s_z+2$,  $b_j=0$, so $\eqref{eq:drlde} = 0$ and, if the target function is \eqref{eq:logpost}, \eqref{eq:partials} determines a horizontal line at $\sigssq = \sigma_s^{2*} \equiv d_{s_z+2} / c_{s_z+2}$, while if the target function is \eqref{eq:reexpress}, $c_{s_z+2} = d_{s_z+2} = 0$; the term for $j=s_z+2$ effectively vanishes; and there is no horizontal line.
+For $j=s_z+2$,  $b_j=0$, so $\eqref{eq:drlde} = 0$ and, if the target function is \eqref{eq:logpost}, \eqref{eq:partials} determines a horizontal line at $\sigssq = \sigma_s^{2*} \equiv d_{s_z+2} / c_{s_z+2}$, while if the target function is \eqref{eq:reexpress}, $c_{s_z+2} = d_{s_z+2} = 0$; the term for $j=s_z+2$ effectively vanishes, and there is no horizontal line.
 
@@
 	\label{fig:oneline}
-\end{figure}
-
-\begin{figure}[h]
-	\label{fig:boundingbox2}
-	\centering
-	\includegraphics[width=.5\linewidth]{figs/boundingbox2.pdf}
-	\caption{Box $B$ satisfies (\ref{eq:boundingbox})}
 \end{figure}
@@
 \textbf{Proof} Let $p_1 = (\sigesq{}_1, \sigssq{}_1)$ be a point such that $\sigssq{}_1 > \sigssq{}^M$ and let $\widetilde{p_1} = (\sigesq{}_1, \sigssq{}^M)$, as illustrated in Figure~(\ref{fig:bigboundingbox}).
-For $j=1, \dots, s_z, s_z+2$, the partial derivatives \eqref{eq:drlde} are negative; for $j=s_z+1$, the partial derivative is zero.
+For $j=1, \dots, s_z, s_z+2$, the partial derivatives \eqref{eq:drlde} are negative everywhere between $p_1$ and $\widetilde{p_1}$; for $j=s_z+1$, the partial derivative is zero.
 Therefore, $\logRL(\widetilde{p_1}) \ge \logRL(p_1)$ and there can be no local maxima of $\logRL$ above the line $\sigssq = \sigssq{}^M$.
 Let $p_2 = (\sigesq{}_2, \sigssq{}_2)$ be a point such that $\sigesq{}_2 > \sigesq{}^M$ and let $\widetilde{p_2} = (\sigesq{}^M, \sigssq{}_2)$.
-For $j=1, \dots, s_z, s_z+1$, the partial derivatives \eqref{eq:drlds} are negative; for $j=s_z+2$, the partial derivative is zero.
+For $j=1, \dots, s_z, s_z+1$, the partial derivatives \eqref{eq:drlds} are negative everywhere between $p_2$ and $\widetilde{p_2}$; for $j=s_z+2$, the partial derivative is zero.
 Therefore, $\logRL(\widetilde{p_2}) \ge \logRL(p_2)$ and there can be no local maxima of $\logRL$ to the right of the line $\sigesq = \sigesq{}^M$.
@@
 By the claim, \mrle\ must lie on or inside $B_1$, so $B_1$ could be passed to an optimizer such as \R's \textsf{optim} or \textsf{nlminb} with potentially better results than using those functions without bounds.
-However, even with known bounds, general purpose optimizers may still miss \mrle\ and regions of high $\logRL$.
+However, even with known bounds, general purpose optimizers may still miss \mrle\ and regions of high $\logRL$ and they emphasize single points of highest local $\logRL$ while possibly ignoring large adjacent regions where $\logRL$ is nearly as large \citep{hill:1965, hodges:2013}.
 We will next find a box $B \supset B_1$ satisfying desideratum \textbf{D1}.
@@
 Let $q_1 = (\sigesq{}_1, \sigssq{}_1)$ be a point satisfying $\sigesq{}_1 \le \sigma_e^{2*}$ and $\sigssq{}_1 \ge \widetilde{\sigma_s^2}$ --- above $B$ and to the left of the $s_z+1$ line.
-The partial derivatives (\ref{eq:drlds}), evaluated at $q_1$ are negative for $j \in \{1, \dots, s_z, s_z+2\}$ but zero for $j=s_z+1$.
-The partial derivatives (\ref{eq:drlde}), evaluated at $q_1$ are negative for $j \in \{1, \dots, s_z, s_z+1\}$ but nonnegative for $j=s_z+2$.
+The partial derivatives (\ref{eq:drlds}), between $q_1$ and $\widetilde{q_1}$ are negative for $j \in \{1, \dots, s_z, s_z+2\}$ but zero for $j=s_z+1$.
+The partial derivatives (\ref{eq:drlde}), betweem $q_1$ and $\widetilde{q_2}$ are negative for $j \in \{1, \dots, s_z, s_z+1\}$ but nonnegative for $j=s_z+2$.
 Let $\logRL_j(\cdot)$ denote the $j$'th term of \eqref{eq:sum} evaluated at $(\cdot)$.

### 2015-06-02 ms.R

@@ -163,12 +163,11 @@ p + annotate ( "line", x=x1, y=y1, lty=3 ) +
   annotate ( "text", x=xhann, y=yhann, size=4, parse=TRUE, label="j==s[Z]+2" ) +
   geom_path ( data=B1, size=.7 ) +
   annotate ( "text", x=B1annx, y=B1anny, size=4, parse=TRUE, label="B[1]") +
-  annotate ( "point", x=c(.5,.5, 1.05, 1, 1.05, 1), y=c(1.2,1,.5,.5,1.1,1) ) +
-  annotate ( "text", x=c(.55,.55,1.1,.95,1.05,1.05),
-                     y=c(1.2,1.05,.5,.5,1.15,1), parse=TRUE,
+  annotate ( "point", x=c(.5,.5, 1.05, 1), y=c(1.2,1,.5,.5) ) +
+  annotate ( "text", x=c(.55,.55,1.1,.95),
+                     y=c(1.2,1.05,.5,.5), parse=TRUE,
              label = c ( "p[1]", "widetilde(p[1])",
-                         "p[2]", "widetilde(p[2])",
-                         "p[3]", "widetilde(p[3])" )
+                         "p[2]", "widetilde(p[2])" )
            )
 dev.off()
 

### 2015-06-02 ms.tex
mostly Andrew's edits

@@
 For $j=s_z+1$, $a_j=0$, so $\eqref{eq:drlds} = 0$ and \eqref{eq:partials} determines a vertical line at $\sigesq = \sigma_e^{2*} \equiv d_{s_z+1} / c_{s_z+1}$.
-For $j=s_z+2$,  $b_j=0$, so $\eqref{eq:drlde} = 0$ and, if the target function is \eqref{eq:logpost}, \eqref{eq:partials} determines a horizontal line at $\sigssq = \sigma_s^{2*} \equiv d_{s_z+2} / c_{s_z+2}$, while if the target function is \eqref{eq:reexpress}, $c_{s_z+2} = d_{s_z+2} = 0$; the term for $j=s_z+2$ effectively vanishes, and there is no horizontal line.
+For $j=s_z+2$,  $b_j=0$, so $\eqref{eq:drlde} = 0$ and, if the target function is \eqref{eq:logpost}, \eqref{eq:partials} determines a horizontal line at $\sigssq = \sigma_s^{2*} \equiv d_{s_z+2} / c_{s_z+2}$, while if the target function is \eqref{eq:reexpress}, $c_{s_z+2} = d_{s_z+2} = 0$; the term for $j=s_z+2$ effectively vanishes and there is no horizontal line.
 
@@
 Let $B_1$ be the box whose lower-left and upper-right corners are $(0,0)$ and $(\sigesq{}^M,\sigssq{}^M)$, respectively.
-Figure~(\ref{fig:bigboundingbox}) illustrates with five lines --- realistic data sets may have more --- labelled $j=1, 2, 3, s_z+1, s_z+2$.\\
+Figure~(\ref{fig:bigboundingbox}) shows $B_1$ and five lines --- realistic data sets may have more --- labelled $j=1, 2, 3, s_z+1, s_z+2$.\\[5pt]
 \textbf{Claim} \emph{All local maxima of $\logRL$ lie within $B_1$.}\\
@@
 \textbf{QED}\\
-
-By the claim, \mrle\ must lie on or inside $B_1$, so $B_1$ could be passed to an optimizer such as \R's \textsf{optim} or \textsf{nlminb} with potentially better results than using those functions without bounds.
-However, even with known bounds, general purpose optimizers may still miss \mrle\ and regions of high $\logRL$ and they emphasize single points of highest local $\logRL$ while possibly ignoring large adjacent regions where $\logRL$ is nearly as large \citep{hill:1965, hodges:2013}.
-We will next find a box $B \supset B_1$ satisfying desideratum \textbf{D1}.
-First, though, we pause to note that the region outlined in bold in Figure~(\ref{fig:smallboundingbox}) is a subset of $B_1$ that must also contain \mrle\ by the same reasoning used to prove the previous claim.
-But it is not rectangular, hence less convenient than $B_1$, so we don't pursue it further.
 
@@
 \end{figure}
+
+By the claim, \mrle\ must lie on or inside $B_1$, so $B_1$ could be passed to an optimizer such as \R's \textsf{optim} or \textsf{nlminb} with potentially better results than using those functions without bounds.
+However, even with known bounds, general purpose optimizers may still miss \mrle\ and regions of high $\logRL$ and they emphasize single points of highest local $\logRL$ while possibly ignoring large adjacent regions where $\logRL$ is nearly as large \citep{hill:1965, hodges:2013}.
+
+We will next find a box $B \supset B_1$ satisfying desideratum \textbf{D1}.
+First, though, we pause to note that the region outlined in bold in Figure~(\ref{fig:smallboundingbox}) is a subset of $B_1$ that must also contain \mrle\ by the same reasoning used to prove the previous claim.
+But it is not rectangular, hence less convenient than $B_1$, so we don't pursue it further.\\[5pt]
 
@@
 Let $q_2 = (\sigesq{}_2, \sigssq{}_2)$ satisfy $\sigesq{}_2 \ge \sigma_e^{2*}$ and $\sigssq{}_2 \ge \widetilde{\sigma_s^2}$ --- above $B$ and to the right of the $s_z+1$ line.
-(Points $q_2$ and $q_3$ in Figure~(\ref{fig:boundingbox2}) both satisfy this criterion.)  The partial derivatives (\ref{eq:drlds}, \ref{eq:drlde}) evaluated at $q_2$ are nonpositive for $j \in \{1, \dots, s_z+2\}$; therefore, $\logRL(\sshat) - \logRL(q_2) \ge L - \logRL(\widetilde{q_2})$.
+(Points $q_2$ and $q_3$ in Figure~(\ref{fig:boundingbox2}) both satisfy these criteria.)  The partial derivatives (\ref{eq:drlds}, \ref{eq:drlde}) evaluated at $q_2$ are nonpositive for $j \in \{1, \dots, s_z+2\}$; therefore, $\logRL(\sshat) - \logRL(q_2) \ge L - \logRL(\widetilde{q_2})$.
 The latter expression is greater than $M$ iff $\logRL(\widetilde{q_2}) < L-M$.
@@
 \end{figure}
-Next we consider the relationship between the $\logRL_j$'s and boxes as depicted in Figure~\ref{fig:boxb}, which shows a box $b$ and three lines.
+Next we consider the relationship between the $\logRL_j$'s and boxes as depicted in Figure~\ref{fig:boxb}, which shows a small box $b$ and three lines.
 For lines such as $j=2$ that lie below or to the left of $b$, the partial derivatives (\ref{eq:drlds}, \ref{eq:drlde}) are nonpositive, so the maximum and minimum of $\logRL_j$ within $b$ are attained at the lower left and upper right corners respectively.
@@
 \section{Examples}
-
+\label{sec:examp}
+Throughout Section~\ref{sec:examp} we use $\epsilon = 1$, $M = 7$, $\delta_e = \delta_s = 0$ and $\maxit = 25$.
+We used a finite value of $\maxit$ for practicality but, because each example took 25 or fewer iterations to adequately resolve $\log f$, $\maxit = 25$ yields the same results as $\maxit = \infty$.
+Computations were performed on an iMac that was new in 2011, having a 2.7 GHz Intel Core i5 processor and 4GB of memory.
 \subsection{HMO premiums}
@@
 The algorithm did not need to divide the boxes with large $\sigesq$ or $\sigssq$ as finely as those with small $\sigesq$ and $\sigssq$ because they more readily satisfy either $U^b < L-M$ or $U^b - L^b < \epsilon$, so become inactive.
-Figure~\ref{fig:hmorll} shows the same boxes, shaded by $L^b$ in each box.
+Figure~\ref{fig:hmorll} shows the same boxes on a logarithmic scale, shaded by $L^b$ in each box.
 The red dot is the lower left corner of the box that maximizes $L^b$.

### 2015-06-02 ms.tex
proof of second claim

@@
 
-Combining the requirements for points like $q_1$, $q_2$, $q_3$, and $q_4$, desideratum \textbf{D1} is satisfied for all points in $B^c$ by choosing $\widetilde{\sigma_e^2}$ and $\widetilde{\sigma_s^2}$ sufficiently large.
+\noindent\textbf{Alt Proof}
+Let $p$ be an arbitrary point inside $B_1$, as illustrated in Figure~(\ref{fig:boundingbox2}), and define $L \equiv \logRL(p)$; we use $L$ as a lower bound on $\logRL(\sshat)$.
+Let $\widetilde{q_1} = (0,\widetilde{\sigma_s^2})$ be the intercept of $B$ and the $\sigssq$ axis; let $\widetilde{q_2} = (\sigma_e^{2*},\widetilde{\sigma_s^2})$; let $\widetilde{q_3} = (\widetilde{\sigma_e^2}, \sigma_s^{2*})$; let $\widetilde{q_4} = (\widetilde{\sigma_e^2},0)$; and let $\logRL_j$ denote the $j$'th term of \eqref{eq:sum}.
+The proof considers in turn four regions $Q_1$, $Q_2$, $Q_3$, and $Q_4$ whose union is $B^c$ and shows that $\widetilde{\sigma_e^2}$ and $\widetilde{\sigma_s^2}$ can be chosen large enough so that \textbf{D1} is satisfied on each region.
+\begin{itemize}
+\item $Q_1: \sigesq \le \sigesq{}^*$ and $\sigssq \ge \widetilde{\sigssq}$, as illustrated by $q_1$ in
+  Figure~\ref{fig:boundingbox2}.  In $Q_1$, by inspection of \eqref{eq:sum} and \eqref{eq:partials},
+  \begin{itemize}
+  \item for $j \ne s_z+1$,
+    \begin{equation*}
+      \frac{\partial\logRL_j}{\partial\sigesq} \le 0 \hspace{1cm} \text{and} \hspace{1cm}
+      \frac{\partial\logRL_j}{\partial\sigssq} \le 0
+    \end{equation*}
+    so for any $q \in Q_1$,
+    \begin{equation*}
+      \logRL_j(q) \le \logRL_j(\widetilde{q_1});
+    \end{equation*}
+  \item  for $j = s_z+1$,
+    \begin{equation*}
+      \frac{\partial\logRL_j}{\partial\sigesq} > 0 \hspace{1cm} \text{and} \hspace{1cm}
+      \frac{\partial\logRL_j}{\partial\sigssq} = 0
+    \end{equation*}
+    so for any $q \in Q_1$,
+    \begin{equation*}
+      \logRL_j(q) \le \logRL_j(\widetilde{q_2}).
+    \end{equation*}
+  \end{itemize}
+  Therefore, 
+  \begin{equation}
+	\logRL(\sshat) - \logRL(q) \ge L - \sum_{j \ne s_z+1} \logRL_j(\widetilde{q_1}) - \logRL_{s_z+1}(\widetilde{q_2}),
+  \end{equation}
+  which is larger than $M$ if
+  \begin{equation}
+    \label{eq:q1}
+    \begin{split}
+      \sum_{j \ne s_z+1} \logRL_j(\widetilde{q_1})
+        &< L - M - \logRL_{s_z+1}(\widetilde{q_2})\\
+        &= L - M + \frac{1}{2} \left[ c_{s_z+1}\log\sigma_e^{2*} + \frac{d_{s_z+1}}{\sigma_e^{2*}}\right].
+    \end{split}
+  \end{equation}
+  Examination of (\ref{eq:sum}) shows that for any fixed $\sigesq$, in particular for $\sigesq=0$,
+  and for $j \ne s_z+1$,\\ $\lim_{\sigssq \rightarrow \infty} \logRL_j(\sigesq,\sigssq) = -\infty$.
+  Thus $\widetilde{\sigma_s^2}$ can be chosen large enough so that the summation on the
+  l.h.s.~of \eqref{eq:q1} is less than the r.h.s.~of \eqref{eq:q1}, and  \textbf{D1} is satisfied.
+\item $Q_2: \sigesq \ge \sigesq{}^*$ and $\sigssq \ge \widetilde{\sigssq}$, as illustrated by $q_2$ and $q_3$
+  in Figure~\ref{fig:boundingbox2}.  In $Q_2$, for all $j$,
+  \begin{equation*}
+    \frac{\partial\logRL_j}{\partial\sigesq} \le 0 \hspace{1cm} \text{and} \hspace{1cm}
+    \frac{\partial\logRL_j}{\partial\sigssq} \le 0
+  \end{equation*}
+  so for any $q \in Q_2$,
+  \begin{equation*}
+    \logRL_j(q) \le \logRL_j(\widetilde{q_2}).
+  \end{equation*}
+  Therefore, $\logRL(\sshat) - \logRL(q) \ge L - \logRL(\widetilde{q_2})$.  The latter expression is greater than
+  $M$ iff $\logRL(\widetilde{q_2}) < L-M$.  But (\ref{eq:sum}) shows that for any fixed
+  $\sigesq$, $\lim_{\sigssq \rightarrow \infty} \logRLss = -\infty$ and therefore $\widetilde{\sigma_s^2}$
+  can be chosen large enough so that \textbf{D1} is satisfied.
+\item $Q_3: \sigesq \ge \widetilde{\sigesq}$ and $\sigssq \ge \sigssq{}^*$, as illustrated by $q_3$.  In $Q_3$,
+  for all $j$,
+  \begin{equation*}
+    \frac{\partial\logRL_j}{\partial\sigesq} \le 0 \hspace{1cm} \text{and} \hspace{1cm}
+    \frac{\partial\logRL_j}{\partial\sigssq} \le 0
+  \end{equation*}
+  so for any $q \in Q_3$,
+  \begin{equation*}
+    \logRL_j(q) \le \logRL_j(\widetilde{q_3}).
+  \end{equation*}
+  Therefore, $\logRL(\sshat) - \logRL(q) \ge L - \logRL(\widetilde{q_3})$.  The latter expression is greater than
+  $M$ iff $\logRL(\widetilde{q}) < L-M$.  But for any fixed $\sigssq$,
+  $\lim_{\sigesq \rightarrow \infty} \logRLss = -\infty$ and therefore $\widetilde{\sigma_e^2}$
+  can be chosen large enough so that \textbf{D1} is satisfied.
+\item $Q_4: \sigesq \ge \widetilde{\sigesq}$ and $\sigssq \le \sigssq{}^*$, as illustrated by $q_4$.  In $Q_4$,
+  \begin{itemize}
+  \item for $j \ne s_z+2$,
+    \begin{equation*}
+      \frac{\partial\logRL_j}{\partial\sigesq} \le 0 \hspace{1cm} \text{and} \hspace{1cm}
+      \frac{\partial\logRL_j}{\partial\sigssq} < 0
+    \end{equation*}
+    so for any $q \in Q_4$,
+    \begin{equation*}
+      \logRL_j(q) \le \logRL_j(\widetilde{q_4});
+    \end{equation*}
+  \item  for $j = s_z+2$,
+    \begin{equation*}
+      \frac{\partial\logRL_j}{\partial\sigesq} = 0 \hspace{1cm} \text{and} \hspace{1cm}
+      \frac{\partial\logRL_j}{\partial\sigssq} \ge 0
+    \end{equation*}
+    so for any $q \in Q_4$,
+    \begin{equation*}
+      \logRL_j(q) \le \logRL_j(\widetilde{q_3}).
+    \end{equation*}
+  \end{itemize}
+  Therefore,
+  \begin{equation*}
+	\logRL(\sshat) - \logRL(q) \ge L - \sum_{j \ne s_z+2} \logRL_j(\widetilde{q_4}) - \logRL_{s_z+2}(\widetilde{q_3})
+  \end{equation*}
+  which is larger than $M$ if
+  \begin{equation}
+  \label{eq:q4}
+    \begin{split}
+      \sum_{j \ne s_z+2} \logRL_j(\widetilde{q_4})
+        &< L - M - \logRL_{s_z+2}(\widetilde{q_3})\\
+        &= L - M + \frac{1}{2} \left[ c_{s_z+2}\log\sigma_s^{2*} + \frac{d_{s_z+2}}{\sigma_s^{2*}}\right].
+    \end{split}
+  \end{equation}
+  For any fixed $\sigssq$, in particular for $\sigssq=0$, and for $j \ne s_z+2$,\\ $\lim_{\sigesq \rightarrow \infty}
+  \logRL_j(\sigesq,\sigssq) = -\infty$.  Thus $\widetilde{\sigma_e^2}$ can be chosen large enough so that the
+  summation on the l.h.s.~of \eqref{eq:q4} is less than the r.h.s.~of \eqref{eq:q4}, and  \textbf{D1} is satisfied.\end{itemize}  
 \textbf{QED}\\

### 2015-06-02 ms.tex
Andrew's minor edits

@@
 \begin{equation*}
-  \sup_{B_1^c} \logRLss \le \logRL(\sshat) - M.
+  \sup_{B^c} \logRLss \le \logRL(\sshat) - M.
 \end{equation*}}
@@
 Let $p$ be an arbitrary point inside $B_1$, as illustrated in Figure~(\ref{fig:boundingbox2}), and define $L \equiv \logRL(p)$; we use $L$ as a lower bound on $\logRL(\sshat)$.
-The proof considers several regions outside $B_1$ in turn, as exemplified by the points $q_1$, $q_2$, $q_3$, and $q_4$ in Figure~\ref{fig:boundingbox2}.
+The proof considers several regions outside $B$ in turn, as exemplified by the points $q_1$, $q_2$, $q_3$, and $q_4$ in Figure~\ref{fig:boundingbox2}.
 Let $\widetilde{q_1} = (0,\widetilde{\sigma_s^2})$ be the intercept of $B$ and the $\sigssq$ axis and let $\widetilde{q_2} = (\sigma_e^{2*},\widetilde{\sigma_s^2})$.
@@
 Let $p$ be an arbitrary point inside $B_1$, as illustrated in Figure~(\ref{fig:boundingbox2}), and define $L \equiv \logRL(p)$; we use $L$ as a lower bound on $\logRL(\sshat)$.
-Let $\widetilde{q_1} = (0,\widetilde{\sigma_s^2})$ be the intercept of $B$ and the $\sigssq$ axis; let $\widetilde{q_2} = (\sigma_e^{2*},\widetilde{\sigma_s^2})$; let $\widetilde{q_3} = (\widetilde{\sigma_e^2}, \sigma_s^{2*})$; let $\widetilde{q_4} = (\widetilde{\sigma_e^2},0)$; and let $\logRL_j$ denote the $j$'th term of \eqref{eq:sum}.
-The proof considers in turn four regions $Q_1$, $Q_2$, $Q_3$, and $Q_4$ whose union is $B^c$ and shows that $\widetilde{\sigma_e^2}$ and $\widetilde{\sigma_s^2}$ can be chosen large enough so that \textbf{D1} is satisfied on each region.
+Let $\widetilde{q_1} = (0,\widetilde{\sigma_s^2})$, the intercept of $B$ and the $\sigssq$ axis; let $\widetilde{q_2} = (\sigma_e^{2*},\widetilde{\sigma_s^2})$; let $\widetilde{q_3} = (\widetilde{\sigma_e^2}, \sigma_s^{2*})$; let $\widetilde{q_4} = (\widetilde{\sigma_e^2},0)$, the intercept of $B$ and the $\sigssq$ axis; and let $\logRL_j$ denote the $j$'th term of \eqref{eq:sum}.
+The proof considers in turn four regions $Q_1$, $Q_2$, $Q_3$, and $Q_4$ whose union is $B^c$.
+Implicitly, $B$ and $B^c$ are functions of $\widetilde{\sigma_e^2}$ and $\widetilde{\sigma_s^2}$.
+The proof shows that $\widetilde{\sigma_e^2}$ and $\widetilde{\sigma_s^2}$ can be chosen large enough so that \textbf{D1} is satisfied on each region.
 \begin{itemize}
@@
   \begin{equation}
-	\logRL(\sshat) - \logRL(q) \ge L - \sum_{j \ne s_z+1} \logRL_j(\widetilde{q_1}) - \logRL_{s_z+1}(\widetilde{q_2}),
-  \end{equation}
-  which is larger than $M$ if
+	\logRL(\sshat) - \logRL(q) \ge L - \sum_{j \ne s_z+1} \logRL_j(\widetilde{q_1}) - \logRL_{s_z+1}(\widetilde{q_2}).
+  \end{equation}
+  The r.h.s.~is larger than $M$ if
   \begin{equation}
@@
   Therefore, $\logRL(\sshat) - \logRL(q) \ge L - \logRL(\widetilde{q_3})$.  The latter expression is greater than
-  $M$ iff $\logRL(\widetilde{q}) < L-M$.  But for any fixed $\sigssq$,
+  $M$ iff $\logRL(\widetilde{q_3}) < L-M$.  But for any fixed $\sigssq$,
   $\lim_{\sigesq \rightarrow \infty} \logRLss = -\infty$ and therefore $\widetilde{\sigma_e^2}$
@@
     \begin{equation*}
-      \frac{\partial\logRL_j}{\partial\sigesq} \le 0 \hspace{1cm} \text{and} \hspace{1cm}
-      \frac{\partial\logRL_j}{\partial\sigssq} < 0
+      \frac{\partial\logRL_j}{\partial\sigesq} < 0 \hspace{1cm} \text{and} \hspace{1cm}
+      \frac{\partial\logRL_j}{\partial\sigssq} \le 0
     \end{equation*}
@@
 The MCMC draws depicted in \cite{henn&hodges:2014} Figure~2c show that the plateau has significant posterior mass.
-%Usual analyses of the posterior density are carried out either by MCMC, in which case one hopes that one's Markov chain happens to sample all regions of high enough posterior density to matter, or by evaluating the posterior on a grid --- in which case one hopes one's grid is fine enough.
-With our analysis there is no need to hope; the algorithm is certain to find all regions of high density.
 
@@
 
-%\begin{figure}
-%  \begin{subfigure}{.5\textwidth}
-%    \centering
-%    \includegraphics[width=.8\linewidth]{figs/gmst-boxes2.jpg}
-%    \caption{Locations of boxes}
-%    \label{fig:gmst-boxes2}
-%  \end{subfigure}
-%  \begin{subfigure}{.5\textwidth}
-%    \centering
-%    \includegraphics[width=.8\linewidth]{figs/gmst-rll2.jpg}
-%    \caption{Grayscale shows $L^b$ in each box.
-%                  Red dot shows $b^* \equiv\argmax L^b$.
-%	          Boxes outlined in blue have $U^b \ge L^{b^*}$.
-%	          Axes are logarithmic.}
-%    \label{fig:gmstrll2}
-%  \end{subfigure}
-%  \caption{The 3,605,095 boxes from the second analysis of global
-%                mean surface temperature.}
-%\label{fig:gmst2}
-%\end{figure}
-
 \section{Discussion}
-This paper explains and illustrates an algorithm for linear mixed models with two variances, to find all regions where either the restricted likelihood function or the joint posterior density of the variances is high, and to evaluate the function there to arbitrary accuracy.
+This paper explains and illustrates an algorithm that facilitates computation for linear mixed models with two variances.
+The algorithm finds all regions where either the restricted likelihood function or the joint posterior density of the variances is high and can evaluate the function there to arbitrary accuracy.
 A natural question to ask is \emph{What about linear mixed models with more than two variances?}  A partial answer is given by \cite{hodges:2013} who shows that some models with more than two variances can be re-expressed similarly to \eqref{eq:reexpress} but others can't.

### 2015-06-03 ms.tex
algorithm and minor edits

@@
 %\usepackage{setspace}
+\usepackage{algorithm}
+\usepackage{algpseudocode}
 
@@
 \newcommand{\argsup}{\operatornamewithlimits{argsup}}
+\newcommand*\DNA{\textsc{dna}}
+\newcommand*\Let[2]{\State #1 $\gets$ #2}
 \renewcommand{\bibname}{References}
@@
 \author[1]{Michael Lavine}
-\author[1,2]{Andrew Bray}
+\author[2]{Andrew Bray}
 \author[3]{James Hodges}
 \affil[1]{Department of Mathematics and Statistics, University of Massachusetts, Amherst, MA 01003 USA}
-\affil[2]{Mount Holyoke College}
-\affil[3]{Division of Biostatistics, University of Minnesota, Minneapolis, MN 55455 USA}
+\affil[2]{Reed College, Portland, OR, 97202 USA}
+\affil[3]{Division of Biostatistics, University of Minnesota, Minneapolis, MN, 55455 USA}
 
@@
 For $j=1, \dots, s_z, s_z+1$, the partial derivatives \eqref{eq:drlds} are negative everywhere between $p_2$ and $\widetilde{p_2}$; for $j=s_z+2$, the partial derivative is zero.
-Therefore, $\logRL(\widetilde{p_2}) \ge \logRL(p_2)$ and there can be no local maxima of $\logRL$ to the right of the line $\sigesq = \sigesq{}^M$.
+Therefore, $\logRL(\widetilde{p_2}) \ge \logRL(p_2)$ and there can be no local maxima of $\logRL$ to the right of the line $\sigesq = \sigesq{}^M$.\\
 \textbf{QED}\\
@@
 \label{sec:algorithm}
-Section~\ref{sec:boundingbox} shows how to determine a box $B_1$ containing all local maxima of $\logRLss$; and it shows that we can find a bigger box $B\supseteq B_1$ satisfying \textbf{D1}.
+Section~\ref{sec:boundingbox} shows how to determine a box $B_1$ containing all local maxima of $\logRLss$ and   that we can find a bigger box $B\supseteq B_1$ satisfying \textbf{D1}.
 We are about to present an algorithm that uses Section~\ref{sec:boundsinboxes} to show that any bounded box $B^0$ can be partitioned into finitely many smaller boxes $B^0_1, \dots B^0_p$  such that, for each $B^0_i$ in the partition, we know for pre-specified constants $M, \epsilon >0$, either
@@
 
-With \textbf{D1} and \textbf{D2} the \textcompute{R} function \textcompute{findf}, sketched below, will evaluate, within a box $B^0$, $\logRL$ to arbitrary accuracy everywhere $\logRL$ is large by performing the following tasks:
-\begin{enumerate*}[label=(\alph*)]
-\item accept as input $\{a_j, b_j, c_j, d_j\}, B^0$ and some tuning constants;
-\item create a list of active boxes, initially consisting of just $B^0$;
-\item create a list of inactive boxes, initially empty; and
-\item for each active box $B_j$, find $(L^{B_j}, U^{B_j})$ and determine whether $B_j$ needs to be subdivided.
-\end{enumerate*}
- These notes explain some parts of the function in more detail.
-\begin{enumerate}
-\item A dataframe \textcompute{lines} is an input to \textcompute{findf}.  \textcompute{lines} contains the
-  $\{a_j, b_j, c_j, d_j \}$ from Section~\ref{subsec:lines}.
-  Details for computing them from $y$, $X$, $Z$, $\Sigma_e$, and $\Sigma_s$ are in the Appendix.
-\item \textcompute{startbox}, or $B^0$, is an input to \textcompute{findf}.  As explained at the beginning of
-  Section~\ref{sec:algorithm}, \textcompute{startbox} could be $B_1$, $B$, or any other box the user chooses.
-\item Constants \maxit, $M$, $\epsilon$, $\delta_e$, and $\delta_s$ are inputs to \textcompute{findf}.
-	\begin{enumerate}[label=(\alph*)]
-	\item \maxit\ is the maximum number (could be $\infty$) of iterations of
-		\textcompute{findf}'s loop.
-	\item We will not further analyze regions of the plane where
-		$\logRLss < \logRL(\sshat) - M$.  ($M$ could be $\infty$.)
-	\item We will evaluate $\logRL$ to within an accuracy of $\epsilon$ (could be 0) inside $B^0$
-	         unless evaluation is stopped by one of the other criteria.
-	\item We will not distinguish values of $\sigesq$ separated by less
-		than $\delta_e$ (could be 0) nor values of $\sigssq$ separated by
-		less than $\delta_s$ (could be 0).  Separation may be specified
-		in either absolute or relative terms. (I.e.\ we look at either the difference
-		in $\sigesq$ or $\log\sigesq$ (or $\sigssq$ or $\log\sigssq$) from one
-		side of the box to the other.)
-	\end{enumerate}
-  Setting $\maxit = \infty$ and $\delta_e = \delta_s = 0$ implies that $B^0$ will be partitioned so that either
-  $\logRL < \logRL(\sshat) - M$ or $\logRL$ is known to within $\epsilon$, for every $B^0_i$ in the partition.
-
-\item A \textcompute{box} $b$ is a list consisting of
-	 \begin{itemize}
-	 \item upper and lower limits on $\sigesq$,
-	 \item upper and lower limits on $\sigssq$,
-	 \item upper and lower bounds on $\logRL$, $U^b$ and $L^b$ from \eqref{eq:bounds},
-	 \item indicators for whether this box lies above (or to the right of), below (or to the left of),
-	   or straddles each of the $s_z+2$ lines.
-	 \end{itemize}
-\item \textcompute{killfunc} is a function, shown below \textcompute{findf}, that determines whether
-	a box $b$ should be divided more finely.
-\item \textcompute{splitbox} is a function that takes a box $b =
-	[\sigesq{}_\text{low}, \sigesq{}_\text{high}] \times [\sigssq{}_\text{low}, \sigssq{}_\text{high}]$ as input 
-	and returns the four boxes created by dividing each side of $b$ at its midpoint.
-%\item With $y$, $X$, $Z$, $\Sigma_e$, and $\Sigma_s$ --- see \eqref{eq:lmm}
-%	and \eqref{eq:rll} for definitions and the Appendix for details --- compute $\{a_j, b_j, c_j, d_j\}$ for $j=1,
-%	\dots, s_z+2$.
-%\item Create two lists of boxes: active and inactive.  Both lists are initially empty.
-%\item Specify a box $B$ as the sole entry of the active list.  $B$ could be,
-%	but doesn't have to be, a box satisfying desideratum \textbf{D1}.
-%\item Either set $L = -\infty$ or evaluate $\logRL(p_i)$ at a few points
-%	$p_1, \dots p_k$ and set $L = \max_i\logRL(p_i)$.  $L$ is our current
-%	lower bound on $\logRL(\sshat)$.
-%\item While the active list is not empty or we have not reached \maxit\
-%	iterations
-%	\begin{enumerate}
-%	\item For each active box $b$:
-%		\begin{enumerate}
-%		\item if $U^b < L-M$ move $b$ to the inactive list.
-%		\item if $U^b - L^b < \epsilon$ move $b$ to the inactive list
-%		\item if the horizontal extent of $b$ is less than $\delta_e$,
-%			move $b$ to the inactive list
-%		\item if the vertical extent of $b$ is less than $\delta_s$,
-%			move $b$ to the inactive list
-%		\item otherwise, $b$ is still active.  Divide $b$ into four subboxes.
-%			Remove $b$ and add the subboxes to the active list.
-%		\end{enumerate}
-%	\item Set $L = \max ( L, \max_{b\in \text{active list}}L^b )$.  $L$ is our new
-%		lower bound on $\logRL(\sshat)$.
-%	\end{enumerate}
-%\item Return the active and inactive lists.
-\end{enumerate}
-\begin{verbatim}
-findf <- function(lines, startbox, eps = 0, delE = 0, delS = 0, M = Inf, maxit = 10,
-                  ratio = FALSE, lognote = "summary") {
-
-    inactive <- list()  # a list of inactive boxes
-    ninact <- 0  # the number of inactive boxes
-
-    active <- list(startbox)  # a list of active boxes
-    nact <- 0  # the number of active boxes
-
-    lowbound <- -Inf  # lowerbound on max(log(f))
-    iter <- 0  # iteration number
-
-    while (nact > 0 && iter < maxit) {
-        # Find the lower bound of each box and the maximum of the lower bounds.  For
-        # each active box, either make it inactive or divide it.
-        low.act <- max(vapply(X = active, FUN = function(box) {
-            box$bounds[1]
-        }, FUN.VALUE = 0.1))
-        lowbound <- max(lowbound, low.act)
-        kill <- vapply(X = active, FUN = killfunc, FUN.VALUE = TRUE, lb = lowbound,
-            M = M, eps = eps, delE - delE, delS = delS, ratio = ratio)
-            # which boxes become inactive?
-        nkill <- sum(kill)
-        if (nkill > 0) {
-            inactive[(ninact + 1):(ninact + nkill)] <- active[kill] 
-            # add those boxes to the inactive list
-        }
-        ninact <- length(inactive)
-        kids <- list()  # subdivisions of active boxes
-        nkids <- 0
-        for (i in which(!kill)) {  # split active boxes into 4 parts
-            kids[(nkids + 1):(nkids + 4)] <- splitbox(active[[i]], lines)
-            nkids <- nkids + 4
-        }
-        active <- kids
-        nact <- length(active)
-
-        iter <- iter + 1
-        write(c("iteration", iter, "nact", nact, "ninact", ninact, "lowbound",
-            lowbound), file = "log.out", ncolumns = 8, append = TRUE)
-    }
-
-    tmp <- t(vapply(X = c(active, inactive), FUN = function(box) {
-        c(box$lims.sigsqs, box$lims.sigsqe, box$bounds)
-    }, FUN.VALUE = c(sigsqs.lo = 0.1, sigsqs.hi = 0.1, sigsqe.lo = 0.1, sigsqe.hi = 0.1,
-        rll.lower = 0.1, rll.upper = 0.1)))
-    end_time <- Sys.time()
-    write(paste(lognote, "runtime: ", round(end_time - start_time, digits = 3)),
-          file = "log.out", ncolumns = 1, append = TRUE)
-    return(data.frame(tmp))
-}
-
-# conditions under which a box becomes inactive
-killfunc <- function(box, lb, M, eps, delE, delS, ratio) {
-    # lb is a global (within startbox) lower bound on max(log(f));
-    #      it changes at each iteration.
-    #  M, eps, delE, delS stay constant throughout the iterations.
-    cond.low <- box$bounds[2] < lb - M
-    cond.eps <- diff(box$bounds) < eps
-    cond.E <- ifelse(ratio, diff(log(box$lims.sigsqe)) < delE,
-                            diff(box$lims.sigsqe) < delE)
-    cond.S <- ifelse(ratio, diff(log(box$lims.sigsqs)) < delS,
-                            diff(box$lims.sigsqs) < delS)
-    return(cond.low || cond.eps || cond.E || cond.S)  # stop dividing this box?
-}
-\end{verbatim}
-
+\begin{algorithm}
+  \caption{Learn areas in $B$ where log $f$ is high to within $\epsilon$
+    \label{alg:findf}}
+  \begin{algorithmic}[1]
+    \Function{findf}{$B, \epsilon, M$}
+      \Let{$active$}{list containing only $B$}
+      \Let{$inactive$}{empty list}
+      \Let{$L$}{$-\infty$}
+      \While{length of $active > 0$}
+        \For{each element $b$ in $active$}
+          \Let{$L^b, U^b$}{get bounds on log $f$ in $b$ \Comment{see section \ref{sec:boundsinboxes}}}
+        \EndFor
+          \Let{tmp}{max($L^b$)}
+         \Let{$L$}{max($L$, tmp)}
+        \For{each element $b$ in $active$}
+          \If{$U^b - L^b < \epsilon \enspace \textrm{or} \enspace L^b < L - M$}
+            \State move $b$ from $active$ to $inactive$
+          \Else
+            \Let{$b_1, b_2, b_3, b_4$}{split $b$ into 4 smaller boxes}
+            \State remove $b$ from $active$
+            \State add $b_1, b_2, b_3, b_4$ to $active$
+          \EndIf
+        \EndFor
+      \EndWhile
+      \State \Return $inactive$ 
+    \EndFunction
+  \end{algorithmic}
+\end{algorithm}
+
+Algorithm~\ref{alg:findf} sketches the basic procedure; details are in an appendix.
 Before giving examples, we mention some considerations in setting the tuning constants $M$ and $\epsilon$, which have to do with the interpretation of likelihood ratios.
@@
 \label{sec:examp}
-Throughout Section~\ref{sec:examp} we use $\epsilon = 1$, $M = 7$, $\delta_e = \delta_s = 0$ and $\maxit = 25$.
+Section~\ref{sec:examp}'s examples use $\epsilon = 1$ and $M = 7$.
+%, $\delta_e = \delta_s = 0$ and $\maxit = 25$.
 We used a finite value of $\maxit$ for practicality but, because each example took 25 or fewer iterations to adequately resolve $\log f$, $\maxit = 25$ yields the same results as $\maxit = \infty$.
@@
 For a $\logRLssorig$ analysis there are 43 lines, as shown in Figure~\ref{fig:hmolines}.
-Running the algorithm on the box $B_1$ determined by the lines' maximum intercepts on the $\sigesq$ and $\sigssq$ axes with the settings
-\begin{equation*}
-	\maxit=\infty; \hspace{1cm} \epsilon=1; \hspace{1cm}
-	\delta_e=0; \hspace{1cm} \delta_s=0; \hspace{1cm} M=7
-\end{equation*}
+Running the algorithm on the box $B_1$ determined by the lines' maximum intercepts on the $\sigesq$ and $\sigssq$ axes
+%with the settings
+%\begin{equation*}
+%	\maxit=\infty; \hspace{1cm} \epsilon=1; \hspace{1cm}
+%	\delta_e=0; \hspace{1cm} \delta_s=0; \hspace{1cm} M=7
+%\end{equation*}
 results in the the output displayed in Table~\ref{table:hmo_HH11_run}, which shows the state of the algorithm after iterations 1 through 15: the numbers of active and inactive boxes and the current value of the lower bound $L$ on $\max\log f$.
@@
 \end{figure}
-Because the $\logRLorig$ analysis in Figure~\ref{fig:hmorll} shows low values for $\sigesq, \sigssq > 1000$, we run the Bayesian analysis on the box $B^0 = (0,1000) \times (0,1000)$, again with control parameters
-\begin{equation*}
-	\maxit=\infty; \hspace{1cm} \epsilon=1; \hspace{1cm}
-	\delta_e=0; \hspace{1cm} \delta_s=0; \hspace{1cm} M=7.
-\end{equation*}
+Because the $\logRLorig$ analysis in Figure~\ref{fig:hmorll} shows low values for $\sigesq, \sigssq > 1000$, we run the Bayesian analysis on the box $B^0 = (0,1000) \times (0,1000)$.
+%, again with control parameters
+%\begin{equation*}
+%	\maxit=\infty; \hspace{1cm} \epsilon=1; \hspace{1cm}
+%	\delta_e=0; \hspace{1cm} \delta_s=0; \hspace{1cm} M=7.
+%\end{equation*}
 The algorithm finished in 22 iterations and took a little under 3 hours.  Table~\ref{table:hmo_HH11Bayes} shows the output and Figure~\ref{fig:hmoBayes} shows the boxes.
@@
 \end{figure}
-As shown in \cite{ruppert_etal:2003}, many splines can be written as as linear mixed models.
+As shown in \cite{ruppert_etal:2003}, many splines can be written as linear mixed models.
 \cite{hodges:2013} fit a piecewise quadratic spline to the GMST data, though a piecewise cubic spline would look similar.
@@
 \end{figure}
-As usual, we use $B^0 = B_1$, the box determined by the largest intercepts of the 31 lines on the $\sigesq$ and $\sigssq$ axes and the control constants 
-\begin{equation*}
-	\maxit=\infty; \hspace{1cm} \epsilon=1; \hspace{1cm}
-	\delta_e= \delta_s=0; \hspace{1cm} M=7.
-\end{equation*}
+As usual, we use $B^0 = B_1$, the box determined by the largest intercepts of the 31 lines on the $\sigesq$ and $\sigssq$ axes.
+%and the control constants 
+%\begin{equation*}
+%	\maxit=\infty; \hspace{1cm} \epsilon=1; \hspace{1cm}
+%	\delta_e= \delta_s=0; \hspace{1cm} M=7.
+%\end{equation*}
 After 25 iterations and about 35 minutes of computing time, all boxes became inactive.  Output is in Table~\ref{table:gmst_output}; boxes are displayed in Figure~\ref{fig:gmst1}.
@@
 As written, our algorithm moves a box $b$ to the inactive list if
-\begin{enumerate}[label=(\alph*)]
+\begin{enumerate*}[label=(\alph*)]
 \item $U^b < L-M$ or
-\item $U^b - L^b < \epsilon$ or
-\item either the vertical or horizontal extent of $b$ is sufficiently small.
-\end{enumerate}
+\item $U^b - L^b < \epsilon$.
+% \item either the vertical or horizontal extent of $b$ is sufficiently small.
+\end{enumerate*}
 But one could construct more elaborate rules.
-One appealing example is to apply criteria (a) and (c) if $U^b \le L-\epsilon_2$ and apply criterion (b) if $U^b > L-\epsilon_2$.
+One appealing example is to apply criterion (a) if $U^b \le L-\epsilon_2$ and apply criterion (b) if $U^b > L-\epsilon_2$.
 Other rules are possible, too.
@@
 
-\section{Appendix}
-\paragraph{Derivation of $\{a_j\}$ and $\{\hat v_j\}$ in \eqref{eq:reexpress}, \eqref{eq:logpost}, and \eqref{eq:sum}}  Our derivation follows \cite{hodges:2013},
+\appendix
+\section{Appendix: Derivation of $\{a_j\}$ and $\{\hat v_j\}$ in \eqref{eq:reexpress}, \eqref{eq:logpost}, and \eqref{eq:sum}}
+Our derivation follows \cite{hodges:2013},
 which contains more details.  There are three steps.
@@
 
+\section{Appendix: Details of the algorithm}
+With \textbf{D1} and \textbf{D2} the \textcompute{R} function \textcompute{findf}, sketched below, will evaluate, within a box $B^0$, $\logRL$ to arbitrary accuracy everywhere $\logRL$ is large by performing the following tasks:
+\begin{enumerate*}[label=(\alph*)]
+\item accept as input $\{a_j, b_j, c_j, d_j\}, B^0$ and some tuning constants;
+\item create a list of active boxes, initially consisting of just $B^0$;
+\item create a list of inactive boxes, initially empty; and
+\item for each active box $B_j$, find $(L^{B_j}, U^{B_j})$ and determine whether $B_j$ needs to be subdivided.
+\end{enumerate*}
+ These notes explain some parts of the function in more detail.
+\begin{enumerate}
+\item A dataframe \textcompute{lines} is an input to \textcompute{findf}.  \textcompute{lines} contains the
+  $\{a_j, b_j, c_j, d_j \}$ from Section~\ref{subsec:lines}.
+  Details for computing them from $y$, $X$, $Z$, $\Sigma_e$, and $\Sigma_s$ are in the Appendix.
+\item \textcompute{startbox}, or $B^0$, is an input to \textcompute{findf}.  As explained at the beginning of
+  Section~\ref{sec:algorithm}, \textcompute{startbox} could be $B_1$, $B$, or any other box the user chooses.
+\item Constants \maxit, $M$, $\epsilon$, $\delta_e$, and $\delta_s$ are inputs to \textcompute{findf}.
+	\begin{enumerate}[label=(\alph*)]
+	\item Though not mentioned in the main text, \maxit\ is the maximum number (could be $\infty$) of iterations of
+		\textcompute{findf}'s loop.
+	\item We will not further analyze regions of the plane where
+		$\logRLss < \logRL(\sshat) - M$.  ($M$ could be $\infty$.)
+	\item We will evaluate $\logRL$ to within an accuracy of $\epsilon$ (could be 0) inside $B^0$
+	         unless evaluation is stopped by one of the other criteria.
+	\item Though not mentioned in the main text, the algorithm can be told not
+		to distinguish values of $\sigesq$ separated by less
+		than $\delta_e$ (could be 0) nor values of $\sigssq$ separated by
+		less than $\delta_s$ (could be 0).  Separation may be specified
+		in either absolute or relative terms. (I.e.\ we look at either the difference
+		in $\sigesq$ or $\log\sigesq$ (or $\sigssq$ or $\log\sigssq$) from one
+		side of the box to the other.)
+	\end{enumerate}
+  Setting $\maxit = \infty$ and $\delta_e = \delta_s = 0$, as we have done in the examples, implies
+  that $B^0$ will be partitioned so that either
+  $\logRL < \logRL(\sshat) - M$ or $\logRL$ is known to within $\epsilon$, for every $B^0_i$ in the partition.
+
+\item A \textcompute{box} $b$ is a list consisting of
+	 \begin{itemize}
+	 \item upper and lower limits on $\sigesq$,
+	 \item upper and lower limits on $\sigssq$,
+	 \item upper and lower bounds on $\logRL$, $U^b$ and $L^b$ from \eqref{eq:bounds},
+	 \item indicators for whether this box lies above (or to the right of), below (or to the left of),
+	   or straddles each of the $s_z+2$ lines.
+	 \end{itemize}
+\item \textcompute{killfunc} is a function, shown below \textcompute{findf}, that determines whether
+	a box $b$ should be divided more finely.
+\item \textcompute{splitbox} is a function that takes a box $b =
+	[\sigesq{}_\text{low}, \sigesq{}_\text{high}] \times [\sigssq{}_\text{low}, \sigssq{}_\text{high}]$ as input 
+	and returns the four boxes created by dividing each side of $b$ at its midpoint.
+\end{enumerate}
+\begin{verbatim}
+findf <- function(lines, startbox, eps = 0, delE = 0, delS = 0, M = Inf, maxit = 10,
+                  ratio = FALSE, lognote = "summary") {
+
+    inactive <- list()  # a list of inactive boxes
+    ninact <- 0  # the number of inactive boxes
+
+    active <- list(startbox)  # a list of active boxes
+    nact <- 0  # the number of active boxes
+
+    lowbound <- -Inf  # lowerbound on max(log(f))
+    iter <- 0  # iteration number
+
+    while (nact > 0 && iter < maxit) {
+        # Find the lower bound of each box and the maximum of the lower bounds.  For
+        # each active box, either make it inactive or divide it.
+        low.act <- max(vapply(X = active, FUN = function(box) {
+            box$bounds[1]
+        }, FUN.VALUE = 0.1))
+        lowbound <- max(lowbound, low.act)
+        kill <- vapply(X = active, FUN = killfunc, FUN.VALUE = TRUE, lb = lowbound,
+            M = M, eps = eps, delE - delE, delS = delS, ratio = ratio)
+            # which boxes become inactive?
+        nkill <- sum(kill)
+        if (nkill > 0) {
+            inactive[(ninact + 1):(ninact + nkill)] <- active[kill] 
+            # add those boxes to the inactive list
+        }
+        ninact <- length(inactive)
+        kids <- list()  # subdivisions of active boxes
+        nkids <- 0
+        for (i in which(!kill)) {  # split active boxes into 4 parts
+            kids[(nkids + 1):(nkids + 4)] <- splitbox(active[[i]], lines)
+            nkids <- nkids + 4
+        }
+        active <- kids
+        nact <- length(active)
+
+        iter <- iter + 1
+        write(c("iteration", iter, "nact", nact, "ninact", ninact, "lowbound",
+            lowbound), file = "log.out", ncolumns = 8, append = TRUE)
+    }
+
+    tmp <- t(vapply(X = c(active, inactive), FUN = function(box) {
+        c(box$lims.sigsqs, box$lims.sigsqe, box$bounds)
+    }, FUN.VALUE = c(sigsqs.lo = 0.1, sigsqs.hi = 0.1, sigsqe.lo = 0.1, sigsqe.hi = 0.1,
+        rll.lower = 0.1, rll.upper = 0.1)))
+    end_time <- Sys.time()
+    write(paste(lognote, "runtime: ", round(end_time - start_time, digits = 3)),
+          file = "log.out", ncolumns = 1, append = TRUE)
+    return(data.frame(tmp))
+}
+
+# conditions under which a box becomes inactive
+killfunc <- function(box, lb, M, eps, delE, delS, ratio) {
+    # lb is a global (within startbox) lower bound on max(log(f));
+    #      it changes at each iteration.
+    #  M, eps, delE, delS stay constant throughout the iterations.
+    cond.low <- box$bounds[2] < lb - M
+    cond.eps <- diff(box$bounds) < eps
+    cond.E <- ifelse(ratio, diff(log(box$lims.sigsqe)) < delE,
+                            diff(box$lims.sigsqe) < delE)
+    cond.S <- ifelse(ratio, diff(log(box$lims.sigsqs)) < delS,
+                            diff(box$lims.sigsqs) < delS)
+    return(cond.low || cond.eps || cond.E || cond.S)  # stop dividing this box?
+}
+\end{verbatim}
+
 \bibliographystyle{jasa}

### 2015-06-03 ms.tex
minor

@@
 By the claim, \mrle\ must lie on or inside $B_1$, so $B_1$ could be passed to an optimizer such as \R's \textsf{optim} or \textsf{nlminb} with potentially better results than using those functions without bounds.
-However, even with known bounds, general purpose optimizers may still miss \mrle\ and regions of high $\logRL$ and they emphasize single points of highest local $\logRL$ while possibly ignoring large adjacent regions where $\logRL$ is nearly as large \citep{hill:1965, hodges:2013}.
+However, even with known bounds, general purpose optimizers may still miss \mrle\ and
+%regions of high $\logRL$ and
+they emphasize single points of highest local $\logRL$ while possibly ignoring large
+%adjacent
+regions where $\logRL$ is nearly as
+high %large
+\citep{hill:1965, hodges:2013}.
 
@@
         \For{each element $b$ in $active$}
-          \If{$U^b - L^b < \epsilon \enspace \textrm{or} \enspace L^b < L - M$}
+          \If{$U^b - L^b < \epsilon \enspace \textrm{or} \enspace U^b < L - M$}
             \State move $b$ from $active$ to $inactive$
@@
 Figure~\ref{fig:hmorll} depicts the same $\logRLorig$ as \cite{henn&hodges:2014}'s Figures~2a (MCMC draws) and~2b ($\logRLorig$ contours), but their Figure~2a was produced by MCMC whereas our Figure~\ref{fig:hmorll} was produced by direct calculation.
-Their Figure~2a shows that the MCMC sampler did not sample any values of $\sigssq$ less than about 10, whereas our Figure~\ref{fig:hmorll} and their Figure~2b shows a region of high $\logRLorig$ extending down to $\sigssq=0$.
+Their Figure~2a shows that the MCMC sampler did not sample any values of $\sigssq$ less than about 10, whereas our Figure~\ref{fig:hmorll} and their Figure~2b show a region of high $\logRLorig$ extending down to $\sigssq=0$.
 In fact, $\logRLorig(500,0) \approx -1241.5$, only about 6 log units below $\logRLorig(\sshat) \approx -1235.7$.

### 2015-06-06 ms.tex
Jim's edits
use alternate proof of claim

@@
 \usepackage{algpseudocode}
+\usepackage{hyperref}
 
@@
 Figure~(\ref{fig:bigboundingbox}) shows $B_1$ and five lines --- realistic data sets may have more --- labelled $j=1, 2, 3, s_z+1, s_z+2$.\\[5pt]
-\textbf{Claim} \emph{All local maxima of $\logRL$ lie within $B_1$.}\\
+\textbf{Claim} \emph{All local maxima of $\logRL$ lie within $B_1$.}\\[5pt]
 \textbf{Proof} Let $p_1 = (\sigesq{}_1, \sigssq{}_1)$ be a point such that $\sigssq{}_1 > \sigssq{}^M$ and let $\widetilde{p_1} = (\sigesq{}_1, \sigssq{}^M)$, as illustrated in Figure~(\ref{fig:bigboundingbox}).
@@
 \end{equation*}}
-I.e., \textbf{D1} is satisfied.\\
-\noindent\textbf{Proof}
-Let $p$ be an arbitrary point inside $B_1$, as illustrated in Figure~(\ref{fig:boundingbox2}), and define $L \equiv \logRL(p)$; we use $L$ as a lower bound on $\logRL(\sshat)$.
+I.e., \textbf{D1} is satisfied.\\[5pt]
+%\noindent\textbf{Proof}
+%Let $p$ be an arbitrary point inside $B_1$, as illustrated in Figure~(\ref{fig:boundingbox2}), and define $L \equiv \logRL(p)$; we use $L$ as a lower bound on $\logRL(\sshat)$.
 The proof considers several regions outside $B$ in turn, as exemplified by the points $q_1$, $q_2$, $q_3$, and $q_4$ in Figure~\ref{fig:boundingbox2}.
@@
 Let $\logRL_j(\cdot)$ denote the $j$'th term of \eqref{eq:sum} evaluated at $(\cdot)$.
-\begin{equation}
-	\logRL(\sshat) - \logRL(q_1) \ge L - \sum_{j \ne s_z+1} \logRL_j(\widetilde{q_1}) - \logRL_{s_z+1}(\widetilde{q_2}),
-\end{equation}
-which is larger than $M$ if
-\begin{equation}
-\label{eq:q1}
-  \begin{split}
-    \sum_{j \ne s_z+1} \logRL_j(\widetilde{q_1})
-      &< L - M - \logRL_{s_z+1}(\widetilde{q_2})\\
-      &= L - M + \frac{1}{2} \left[ c_{s_z+1}\log\sigma_e^{2*} + \frac{d_{s_z+1}}{\sigma_e^{2*}}\right].
-  \end{split}
-\end{equation}
-Examination of (\ref{eq:sum}) shows that for any fixed $\sigesq$, in particular for $\sigesq=0$, and for $j \ne s_z+1$,\\ $\lim_{\sigssq \rightarrow \infty} \logRL_j(\sigesq,\sigssq) = -\infty$.
+%\begin{equation}
+%	\logRL(\sshat) - \logRL(q_1) \ge L - \sum_{j \ne s_z+1} \logRL_j(\widetilde{q_1}) - \logRL_{s_z+1}(\widetilde{q_2}),
+%\end{equation}
+%which is larger than $M$ if
+%\begin{equation}
+%\label{eq:q1}
+%  \begin{split}
+%    \sum_{j \ne s_z+1} \logRL_j(\widetilde{q_1})
+%      &< L - M - \logRL_{s_z+1}(\widetilde{q_2})\\
+%      &= L - M + \frac{1}{2} \left[ c_{s_z+1}\log\sigma_e^{2*} + \frac{d_{s_z+1}}{\sigma_e^{2*}}\right].
+%  \end{split}
+%\end{equation}
+%Examination of (\ref{eq:sum}) shows that for any fixed $\sigesq$, in particular for $\sigesq=0$, and for $j \ne s_z+1$,\\ $\lim_{\sigssq \rightarrow \infty} \logRL_j(\sigesq,\sigssq) = -\infty$.
 Thus $\widetilde{\sigma_s^2}$ can be chosen large enough so that the summation on the l.h.s.~of \eqref{eq:q1} is less than the r.h.s.~of \eqref{eq:q1}, and  \textbf{D1} is satisfied.
-
-Let $q_2 = (\sigesq{}_2, \sigssq{}_2)$ satisfy $\sigesq{}_2 \ge \sigma_e^{2*}$ and $\sigssq{}_2 \ge \widetilde{\sigma_s^2}$ --- above $B$ and to the right of the $s_z+1$ line.
+%
+%Let $q_2 = (\sigesq{}_2, \sigssq{}_2)$ satisfy $\sigesq{}_2 \ge \sigma_e^{2*}$ and $\sigssq{}_2 \ge \widetilde{\sigma_s^2}$ --- above $B$ and to the right of the $s_z+1$ line.
 (Points $q_2$ and $q_3$ in Figure~(\ref{fig:boundingbox2}) both satisfy these criteria.)  The partial derivatives (\ref{eq:drlds}, \ref{eq:drlde}) evaluated at $q_2$ are nonpositive for $j \in \{1, \dots, s_z+2\}$; therefore, $\logRL(\sshat) - \logRL(q_2) \ge L - \logRL(\widetilde{q_2})$.
@@
 But examination of (\ref{eq:sum}) shows that for any fixed $\sigesq$, $\lim_{\sigssq \rightarrow \infty} \logRLss = -\infty$ and therefore $\widetilde{\sigma_s^2}$ can be chosen large enough so that \textbf{D1} is satisfied.
-
-Let $\widetilde{q_3} = (\widetilde{\sigma_e^2}, \sigma_s^{2*})$ and let $q_3 = (\sigesq{}_3, \sigssq{}_3)$ satisfy $\sigesq{}_3 \ge \widetilde{\sigma_e^2}$ and $\sigssq{}_3 \ge \sigma_s^{2*}$.
+%
+%Let $\widetilde{q_3} = (\widetilde{\sigma_e^2}, \sigma_s^{2*})$ and let $q_3 = (\sigesq{}_3, \sigssq{}_3)$ satisfy $\sigesq{}_3 \ge \widetilde{\sigma_e^2}$ and $\sigssq{}_3 \ge \sigma_s^{2*}$.
 The partial derivatives (\ref{eq:drlds}, \ref{eq:drlde}) evaluated at $q_3$ are nonpositive for $j \in \{1, \dots, s_{z+2}\}$; therefore, $\logRL(\sshat) - \logRL(q_3) \ge L - \logRL(\widetilde{q_3})$.
@@
 But examination of (\ref{eq:sum}) shows that for any fixed $\sigssq$, $\lim_{\sigesq \rightarrow \infty} \logRLss = -\infty$ and therefore $\widetilde{\sigma_e^2}$ can be chosen large enough so that \textbf{D1} is satisfied.
-
-Let $\widetilde{q_4} = (\widetilde{\sigma_e^2},0)$ and let $q_4 = (\sigesq{}_4, \sigssq{}_4)$ satisfy $\sigssq{}_4 \le \sigma_s^{2*}$ and $\sigesq{}_4 \ge \widetilde{q_4}$.
+%
+%Let $\widetilde{q_4} = (\widetilde{\sigma_e^2},0)$ and let $q_4 = (\sigesq{}_4, \sigssq{}_4)$ satisfy $\sigssq{}_4 \le \sigma_s^{2*}$ and $\sigesq{}_4 \ge \widetilde{q_4}$.
 The partial derivatives \eqref{eq:drlds} evaluated at $q_4$ are nonpositive for $j \in \{1, \dots, s_z+1\}$  but nonnegative for $j=s_z+2$.
@@
 Therefore,
-\begin{equation*}
-	\logRL(\sshat) - \logRL(q_4) \ge L - \sum_{j \ne s_z+2} \logRL_j(\widetilde{q_4}) - \logRL_{s_z+2}(\widetilde{q_3})
-\end{equation*}
-which is larger than $M$ if
-\begin{equation}
-\label{eq:q4}
-  \begin{split}
-    \sum_{j \ne s_z+2} \logRL_j(\widetilde{q_4})
-      &< L - M - \logRL_{s_z+2}(\widetilde{q_3})\\
-      &= L - M + \frac{1}{2} \left[ c_{s_z+2}\log\sigma_s^{2*} + \frac{d_{s_z+2}}{\sigma_s^{2*}}\right].
-  \end{split}
-\end{equation}
-Examination of (\ref{eq:sum}) shows that for any fixed $\sigssq$, in particular for $\sigssq=0$, and for $j \ne s_z+2$,\\ $\lim_{\sigesq \rightarrow \infty} \logRL_j(\sigesq,\sigssq) = -\infty$.
+%\begin{equation*}
+%	\logRL(\sshat) - \logRL(q_4) \ge L - \sum_{j \ne s_z+2} \logRL_j(\widetilde{q_4}) - \logRL_{s_z+2}(\widetilde{q_3})
+%\end{equation*}
+%which is larger than $M$ if
+%\begin{equation}
+%\label{eq:q4}
+%  \begin{split}
+%    \sum_{j \ne s_z+2} \logRL_j(\widetilde{q_4})
+%      &< L - M - \logRL_{s_z+2}(\widetilde{q_3})\\
+%      &= L - M + \frac{1}{2} \left[ c_{s_z+2}\log\sigma_s^{2*} + \frac{d_{s_z+2}}{\sigma_s^{2*}}\right].
+%  \end{split}
+%\end{equation}
+%Examination of (\ref{eq:sum}) shows that for any fixed $\sigssq$, in particular for $\sigssq=0$, and for $j \ne s_z+2$,\\ $\lim_{\sigesq \rightarrow \infty} \logRL_j(\sigesq,\sigssq) = -\infty$.
 Thus $\widetilde{\sigma_e^2}$ can be chosen large enough so that the summation on the l.h.s.~of \eqref{eq:q4} is less than the r.h.s.~of \eqref{eq:q4}, and  \textbf{D1} is satisfied.
-
-\noindent\textbf{Alt Proof}
+%
+\noindent\textbf{Proof}
 Let $p$ be an arbitrary point inside $B_1$, as illustrated in Figure~(\ref{fig:boundingbox2}), and define $L \equiv \logRL(p)$; we use $L$ as a lower bound on $\logRL(\sshat)$.
@@
 I.e., $X$ is a $341 \times 3$ matrix with columns for the intercept and two fixed effects and $Z$ is a $341 \times 45$ matrix whose columns are indicators of the 45 jurisdictions.
+The data are available at \url{http://www.biostat.umn.edu/~hodges/RPLMBook/Datasets/09_HMO_premiums/Ex9.html}.
 Because the span of $Z$'s columns contains the span of $X$'s, $s_z = 42$.
@@
 $Z$ is $125 \times 30$: one row for each year; one column for each knot.
-\begin{equation*}
-Z =	\begin{bmatrix}
-		0 & 0 & \dots & 0\\
-		0 & 0 & \dots & 0\\
-		0 & 0 & \dots & 0\\
-		0 & 0 & \dots & 0\\
-		1 & 0 & \dots & 0\\
-		4 & 0 & \dots & 0\\
-		9 & 0 & \dots & 0\\
-		16 & 0 & \dots & 0\\
-		25 & 1 & \dots & 0\\
-		\vdots & \vdots & \vdots & \vdots\\
-		13924 & 12996 & \dots & 4\\
-		14161 & 13225 & \dots & 9\\
-		14400 & 13456 & \dots & 16\\
-		14641 & 13689 & \dots & 25\\
-	\end{bmatrix}; \hspace{.5cm} \Sigma_e = \mathbf{1}_{125}; \hspace{.5cm}
-		\Sigma_s = \mathbf{1}_{30}
-\end{equation*}
+%\begin{equation*}
+%Z =	\begin{bmatrix}
+%		0 & 0 & \dots & 0\\
+%		0 & 0 & \dots & 0\\
+%		0 & 0 & \dots & 0\\
+%		0 & 0 & \dots & 0\\
+%		1 & 0 & \dots & 0\\
+%		4 & 0 & \dots & 0\\
+%		9 & 0 & \dots & 0\\
+%		16 & 0 & \dots & 0\\
+%		25 & 1 & \dots & 0\\
+%		\vdots & \vdots & \vdots & \vdots\\
+%		13924 & 12996 & \dots & 4\\
+%		14161 & 13225 & \dots & 9\\
+%		14400 & 13456 & \dots & 16\\
+%		14641 & 13689 & \dots & 25\\
+%	\end{bmatrix}; \hspace{.5cm} \Sigma_e = \mathbf{1}_{125}; \hspace{.5cm}
+%		\Sigma_s = \mathbf{1}_{30}
+%\end{equation*}
 Because we fit a quadratic spline, the entries in $Z$ are squares.

### 2015-06-06 ms.tex
hodges' edits + remove comments

@@
 By the claim, \mrle\ must lie on or inside $B_1$, so $B_1$ could be passed to an optimizer such as \R's \textsf{optim} or \textsf{nlminb} with potentially better results than using those functions without bounds.
-However, even with known bounds, general purpose optimizers may still miss \mrle\ and
-%regions of high $\logRL$ and
-they emphasize single points of highest local $\logRL$ while possibly ignoring large
-%adjacent
-regions where $\logRL$ is nearly as
-high %large
-\citep{hill:1965, hodges:2013}.
+However, even with known bounds, general purpose optimizers may still miss \mrle\ and they emphasize single points of highest local $\logRL$ while possibly ignoring large regions where $\logRL$ is nearly as high \citep[][Section 18.1.1]{hill:1965, hodges:2013}.
 
@@
 Section~\ref{sec:examp}'s examples use $\epsilon = 1$ and $M = 7$.
-%, $\delta_e = \delta_s = 0$ and $\maxit = 25$.
-We used a finite value of $\maxit$ for practicality but, because each example took 25 or fewer iterations to adequately resolve $\log f$, $\maxit = 25$ yields the same results as $\maxit = \infty$.
 Computations were performed on an iMac that was new in 2011, having a 2.7 GHz Intel Core i5 processor and 4GB of memory.
@@
 \subsubsection{Introduction to the Data}
-Our first example is a traditional linear mixed model previously analyzed in \cite{hodges:98}, \cite{hodges:2013}, and \cite{henn&hodges:2014} who reported a bimodal log posterior density for $(\sigesq,\sigssq)$.
+Our first example is a traditional linear mixed model previously analyzed in \citet{hodges:98},  \cite{wakefield:1998}, \cite{hodges:2013}, and \cite{henn&hodges:2014}.
+\cite{wakefield:1998} reported a bimodal log posterior density for $(\sigesq,\sigssq)$.
 Quoting from \cite{henn&hodges:2014},
@@
 For a $\logRLssorig$ analysis there are 43 lines, as shown in Figure~\ref{fig:hmolines}.
-Running the algorithm on the box $B_1$ determined by the lines' maximum intercepts on the $\sigesq$ and $\sigssq$ axes
-%with the settings
-%\begin{equation*}
-%	\maxit=\infty; \hspace{1cm} \epsilon=1; \hspace{1cm}
-%	\delta_e=0; \hspace{1cm} \delta_s=0; \hspace{1cm} M=7
-%\end{equation*}
-results in the the output displayed in Table~\ref{table:hmo_HH11_run}, which shows the state of the algorithm after iterations 1 through 15: the numbers of active and inactive boxes and the current value of the lower bound $L$ on $\max\log f$.
+Running the algorithm on the box $B_1$ determined by the lines' maximum intercepts on the $\sigesq$ and $\sigssq$ axes results in the the output displayed in Table~\ref{table:hmo_HH11_run}, which shows the state of the algorithm after iterations 1 through 15: the numbers of active and inactive boxes and the current value of the lower bound $L$ on $\max\log f$.
 The run finished in less than 10 seconds.
@@
 Boxes with $U^b \ge L \equiv \max L^b$ are outlined in blue; $(\sshat)$ must lie within the blue region.
-%Figure~\ref{fig:hmorll}  suggests that we can restrict attention to a box determined by $\sigesq\in(300,1000)$ and $\sigssq\in(0,1000)$.
-So we rerun the algorithm on that box and with more stringent control parameters:
-%\begin{equation*}
-%	\maxit=15; \hspace{1cm} \epsilon=1; \hspace{1cm}
-%	\delta_e=0; \hspace{1cm} \delta_s=0; \hspace{1cm} M=10
-%\end{equation*}
-%Table~\ref{table:hmo_HH11_run2} shows the output.  The algorithm needed 12 iterations to move all boxes to the inactive list.  The resulting 37,516 boxes are shown in Figure~\ref{fig:hmo2}.
 For comparison, the standard REML analysis using \R's \textcompute{lme} function yields $\sshat \approx (495, 99)$ with 95\% confidence intervals of $(421, 582)$ and $(39, 248)$.
@@
 In contrast, our algorithm guarantees there are no undiscovered points where $\logRLorig$ is more than $\epsilon$ above $L$.
-
-%The HMO $\logRLorig$ analysis illustrates a typical workflow: start with a crudely determined box and coarse settings of the control parameters, then refine the box and the settings as suggested by the output of the first run.
-Repeat as needed.
 
@@
 Because the $\logRLorig$ analysis in Figure~\ref{fig:hmorll} shows low values for $\sigesq, \sigssq > 1000$, we run the Bayesian analysis on the box $B^0 = (0,1000) \times (0,1000)$.
-%, again with control parameters
-%\begin{equation*}
-%	\maxit=\infty; \hspace{1cm} \epsilon=1; \hspace{1cm}
-%	\delta_e=0; \hspace{1cm} \delta_s=0; \hspace{1cm} M=7.
-%\end{equation*}
 The algorithm finished in 22 iterations and took a little under 3 hours.  Table~\ref{table:hmo_HH11Bayes} shows the output and Figure~\ref{fig:hmoBayes} shows the boxes.
@@
     such that the columns of $\left[\Gamma_X | \Gamma_Z \right]$ are an orthonormal basis for the column space
-    of $\left[X|Z\right]$.  Let $\Gamma_c$ be an $n \times n-s_X-s_Z$ matrix such that the columns of
+    of $\left[X|Z\right]$.  Let $\Gamma_c$ be an $n \times (n-s_X-s_Z)$ matrix such that the columns of
     $\left[\Gamma_X | \Gamma_Z | \Gamma_c\right]$ are an orthonormal basis for $\mathbb{R}^n$.  Define the matrix
@@
       \begin{split}
-        y &= \begin{bmatrix} X Z \end{bmatrix} \begin{bmatrix} \beta \\ u \end{bmatrix} + \epsilon\\
-        &= \begin{bmatrix} \Gamma_X  \Gamma_Z \end{bmatrix} M \begin{bmatrix} \beta \\ u \end{bmatrix} + \epsilon\\
-        &= \begin{bmatrix} \Gamma_X \Gamma_ZP \end{bmatrix} \begin{bmatrix} \beta^* \\ v \end{bmatrix} + \epsilon
+        y &= \begin{bmatrix} X | Z \end{bmatrix} \begin{bmatrix} \beta \\ u \end{bmatrix} + \epsilon\\
+        &= \begin{bmatrix} \Gamma_X | \Gamma_Z \end{bmatrix} M \begin{bmatrix} \beta \\ u \end{bmatrix} + \epsilon\\
+        &= \begin{bmatrix} \Gamma_X | \Gamma_Z | P \end{bmatrix} \begin{bmatrix} \beta^* \\ v \end{bmatrix} + \epsilon
       \end{split}

### 2015-06-08 ms.tex
minor correction from Hodges

@@
         &= \begin{bmatrix} \Gamma_X | \Gamma_Z \end{bmatrix} M \begin{bmatrix} \beta \\ u \end{bmatrix} + \epsilon\\
-        &= \begin{bmatrix} \Gamma_X | \Gamma_Z | P \end{bmatrix} \begin{bmatrix} \beta^* \\ v \end{bmatrix} + \epsilon
+        &= \begin{bmatrix} \Gamma_X | \Gamma_Z P \end{bmatrix} \begin{bmatrix} \beta^* \\ v \end{bmatrix} + \epsilon
       \end{split}

### 2015-09-21 ms.tex
repositioned Algorithm 1
various
minor

@@
 \renewcommand{\bibname}{References}
+\linespread{1.6}
 
@@
 Multiple maxima in restricted likelihoods have received far less attention and any statement about their incidence would be speculation.
-To our knowledge, two naturally-occurring cases have have been reported, in \cite{welham_thompson:2009} and \cite{reiss_etal:2014};  \cite{henn&hodges:2014} report an artificial case and give a recipe for manufacturing examples.
+To our knowledge, two naturally-occurring cases have been reported, in \cite{welham_thompson:2009} and \cite{reiss_etal:2014};  \cite{henn&hodges:2014} report an artificial case and give a recipe for manufacturing examples.
 
@@
   \frac{\partial\logRLss}{\partial\sigssq}
-  = -\frac{1}{2} \sum_j
+  = -\frac{1}{2} \sum_{j=1}^{s_z+2}
         \left[ \frac{a_jc_j}{\abss} - \frac{a_j d_j}{(\abss)^2} \right]
-        = -\frac{1}{2} \sum_j
+        = -\frac{1}{2} \sum_{j=1}^{s_z+2}
             \frac{a_j \left( a_jc_j\sigssq + b_jc_j\sigesq  - d_j\right)}{(\abss)^2}\\
@@
   \frac{\partial\logRLss}{\partial\sigesq} =
-    -\frac{1}{2} \sum_j \left[ \frac{b_jc_j}{\abss} - \frac{b_j d_j}{(\abss)^2} \right]
-    = - \frac{1}{2} \sum_j \frac{b_j \left( a_jc_j\sigssq + b_jc_j\sigesq  - d_j\right)}{(\abss)^2}.
+    -\frac{1}{2} \sum_{j=1}^{s_z+2} \left[ \frac{b_jc_j}{\abss} - \frac{b_j d_j}{(\abss)^2} \right]
+    = - \frac{1}{2} \sum_{j=1}^{s_z+2}
+         \frac{b_j \left( a_jc_j\sigssq + b_jc_j\sigesq  - d_j\right)}{(\abss)^2}.
 \end{equation}
@@
 I.e., \textbf{D1} is satisfied.\\[5pt]
-%\noindent\textbf{Proof}
-%Let $p$ be an arbitrary point inside $B_1$, as illustrated in Figure~(\ref{fig:boundingbox2}), and define $L \equiv \logRL(p)$; we use $L$ as a lower bound on $\logRL(\sshat)$.
-The proof considers several regions outside $B$ in turn, as exemplified by the points $q_1$, $q_2$, $q_3$, and $q_4$ in Figure~\ref{fig:boundingbox2}.
-Let $\widetilde{q_1} = (0,\widetilde{\sigma_s^2})$ be the intercept of $B$ and the $\sigssq$ axis and let $\widetilde{q_2} = (\sigma_e^{2*},\widetilde{\sigma_s^2})$.
-Let $q_1 = (\sigesq{}_1, \sigssq{}_1)$ be a point satisfying $\sigesq{}_1 \le \sigma_e^{2*}$ and $\sigssq{}_1 \ge \widetilde{\sigma_s^2}$ --- above $B$ and to the left of the $s_z+1$ line.
-The partial derivatives (\ref{eq:drlds}), between $q_1$ and $\widetilde{q_1}$ are negative for $j \in \{1, \dots, s_z, s_z+2\}$ but zero for $j=s_z+1$.
-The partial derivatives (\ref{eq:drlde}), betweem $q_1$ and $\widetilde{q_2}$ are negative for $j \in \{1, \dots, s_z, s_z+1\}$ but nonnegative for $j=s_z+2$.
-Let $\logRL_j(\cdot)$ denote the $j$'th term of \eqref{eq:sum} evaluated at $(\cdot)$.
-%\begin{equation}
-%	\logRL(\sshat) - \logRL(q_1) \ge L - \sum_{j \ne s_z+1} \logRL_j(\widetilde{q_1}) - \logRL_{s_z+1}(\widetilde{q_2}),
-%\end{equation}
-%which is larger than $M$ if
-%\begin{equation}
-%\label{eq:q1}
-%  \begin{split}
-%    \sum_{j \ne s_z+1} \logRL_j(\widetilde{q_1})
-%      &< L - M - \logRL_{s_z+1}(\widetilde{q_2})\\
-%      &= L - M + \frac{1}{2} \left[ c_{s_z+1}\log\sigma_e^{2*} + \frac{d_{s_z+1}}{\sigma_e^{2*}}\right].
-%  \end{split}
-%\end{equation}
-%Examination of (\ref{eq:sum}) shows that for any fixed $\sigesq$, in particular for $\sigesq=0$, and for $j \ne s_z+1$,\\ $\lim_{\sigssq \rightarrow \infty} \logRL_j(\sigesq,\sigssq) = -\infty$.
-Thus $\widetilde{\sigma_s^2}$ can be chosen large enough so that the summation on the l.h.s.~of \eqref{eq:q1} is less than the r.h.s.~of \eqref{eq:q1}, and  \textbf{D1} is satisfied.
-%
-%Let $q_2 = (\sigesq{}_2, \sigssq{}_2)$ satisfy $\sigesq{}_2 \ge \sigma_e^{2*}$ and $\sigssq{}_2 \ge \widetilde{\sigma_s^2}$ --- above $B$ and to the right of the $s_z+1$ line.
-(Points $q_2$ and $q_3$ in Figure~(\ref{fig:boundingbox2}) both satisfy these criteria.)  The partial derivatives (\ref{eq:drlds}, \ref{eq:drlde}) evaluated at $q_2$ are nonpositive for $j \in \{1, \dots, s_z+2\}$; therefore, $\logRL(\sshat) - \logRL(q_2) \ge L - \logRL(\widetilde{q_2})$.
-The latter expression is greater than $M$ iff $\logRL(\widetilde{q_2}) < L-M$.
-But examination of (\ref{eq:sum}) shows that for any fixed $\sigesq$, $\lim_{\sigssq \rightarrow \infty} \logRLss = -\infty$ and therefore $\widetilde{\sigma_s^2}$ can be chosen large enough so that \textbf{D1} is satisfied.
-%
-%Let $\widetilde{q_3} = (\widetilde{\sigma_e^2}, \sigma_s^{2*})$ and let $q_3 = (\sigesq{}_3, \sigssq{}_3)$ satisfy $\sigesq{}_3 \ge \widetilde{\sigma_e^2}$ and $\sigssq{}_3 \ge \sigma_s^{2*}$.
-The partial derivatives (\ref{eq:drlds}, \ref{eq:drlde}) evaluated at $q_3$ are nonpositive for $j \in \{1, \dots, s_{z+2}\}$; therefore, $\logRL(\sshat) - \logRL(q_3) \ge L - \logRL(\widetilde{q_3})$.
-The latter expression is greater than $M$ iff $\logRL(\widetilde{q_3}) < L-M$.
-But examination of (\ref{eq:sum}) shows that for any fixed $\sigssq$, $\lim_{\sigesq \rightarrow \infty} \logRLss = -\infty$ and therefore $\widetilde{\sigma_e^2}$ can be chosen large enough so that \textbf{D1} is satisfied.
-%
-%Let $\widetilde{q_4} = (\widetilde{\sigma_e^2},0)$ and let $q_4 = (\sigesq{}_4, \sigssq{}_4)$ satisfy $\sigssq{}_4 \le \sigma_s^{2*}$ and $\sigesq{}_4 \ge \widetilde{q_4}$.
-The partial derivatives \eqref{eq:drlds} evaluated at $q_4$ are nonpositive for $j \in \{1, \dots, s_z+1\}$  but nonnegative for $j=s_z+2$.
-The partial derivatives \eqref{eq:drlde} evaluated at $q_4$ are nonpositive for $j \in \{1, \dots, s_z, s_z+2\}$  but nonnegative for $j=s_z+1$.
-Therefore,
-%\begin{equation*}
-%	\logRL(\sshat) - \logRL(q_4) \ge L - \sum_{j \ne s_z+2} \logRL_j(\widetilde{q_4}) - \logRL_{s_z+2}(\widetilde{q_3})
-%\end{equation*}
-%which is larger than $M$ if
-%\begin{equation}
-%\label{eq:q4}
-%  \begin{split}
-%    \sum_{j \ne s_z+2} \logRL_j(\widetilde{q_4})
-%      &< L - M - \logRL_{s_z+2}(\widetilde{q_3})\\
-%      &= L - M + \frac{1}{2} \left[ c_{s_z+2}\log\sigma_s^{2*} + \frac{d_{s_z+2}}{\sigma_s^{2*}}\right].
-%  \end{split}
-%\end{equation}
-%Examination of (\ref{eq:sum}) shows that for any fixed $\sigssq$, in particular for $\sigssq=0$, and for $j \ne s_z+2$,\\ $\lim_{\sigesq \rightarrow \infty} \logRL_j(\sigesq,\sigssq) = -\infty$.
-Thus $\widetilde{\sigma_e^2}$ can be chosen large enough so that the summation on the l.h.s.~of \eqref{eq:q4} is less than the r.h.s.~of \eqref{eq:q4}, and  \textbf{D1} is satisfied.
-%
 \noindent\textbf{Proof}
@@
   \begin{equation}
-	\logRL(\sshat) - \logRL(q) \ge L - \sum_{j \ne s_z+1} \logRL_j(\widetilde{q_1}) - \logRL_{s_z+1}(\widetilde{q_2}).
+	\logRL(\sshat) - \logRL(q) \ge L -
+	  \sum_{\substack{j=1\\ j \ne s_z+1}}^{s_z+2} \logRL_j(\widetilde{q_1}) - \logRL_{s_z+1}(\widetilde{q_2}).
   \end{equation}
@@
     \begin{split}
-      \sum_{j \ne s_z+1} \logRL_j(\widetilde{q_1})
+      \sum_{\substack{j=1\\ j \ne s_z+1}}^{s_z+2} \logRL_j(\widetilde{q_1})
         &< L - M - \logRL_{s_z+1}(\widetilde{q_2})\\
@@
   \begin{equation*}
-	\logRL(\sshat) - \logRL(q) \ge L - \sum_{j \ne s_z+2} \logRL_j(\widetilde{q_4}) - \logRL_{s_z+2}(\widetilde{q_3})
+	\logRL(\sshat) - \logRL(q) \ge L - \sum_{j=1}^{s_z+1} \logRL_j(\widetilde{q_4}) - \logRL_{s_z+2}(\widetilde{q_3})
   \end{equation*}
@@
     \begin{split}
-      \sum_{j \ne s_z+2} \logRL_j(\widetilde{q_4})
+      \sum_{j=1}^{s_z+1} \logRL_j(\widetilde{q_4})
         &< L - M - \logRL_{s_z+2}(\widetilde{q_3})\\
@@
 \label{eq:bounds}
-	L^b \equiv \sum_j L^b_j \le \inf_{(\sigssq,\sigesq)\in b} \logRLss \le
-	\sup_{(\sigssq,\sigesq)\in b} \logRLss \le \sum_j U^b_j \equiv U^b.
+	L^b \equiv \sum_{j=1}^{s_z+2} L^b \le \inf_{(\sigssq,\sigesq)\in b} \logRLss \le
+	\sup_{(\sigssq,\sigesq)\in b} \logRLss \le \sum_{j=1}^{s_z+2} U^b_j \equiv U^b.
 \end{equation}
@@
 
-\begin{algorithm}
+Algorithm~\ref{alg:findf} sketches the basic procedure; details are in an appendix.
+Before giving examples, we mention some considerations in setting the tuning constants $M$ and $\epsilon$, which have to do with the interpretation of likelihood ratios.
+We imagine a reference experiment in which a coin is chosen, either fair or two-headed, and tossed.
+If the coin is tossed twice and lands Heads both times, it produces a likelihood ratio of 4 in favor of the two-headed coin.
+That's only weak evidence, so we don't see the need to resolve likelihood functions much beyond a factor of four, or loglikelihood functions much beyond a factor of $\log(4) \approx 1.4$.
+In practice, we use $\epsilon = 1$ as a convenient default.
+If the coin were tossed 10 times and yielded 10 Heads, the likelihood ratio would be $2^{10} = 1024$.
+That's strong evidence, so we don't see the need to resolve $\logRL$ where it is less than .001 times its maximum.
+$\log(1000) \approx 7$, so we use $M = 7$ as a convenient default.
+
+\begin{algorithm}[H]
   \caption{Learn areas in $B$ where log $f$ is high to within $\epsilon$
@@
 \end{algorithm}
-
-Algorithm~\ref{alg:findf} sketches the basic procedure; details are in an appendix.
-Before giving examples, we mention some considerations in setting the tuning constants $M$ and $\epsilon$, which have to do with the interpretation of likelihood ratios.
-We imagine a reference experiment in which a coin is chosen, either fair or two-headed, and tossed.
-If the coin is tossed twice and lands Heads both times, it produces a likelihood ratio of 4 in favor of the two-headed coin.
-That's only weak evidence, so we don't see the need to resolve likelihood functions much beyond a factor of four, or loglikelihood functions much beyond a factor of $\log(4) \approx 1.4$.
-In practice, we use $\epsilon = 1$ as a convenient default.
-If the coin were tossed 10 times and yielded 10 Heads, the likelihood ratio would be $2^{10} = 1024$.
-That's strong evidence, so we don't see the need to resolve $\logRL$ where it is less than .001 times its maximum.
-$\log(1000) \approx 7$, so we use $M = 7$ as a convenient default.
 
@@
 After 25 iterations and about 35 minutes of computing time, all boxes became inactive.  Output is in Table~\ref{table:gmst_output}; boxes are displayed in Figure~\ref{fig:gmst1}.
-\begin{table}[H]
+\begin{table}%[H]
 \centering

### 2015-09-21 ms.tex
splitbox

@@
 Let $p$ be an arbitrary point inside $B_1$, as illustrated in Figure~(\ref{fig:boundingbox2}), and define $L \equiv \logRL(p)$; we use $L$ as a lower bound on $\logRL(\sshat)$.
-Let $\widetilde{q_1} = (0,\widetilde{\sigma_s^2})$, the intercept of $B$ and the $\sigssq$ axis; let $\widetilde{q_2} = (\sigma_e^{2*},\widetilde{\sigma_s^2})$; let $\widetilde{q_3} = (\widetilde{\sigma_e^2}, \sigma_s^{2*})$; let $\widetilde{q_4} = (\widetilde{\sigma_e^2},0)$, the intercept of $B$ and the $\sigssq$ axis; and let $\logRL_j$ denote the $j$'th term of \eqref{eq:sum}.
+Let $\widetilde{q_1} = (0,\widetilde{\sigma_s^2})$, the intercept of $B$ and the $\sigssq$ axis; let $\widetilde{q_2} = (\sigma_e^{2*},\widetilde{\sigma_s^2})$; let $\widetilde{q_3} = (\widetilde{\sigma_e^2}, \sigma_s^{2*})$; let $\widetilde{q_4} = (\widetilde{\sigma_e^2},0)$, the intercept of $B$ and the $\sigesq$ axis; and let $\logRL_j$ denote the $j$'th term of \eqref{eq:sum}.
 The proof considers in turn four regions $Q_1$, $Q_2$, $Q_3$, and $Q_4$ whose union is $B^c$.
@@
 \end{enumerate}
+
+\paragraph{\textcompute{findf}: The main function}
 \begin{verbatim}
@@
 }
-
+\end{verbatim}
+
+\paragraph{\textcompute{killfunc}: determine whether to make a box inactive}
+\begin{verbatim}
 # conditions under which a box becomes inactive
@@
 
+\paragraph{\textcompute{splitbox}: split a box into four children}
+\begin{verbatim}
+splitbox <- function(box, lines) {
+    # split a box into four children
+    NW <- with(box, makebox(lims.sigsqe = c(lims.sigsqe[1], mean(lims.sigsqe)),
+        lims.sigsqs = c(mean(lims.sigsqs), lims.sigsqs[2]), status = status,
+        lines = lines))
+
+    NE <- with(box, makebox(lims.sigsqe = c(mean(lims.sigsqe), lims.sigsqe[2]),
+        lims.sigsqs = c(mean(lims.sigsqs), lims.sigsqs[2]), status = status,
+        lines = lines))
+
+    SW <- with(box, makebox(lims.sigsqe = c(lims.sigsqe[1], mean(lims.sigsqe)),
+        lims.sigsqs = c(lims.sigsqs[1], mean(lims.sigsqs)), status = status,
+        lines = lines))
+
+    SE <- with(box, makebox(lims.sigsqe = c(mean(lims.sigsqe), lims.sigsqe[2]),
+        lims.sigsqs = c(lims.sigsqs[1], mean(lims.sigsqs)), status = status,
+        lines = lines))
+
+    return(list(NW, NE, SW, SE))
+}
+
+\end{verbatim}
+
 \bibliographystyle{jasa}

### 2015-09-24 ejs.tex
ejs version

--- 2015-09-21 ms.tex
@@
-\documentclass{report}
+%\documentclass{report}
+\documentclass[ejs]{imsart}
 
 \usepackage{amsmath,amsfonts}
-\usepackage{authblk}
 \usepackage{graphicx}
 \usepackage{caption,subcaption}
-%\usepackage{color}
-%\usepackage{etoolbox}
 \usepackage{float}
 \usepackage{fullpage}
-%\usepackage{mathtools}
-\usepackage{natbib}
+\usepackage[numbers]{natbib}
 \usepackage[inline]{enumitem}
 \usepackage{subdepth}
-%\usepackage{appendix}
-%\usepackage{setspace}
 \usepackage{algorithm}
@@
 \usepackage{hyperref}
-
+\usepackage{xspace}
+
+\startlocaldefs
 \renewcommand\thesection{\arabic{section}}
-\usepackage{xspace}
 \newcommand{\textcompute}{\textsf}
@@
 \newcommand{\fss}{f(\sigesq,\sigssq)\,}
-%\newcommand{\mrle}{MRLE}
 \newcommand{\mrle}{$\argsup\log f$}
-%\newcommand{\lone}{\textbf{l}_1}
-%\newcommand{\pone}{\textbf{p}_1}
-%\newcommand{\ltwo}{\textbf{l}_2}
-%\newcommand{\ptwo}{\textbf{p}_2}
-%\newcommand{\pthree}{\textbf{p}_3}
-%\newcommand{\given}{\,|\,}
 \newcommand{\g}{\,|\,}
@@
 \renewcommand{\bibname}{References}
+\endlocaldefs
+
 \linespread{1.6}
@@
 
+\begin{frontmatter}
+
 \title{Approximately Exact Calculations for Linear Mixed Models}
-\author[1]{Michael Lavine}
-\author[2]{Andrew Bray}
-\author[3]{James Hodges}
-\affil[1]{Department of Mathematics and Statistics, University of Massachusetts, Amherst, MA 01003 USA}
-\affil[2]{Reed College, Portland, OR, 97202 USA}
-\affil[3]{Division of Biostatistics, University of Minnesota, Minneapolis, MN, 55455 USA}
-
-\maketitle
+\runtitle{Approximately Exact Calculations}
+
+\begin{aug}
+
+\author{ \fnms{Michael} \snm{Lavine}  \ead[label=e1]{lavine@math.umass.edu} }
+\address{Department of Mathematics and Statistics\\ UMass Amherst\\ Amherst, MA 01003\\ USA}
+\printead{e1}
+
+\author{ \fnms{Andrew} \snm{Bray}  \ead[label=e2]{abray@reed.edu} }
+\address{Reed College\\ 3203 Southeast Woodstock Boulevard\\ Portland, Oregon 97202-8199\\ USA}
+\printead{e2}
+
+\author{ \fnms{Jim} \snm{Hodges}  \ead[label=e3]{hodge003@umn.edu} }
+\address{Division of Biostatistics\\
+               School of Public Health\\
+               University of Minnesota\\
+               2221 University Ave SE, Suite 200\\
+               Minneapolis, MN 55414\\
+               USA
+}
+\printead{e3}
+
+\runauthor{M. Lavine et al.}
+\affiliation{Some University and Another University}
+
+\end{aug}
 
@@
 \end{abstract}
+
+\begin{keyword}[class=MSC]
+\kwd[Primary ]{62J05}
+\kwd[; secondary ]{62F99}
+\end{keyword}
+
+\tableofcontents
+
+\end{frontmatter}
 
@@
 \subsubsection{Introduction to the Data}
-Our first example is a traditional linear mixed model previously analyzed in \citet{hodges:98},  \cite{wakefield:1998}, \cite{hodges:2013}, and \cite{henn&hodges:2014}.
+Our first example is a traditional linear mixed model previously analyzed in \cite{hodges:98},  \cite{wakefield:1998}, \cite{hodges:2013}, and \cite{henn&hodges:2014}.
 \cite{wakefield:1998} reported a bimodal log posterior density for $(\sigesq,\sigssq)$.
@@
 In fact, $\logRLorig(500,0) \approx -1241.5$, only about 6 log units below $\logRLorig(\sshat) \approx -1235.7$.
-Further, about their Figure~2a, Henn and Hodges say, ``No change in contour shape indicative of a local maximum could be found in the \dots region of $(500, 600) \times (10^{-3}, 1)$, regardless of contour resolution."  I.e., they cannot be sure there are no undiscovered points with large $\logRLorig$.
+Further, about their Figure~2a, \cite{henn&hodges:2014} say, ``No change in contour shape indicative of a local maximum could be found in the \dots region of $(500, 600) \times (10^{-3}, 1)$, regardless of contour resolution."  I.e., they cannot be sure there are no undiscovered points with large $\logRLorig$.
 In contrast, our algorithm guarantees there are no undiscovered points where $\logRLorig$ is more than $\epsilon$ above $L$.
@@
 Here we reproduce the analysis from \cite{hodges:98} which used inverse Gamma priors for $(\sigesq,\sigssq)$ with $\alpha_e = 1$; $\beta_e = 0$; $\alpha_s = 1.1$; and $\beta_s =0.1$.
-(We don't defend the prior; we use it so we can compare to Hodges.)
+(We don't defend the prior; we use it so we can compare to \cite{hodges:98}.)
 
@@
 Both splines can be formulated as linear mixed models.
-We follow his lead in fitting a quadratic spline with knots at 1880, 1884, 1888, \dots, 2004.
+We follow \cite{hodges:2013}'s lead in fitting a quadratic spline with knots at 1880, 1884, 1888, \dots, 2004.
 $X$ has three columns: $\textcompute{1}, \textcompute{year}, \textcompute{year}^2$.
 $Z$ is $125 \times 30$: one row for each year; one column for each knot.
-%\begin{equation*}
-%Z =	\begin{bmatrix}
-%		0 & 0 & \dots & 0\\
-%		0 & 0 & \dots & 0\\
-%		0 & 0 & \dots & 0\\
-%		0 & 0 & \dots & 0\\
-%		1 & 0 & \dots & 0\\
-%		4 & 0 & \dots & 0\\
-%		9 & 0 & \dots & 0\\
-%		16 & 0 & \dots & 0\\
-%		25 & 1 & \dots & 0\\
-%		\vdots & \vdots & \vdots & \vdots\\
-%		13924 & 12996 & \dots & 4\\
-%		14161 & 13225 & \dots & 9\\
-%		14400 & 13456 & \dots & 16\\
-%		14641 & 13689 & \dots & 25\\
-%	\end{bmatrix}; \hspace{.5cm} \Sigma_e = \mathbf{1}_{125}; \hspace{.5cm}
-%		\Sigma_s = \mathbf{1}_{30}
-%\end{equation*}
 Because we fit a quadratic spline, the entries in $Z$ are squares.
 $\Sigma_e$ and $\Sigma_s$ are identity matrices of the appropriate dimension; see \cite{hodges:2013} for details.
-Following Hodges, we center and scale the \textcompute{year} column of X, then compute the $\textcompute{year}^2$ column of $X$ and all the columns of Z from the transformed \textcompute{year}, so Z becomes
+Following \cite{hodges:2013}, we center and scale the \textcompute{year} column of X, then compute the $\textcompute{year}^2$ column of $X$ and all the columns of Z from the transformed \textcompute{year}, so Z becomes
 \begin{equation*}
@@
 \end{equation*}
-Centering and scaling changes only the scale on which $\sigssq$ is measured; we do it to more easily compare our result to Hodges'.
+Centering and scaling changes only the scale on which $\sigssq$ is measured; we do it to more easily compare our result to \cite{hodges:2013}'s.
 
@@
 	 \end{itemize}
-\item \textcompute{killfunc} is a function, shown below \textcompute{findf}, that determines whether
+\item \textcompute{killfunc} is a function, defined within \textcompute{findf}, that determines whether
 	a box $b$ should be divided more finely.
+\item \textcompute{makebox} is a function that takes limits on $\sigesq$, $\sigssq$ and returns a
+	\textcompute{box} structure. 
 \item \textcompute{splitbox} is a function that takes a box $b =
@@
 	and returns the four boxes created by dividing each side of $b$ at its midpoint.
+\item \textcompute{getstatus} is a function that calculates whether a box is above, below, or straddles
+	each line.
+\item \textcompute{getbounds} is a function that computes $\{L^b_j, U^b_j\}$.
+\item \textcompute{findlines} is a function that finds the lines for $j=1,\dots,s_z+1$ corresponding to $\logRLorig$.
+\item \textcompute{addprior} is a function that adjusts the $s_z+1$ line and adds the $s_z+2$ line
+	to account for a conjugate prior.
 \end{enumerate}
+
+%findf <- function(lines, startbox, eps = 0, delE = 0, delS = 0, M = Inf, maxit = 10,
+%                  ratio = FALSE, lognote = "summary") {
+%
+%
+%    inactive <- list()  # a list of inactive boxes
+%    ninact <- 0  # the number of inactive boxes
+%    
+%    active <- list(startbox)  # a list of active boxes
+%    nact <- 0  # the number of active boxes
+%
+%    lowbound <- -Inf  # lowerbound on max(log(f))
+%    iter <- 0  # iteration number
+%
+%    while (nact > 0 && iter < maxit) {
+%        # Find the lower bound of each box and the maximum of the lower bounds.  For
+%        # each active box, either make it inactive or divide it.
+%        low.act <- max(vapply(X = active, FUN = function(box) {
+%            box$bounds[1]
+%        }, FUN.VALUE = 0.1))
+%        lowbound <- max(lowbound, low.act)
+%        kill <- vapply(X = active, FUN = killfunc, FUN.VALUE = TRUE, lb = lowbound,
+%            M = M, eps = eps, delE - delE, delS = delS, ratio = ratio)
+%            # which boxes become inactive?
+%        nkill <- sum(kill)
+%        if (nkill > 0) {
+%            inactive[(ninact + 1):(ninact + nkill)] <- active[kill] 
+%            # add those boxes to the inactive list
+%        }
+%        ninact <- length(inactive)
+%        kids <- list()  # subdivisions of active boxes
+%        nkids <- 0
+%        for (i in which(!kill)) {  # split active boxes into 4 parts
+%            kids[(nkids + 1):(nkids + 4)] <- splitbox(active[[i]], lines)
+%            nkids <- nkids + 4
+%        }
+%        active <- kids
+%        nact <- length(active)
+%
+%        iter <- iter + 1
+%        write(c("iteration", iter, "nact", nact, "ninact", ninact, "lowbound",
+%            lowbound), file = "log.out", ncolumns = 8, append = TRUE)
+%    }
+%
+%
+%    tmp <- t(vapply(X = c(active, inactive), FUN = function(box) {
+%        c(box$lims.sigsqs, box$lims.sigsqe, box$bounds)
+%    }, FUN.VALUE = c(sigsqs.lo = 0.1, sigsqs.hi = 0.1, sigsqe.lo = 0.1, sigsqe.hi = 0.1,
+%        rll.lower = 0.1, rll.upper = 0.1)))
+%    end_time <- Sys.time()
+%    write(paste(lognote, "runtime: ", round(end_time - start_time, digits = 3)),
+%          file = "log.out", ncolumns = 1, append = TRUE)
+%    return(data.frame(tmp))
+%}
+
+\paragraph{\textcompute{findf}: The main function}
+
 \begin{verbatim}
+#' Learn RLL or Log-posterior for two variance MM.
+#'
+#' Learns the shape of the objective function within epsilon.
+#'
+#' This is the primary function that implements the branch-bound-kill
+#' algorithm described in "Approximately Exact Calculations for Linear
+#' Mixed Models" by Lavine & Hodges (2015). This an interative algorithm
+#' that can require substantial computation time. It is recommended that
+#' the user start with a conservative \code{maxit} and do additional
+#' computations as necessary.
+#'
+#' The arguments of this function include five settings to control when
+#' to stop branching (subdividing) a box:
+#' \enumerate{
+#'   \item when the upper and lower bounds are within \code{eps} of one another.
+#'   \item when the width of \eqn{\sigma^2_e} is less than \code{delE}.
+#'   \item when the width of \eqn{\sigma^2_s} is less than \code{delS}.
+#'   \item when the upper bound is more than \code{M} below the highest global
+#'   lower bound.
+#'   \item when the number of iterations reaches \code{maxit}.
+#'   }
+#'
+#' @param lines a dataframe that contains the constants that define the
+#'   line represented by each term in the sum. Created as output from
+#'   \code{\link{findlines}}.
+#' @param startbox a list of boxes and their bounds, possibly the output
+#'   from a previous call to \code{findf}. If left empty, will create
+#'   a new startbox from a call to \code{makebox}.
+#' @param eps a non-negative numeric indicating the tolerance within which you
+#'   would like to learn the function. Default is 0.
+#' @param delE a non-negative numeric indicating the width of \eqn{\sigma^2_e}
+#'   beyond which the algorithm will stop branching boxes. Default is 0.
+#' @param delS a non-negative numeric indicating the width of \eqn{\sigma^2_s}
+#'   beyond which the algorithm will stop branching boxes. Default is 0.
+#' @param M a non-negative numeric indicating the size of the buffer between
+#'   the highest global lower bound and the upper bound of a given box past
+#'   that box will now be branched further.  Default is \code{Inf}.
+#' @param maxit a positive integer indicating the maximum number of iterations
+#'   of the algorithm.  Default is 10.
+#' @param ratio a logical indicating if \code{delE} and \code{delS} are specified as
+#'   ZZQ.  Default is \code{FALSE}.
+#' @param lognote a string to append to the "log.out" log file to annotate
+#'   the purpose of each run of the algorithm. For use in benchmarking computation
+#'   time. Default is \code{"summary"}.
+#'
+#' @return A list of boxes including their limits in \eqn{\sigma^2_e} and
+#'   \eqn{\sigma^2_e} as well as their bounds. Running this function also results
+#'   in the creation of a log file called "log.out" containing box counts at every
+#'   iteration.
+
 findf <- function(lines, startbox, eps = 0, delE = 0, delS = 0, M = Inf, maxit = 10,
@@
 
-    inactive <- list()  # a list of inactive boxes
-    ninact <- 0  # the number of inactive boxes
-
-    active <- list(startbox)  # a list of active boxes
-    nact <- 0  # the number of active boxes
-
-    lowbound <- -Inf  # lowerbound on max(log(f))
-    iter <- 0  # iteration number
+    start_time <- Sys.time()
+
+    if (missing(lines)) {
+        print("please supply lines")
+        return
+    }
+    if (missing(startbox)) {
+        startbox <- makebox(lines = lines, lims.sigsqs = c(0, max(lines$int.sigsqs[is.finite(lines$int.sigsqs)])),
+            lims.sigsqe = c(0, max(lines$int.sigsqe[is.finite(lines$int.sigsqe)])),
+            status = rep("straddle", nrow(lines)))
+    }
+    inactive <- list()
+    ninact <- 0
+    # check whether startbox is a box or a list of boxes
+    active <- ifelse(length(startbox) == 4 && identical(names(startbox), c("lims.sigsqe",
+        "lims.sigsqs", "status", "bounds")), list(startbox), startbox)
+    nact <- length(active)
+
+    lowbound <- -Inf
+    iter <- 0
+
+    # conditions under which a box becomes inactive
+    killfunc <- function(box, lb, M, eps, delE, delS, ratio) {
+        # lb is a lower bound on max f; it changes at each iteration. M, eps,
+        # delE, delS stay constant throughout the iterations.
+        cond.low <- box$bounds[2] < lb - M
+        cond.eps <- diff(box$bounds) < eps
+        cond.E <- ifelse(ratio, diff(log(box$lims.sigsqe)) < delE, diff(box$lims.sigsqe) <
+            delE)
+        cond.S <- ifelse(ratio, diff(log(box$lims.sigsqs)) < delS, diff(box$lims.sigsqs) <
+            delS)
+        return(cond.low || cond.eps || cond.E || cond.S)
+    }
 
@@
             M = M, eps = eps, delE - delE, delS = delS, ratio = ratio)
-            # which boxes become inactive?
         nkill <- sum(kill)
@@
             inactive[(ninact + 1):(ninact + nkill)] <- active[kill]
-            # add those boxes to the inactive list
         }
         ninact <- length(inactive)
-        kids <- list()  # subdivisions of active boxes
+        kids <- list()
         nkids <- 0
-        for (i in which(!kill)) {  # split active boxes into 4 parts
-            kids[(nkids + 1):(nkids + 4)] <- splitbox(active[[i]], lines)
+        for (i in which(!kill)) {
+            kids[(nkids + 1):(nkids + 4)] <- splitbox(active[[i]], lines)  # boxes are split into 4 parts
             nkids <- nkids + 4
@@
 }
-
-# conditions under which a box becomes inactive
-killfunc <- function(box, lb, M, eps, delE, delS, ratio) {
-    # lb is a global (within startbox) lower bound on max(log(f));
-    #      it changes at each iteration.
-    #  M, eps, delE, delS stay constant throughout the iterations.
-    cond.low <- box$bounds[2] < lb - M
-    cond.eps <- diff(box$bounds) < eps
-    cond.E <- ifelse(ratio, diff(log(box$lims.sigsqe)) < delE,
-                            diff(box$lims.sigsqe) < delE)
-    cond.S <- ifelse(ratio, diff(log(box$lims.sigsqs)) < delS,
-                            diff(box$lims.sigsqs) < delS)
-    return(cond.low || cond.eps || cond.E || cond.S)  # stop dividing this box?
+\end{verbatim}
+
+%\paragraph{\textcompute{killfunc}: determine whether to make a box inactive}
+%\begin{verbatim}
+%# conditions under which a box becomes inactive
+%killfunc <- function(box, lb, M, eps, delE, delS, ratio) {
+%    # lb is a global (within startbox) lower bound on max(log(f));
+%    #      it changes at each iteration.
+%    #  M, eps, delE, delS stay constant throughout the iterations.
+%    cond.low <- box$bounds[2] < lb - M
+%    cond.eps <- diff(box$bounds) < eps
+%    cond.E <- ifelse(ratio, diff(log(box$lims.sigsqe)) < delE,
+%                            diff(box$lims.sigsqe) < delE)
+%    cond.S <- ifelse(ratio, diff(log(box$lims.sigsqs)) < delS,
+%                            diff(box$lims.sigsqs) < delS)
+%    return(cond.low || cond.eps || cond.E || cond.S)  # stop dividing this box?
+%}
+%\end{verbatim}
+
+\paragraph{\textcompute{makebox}: make a box}
+
+\begin{verbatim}
+#' Make a box with bounds
+#'
+#' Given the limits of the box, calculate its status relative to each line and
+#' calculate the bounds on the objective function within that box.
+#'
+#' @param lims.sigsqe numeric vector of length 2 of the form \code{c(lower, upper)}
+#'   containing the x-limits of the box.
+#' @param lims.sigsqs numeric vector of length 2 of the form \code{c(lower, upper)}.
+#'   containing the y-limits of the box.
+#' @param status factor of length of \code{lines} denoting the location of the box
+#'   relative to each line. If the box has a parent, these correspond to the status
+#'   of that parent. The levels are \{above, below, straddle\}.
+#' @param lines dataframe containing the constants that define the shape of each
+#'   term in the sum. Output from \code{\link{findlines}}.
+#'
+#' @return A list containing:
+#' \itemize{
+#'   \item limits of the box in \eqn{\sigma^2_e} (numeric vector of length 2)
+#'   \item limits of the box in \eqn{\sigma^2_s} (numeric vector of length 2)
+#'   \item the status of the box relative to each line (factor of length of the
+#'   number of terms in the sum)
+#'   \item the bounds on the objective function (numeric vector of length 2 of
+#'   the form \code{c(lower, upper)})
+#'   }
+
+makebox <- function(lims.sigsqe = NA, lims.sigsqs = NA, status = NA, lines) {
+
+    # sanity checks
+    if (missing(lims.sigsqs) || missing(lims.sigsqe))
+        print("please supply lims.sigsqs and lims.sigsqe")
+    if (missing(status))
+        print("please supply status")
+    if (missing(lines))
+        print("please supply lines")
+
+    # If the box has a parent then it inherits the parent's status.  But if the
+    # parent straddles a line, we must check whether the child also straddles the
+    # line.
+    strad <- which(status == "straddle")
+    status[strad] <- getstatus(lims.sigsqe = lims.sigsqe, lims.sigsqs = lims.sigsqs,
+        lines = lines[strad, ])
+
+    # we could get some of the bounds from the parent, but it's just as easy to
+    # recalculate them
+    bounds <- getbounds(lims.sigsqe = lims.sigsqe, lims.sigsqs = lims.sigsqs,
+        status = status, lines = lines)
+
+    return(list(lims.sigsqe = lims.sigsqe, lims.sigsqs = lims.sigsqs, status = status,
+        bounds = colSums(bounds)))
+}
+\end{verbatim}
+
+\paragraph{\textcompute{splitbox}: split a box into 4 children}
+
+\begin{verbatim}
+#' Split a parent box into 4 children.
+#'
+#' Split a parent box into four equally sized boxes as part of the branching part
+#' of the algorithm.
+#'
+#' @param box list containing the properties of a parent box. Created as output
+#'   from \code{\link{makebox}}.
+#' @param lines dataframe containing the constants that define the shape of each
+#'   term in the sum. Output from \code{\link{findlines}}.
+#'
+#' @return a list of the four child boxes.
+
+splitbox <- function(box, lines) {
+    # split a box into four children
+    NW <- with(box, makebox(lims.sigsqe = c(lims.sigsqe[1], mean(lims.sigsqe)),
+        lims.sigsqs = c(mean(lims.sigsqs), lims.sigsqs[2]), status = status,
+        lines = lines))
+    NE <- with(box, makebox(lims.sigsqe = c(mean(lims.sigsqe), lims.sigsqe[2]),
+        lims.sigsqs = c(mean(lims.sigsqs), lims.sigsqs[2]), status = status,
+        lines = lines))
+    SW <- with(box, makebox(lims.sigsqe = c(lims.sigsqe[1], mean(lims.sigsqe)),
+        lims.sigsqs = c(lims.sigsqs[1], mean(lims.sigsqs)), status = status,
+        lines = lines))
+    SE <- with(box, makebox(lims.sigsqe = c(mean(lims.sigsqe), lims.sigsqe[2]),
+        lims.sigsqs = c(lims.sigsqs[1], mean(lims.sigsqs)), status = status,
+        lines = lines))
+    return(list(NW, NE, SW, SE))
+}
+\end{verbatim}
+
+\paragraph{\textcompute{getstatus}: determine where a box is, relative to the lines}
+
+\begin{verbatim}
+#' Get the location of a box relative to lines.
+#'
+#' Computes the location of a box with given limits relative to each of
+#' the lines.
+#'
+#' @param lims.sigsqe numeric vector of length 2 of the form \code{c(lower, upper)}
+#'   containing the x-limits of the box.
+#' @param lims.sigsqs numeric vector of length 2 of the form \code{c(lower, upper)}
+#'   containing the y-limits of the box.
+#' @param lines dataframe containing the constants that define the shape of each
+#'   term in the sum. Output from \code{\link{findlines}}.
+#'
+#' @return A factor of length of \code{lines} denoting the location of the box
+#'   relative to each line.
+
+getstatus <- function(lims.sigsqe, lims.sigsqs, lines) {
+    # Is a box above, below, or straddling the lines with these slopes and
+    # intercepts?
+
+    # value of the lines at the left side of the box
+    tmp1 <- lines$int.sigsqs + lines$slope * lims.sigsqe[1]
+    tmp1[is.infinite(tmp1)] <- NA
+
+    # value of the lines at the right side of the box
+    tmp2 <- lines$int.sigsqs + lines$slope * lims.sigsqe[2]
+    tmp2[is.infinite(tmp2)] <- NA
+
+    # where is the box relative to the lines?
+    above <- with(lines, ifelse(slope > -Inf, lims.sigsqs[1] > tmp1, lims.sigsqe[1] > int.sigsqe))
+    below <- with(lines, ifelse(slope > -Inf, lims.sigsqs[2] < tmp2, lims.sigsqe[2] < int.sigsqe))
+    status <- rep("straddle", nrow(lines))
+    status[above] <- "above"
+    status[below] <- "below"
+
+    return(status)
+}
+\end{verbatim}
+
+\paragraph{\textcompute{getbounds}: compute $\{L^b_j, U^b_j\}$}
+
+\begin{verbatim}
+#' Get bounds for a given box
+#'
+#' Computes the lower and upper bounds on every term in the sum of
+#' the objective function within a box.
+#'
+#' @param lims.sigsqe numeric vector of length 2 of the form \code{c(lower, upper)}
+#'   containing the x-limits of the box.
+#' @param lims.sigsqs numeric vector of length 2 of the form \code{c(lower, upper)}
+#'   containing the y-limits of the box.
+#' @param lines dataframe containing the constants that define the shape of each
+#'   term in the sum. Output from \code{\link{findlines}}.
+#' @param status factor of length of \code{lines} denoting the location of the box
+#'   relative to each line.
+#'
+#' @return A matrix with a row corresponding each term in the sum. The first column
+#' contains the lower bound of that term within the box and the second column
+#' contains the upper bound. Note that the \code{colSums} of this matrix yields the
+#' \code{c(lower, upper)} bounds on the full objective function.
+
+getbounds <- function(lims.sigsqe, lims.sigsqs, lines, status) {
+    # small sanity check
+    if (missing(lims.sigsqe) || missing(lims.sigsqs))
+        print("Please supply lims.sigsqe and lims.sigsqs")
+    if (missing(lines))
+        print("Please supply lines")
+    if (missing(status))
+        print("Please supply status")
+    if (length(status) != nrow(lines))
+        print("length(status) != nrow(lines)")
+
+    # evaluate each line at the upper-right corner of the box
+    ur <- with(lines, a * lims.sigsqs[2] + b * lims.sigsqe[2])
+    eval.ur <- with(lines, -0.5 * (multiplier.log * log(ur) + multiplier.inv/ur))
+    # evaluate each line at the lower-left corner of the box
+    ll <- with(lines, a * lims.sigsqs[1] + b * lims.sigsqe[1])
+    eval.ll <- ifelse(ll == 0, -Inf, with(lines, -0.5 * (multiplier.log * log(ll) +
+        multiplier.inv/ll)))
+
+    bounds <- matrix(rep(eval.ur, 2), ncol = 2)
+    # The next two lines of code are for lines that are not straddled.  'above'
+    # means the box is above the line
+    bounds[status != "above", 1] <- eval.ll[status != "above"] # lower
+    bounds[status == "above", 2] <- eval.ll[status == "above"] # upper
+    # now we'll take care of straddled lines
+    strad <- status == "straddle"
+    bounds[strad, 1] <- pmin(eval.ur[strad], eval.ll[strad]) # lower
+    # for the upper bound, we can evaluate anywhere on the line, so we might as
+    # well evaluate at (int.sigsqe,0)
+    bounds[strad, 2] <- with(lines[strad, ], ifelse(is.na(int.sigsqe), -0.5 *
+       (multiplier.log * log(int.sigsqs) + multiplier.inv/int.sigsqs), -0.5 *
+       (multiplier.log * log(int.sigsqe) + multiplier.inv/int.sigsqe)))
+
+    return(bounds)
+}
+\end{verbatim}
+
+\paragraph{\textcompute{findlines}: find the $s_z+1$ lines for $\logRLorig$}
+
+\begin{verbatim}
+#' Find the constants that define each line.
+#'
+#' Given X, Y, Z, find the constants {a_j, b_j, c_j, d_j} that define the shape
+#' of each term in the sum.
+#'
+#' @param x a matrix with n rows corresponding to the fixed effects.
+#' @param z a matrix with n rows corresponding to the random effects.
+#' @param y a numeric vector of length n.
+#' @param SigE an n x n covariance matrix for the random error.
+#' @param SigS an n x n covariance matrix for the random effects.
+#'
+#' @return A dataframe of containing the constants that define the shape of each
+#'   term in the sum.
+
+findlines <- function(x, z, y, SigE, SigS) {
+    if (!is.matrix(x))
+        print("x should be a matrix")
+    if (!is.matrix(z))
+        print("z should be a matrix")
+    if (!is.vector(y))
+        print("y should be a vector")
+
+    n <- length(y)
+    nx <- ncol(x)
+    nz <- ncol(z)
+
+    if (nrow(x) != n)
+        print("nrow(x) != length(y)")
+    if (nrow(z) != n)
+        print("nrow(z) != length(y)")
+
+    if (!is.matrix(SigE))
+        print("SigE should be a matrix")
+    if (!is.matrix(SigS))
+        print("SigS should be a matrix")
+    if (nrow(SigE) != n || ncol(SigE) != n)
+        print("SigE should be a square matrix to match length(y)")
+    if (nrow(SigS) != nz || ncol(SigS) != nz)
+        print("SigS should be a square matrix to match ncol(z)")
+
+    # Is SigE the identity?  If not, make it so.
+    if (!identical(SigE, diag(n))) {
+        tmp <- inv.sqrt(SigE)
+        y <- tmp %*% y
+        x <- tmp %*% x
+        z <- tmp %*% z
+        SigE <- diag(n)
+    }
+
+    # Is SigS the identity?  If not, make it so.
+    if (!identical(SigS, diag(nz))) {
+        tmp <- sqrt.m(SigS)
+        z <- z %*% tmp
+        SigS <- diag(nz)
+    }
+
+    # sx, sz, Gamma_x, Gamma_z, Gamma_c # Is Gamma_c really needed?
+    qrx <- qr(x, LAPACK = FALSE)  # use LINPACK to get rank(x)
+    sx <- qrx$rank
+    Gamx <- qr.Q(qrx)[, 1:sx]
+    if (sx == 1)
+        Gamx <- matrix(Gamx, ncol = 1)
+
+    tmp <- qr.resid(qrx, z)
+    qrz <- qr(tmp, LAPACK = FALSE)  # use LINPACK to get rank(x)
+    sz <- qrz$rank
+    Gamz <- qr.Q(qrz)[, 1:sz]
+    if (sz == 1)
+        Gamz <- matrix(Gamz, ncol = 1)
+
+    M <- qr.solve(cbind(Gamx, Gamz), cbind(x, z))
+    M.zz <- M[-(1:sx), -(1:ncol(x)), drop = FALSE]
+    tmp <- svd(M.zz)
+    a <- tmp$d^2  # follows Eq (15.1)
+    v <- t(tmp$u) %*% t(Gamz) %*% y  # follows Eq (15.4)
+    if (length(a) != sz)
+        print("length(a) != sz")
+    if (length(v) != sz)
+        print("length(v) != sz")
+    rss <- sum(resid(lm(y ~ cbind(Gamx, Gamz)))^2)
+
+    lines <- data.frame(a = c(a, 0), v = c(v, sqrt(rss)), int.sigsqs = c(v^2/a,
+        NA), int.sigsqe = c(v^2, rss/(n - (sx + sz))), slope = c(-1/a, -Inf),
+        multiplier.log = c(rep(1, sz), n - (sx + sz)),
+        multiplier.inv = c(v^2, rss), b = 1)
+    return(lines)
+}
+\end{verbatim}
+
+\paragraph{\textcompute{addprior}: adjust the $s_z+1$ line and add the $s_z+2$ line to account for a prior}
+
+\begin{verbatim}
+#' Add a prior to the RLL
+#'
+#' Modifies the \code{lines} dataframe that is output from the
+#' \code{\link{findlines}} function to add a term corresponding to the
+#' prior for use in a Bayesian analysis.
+#'
+#' Working with the posterior density of \eqn{\sigma^2_e, \sigma^2_e} instead
+#' of the RLL requires this incorporation of a prior distribution on those same
+#' parameters. We work here with the inverse Gamma, which is the conjugate prior,
+#' \enumerate{
+#'  \item \eqn{\sigma^2_e ~ IG(\alpha_e, \beta_e)}
+#'  \item \eqn{\sigma^2_s ~ IG(\alpha_s, \beta_s)}
+#'  }
+#' Given a dataframe output from \code{findlines},
+#'
+#' @param alpha_e a positive numeric. The shape parameter for the IG prior on
+#'   \eqn{\sigma^2_e}.
+#' @param beta_e a positive numeric. The scale parameter for the IG prior on
+#'   \eqn{\sigma^2_e}.
+#' @param alpha_s a positive numeric. The shape parameter for the IG prior on
+#'   \eqn{\sigma^2_s}.
+#' @param beta_s a positive numeric. The scale parameter for the IG prior on
+#'   \eqn{\sigma^2_s}.
+#'
+#' @return Returns a new \code{lines} dataframe with the constants associated
+#' with the prior term appended to the bottom.
+
+addprior <- function(lines, alpha_e = 0, beta_e = 0, alpha_s = 0, beta_s = 0) {
+    if ((alpha_e != 0) || (beta_e != 0)) {
+        vert <- which(lines$slope == -Inf)
+        lines$multiplier.log[vert] <- lines$multiplier.log[vert] + 2 * (alpha_e +
+            1)
+        lines$multiplier.inv[vert] <- lines$multiplier.inv[vert] + beta_e
+        lines$v[vert] <- sqrt(lines$multiplier.inv[vert])
+        lines$int.sigsqe[vert] <- lines$multiplier.inv[vert]/lines$multiplier.log[vert]
+    }
+    if ((alpha_s != 0) || (beta_s != 0)) {
+        horiz <- data.frame(a = 1, multiplier.inv = 2 * beta_s, multiplier.log = 2 *
+            (alpha_s + 1), v = sqrt(2 * beta_s), int.sigsqs = beta_s/(alpha_s + 1),
+            int.sigsqe = NA, slope = 0, b = 0)
+        lines <- rbind(lines, horiz)
+    }
+    return(lines)
 }
